### 2009-05-13 product-graphics.tex
Separate paper, notes and code

\documentclass[letterpaper,oneside]{scrartcl}
\usepackage{fullpage}
\usepackage[utf8]{inputenc}
\usepackage[pdftex]{graphicx}
\DeclareGraphicsExtensions{.png,.pdf}
\usepackage{hyperref}
\usepackage{color}
\definecolor{slateblue}{rgb}{0.07,0.07,0.488}
\hypersetup{colorlinks=true,linkcolor=slateblue,anchorcolor=slateblue,citecolor=slateblue,filecolor=slateblue,urlcolor=slateblue,bookmarksnumbered=true,pdfview=FitB}
\usepackage{url}
\usepackage[round,sectionbib]{natbib}
\bibliographystyle{abbrvnat}
\usepackage[small]{caption2}
\usepackage[small]{titlesec}
\renewcommand\familydefault{bch}

\title{Product graphics}
\author{Heike Hofmann and Hadley Wickham}
\begin{document}
\maketitle

\section{Introduction}

The Grammar of Graphics [reference] is a grammar describing the components of statistical graphics and how they can be combined together.  These components are very flexible and can be used to describe any type of plot.  However, the book (Grammar with a capital G) does not present many examples of graphics for categorical data, and in those examples the use of the grammar gives little insight into plot.  This paper describes common categorical plots, and presents a specialisation of the Grammar of Graphics specifically designed to describe these types of plots and give insight into the types of relationships that they can show.

Area plots [reference] display categorical data with a rectangle for each combination of the factors of interest, with the area of the rectangle proportional to the number of observations in that combination.  The layout of the rectangles give rise to the difference categorical plots that we are familiar with: bar charts, spine charts, mosaic plots, equal-bin-size plots, and fluctuation diagrams.  Trellis graphics are another type of display that uses categorical variables to create small multiples for different subsets of the data.  Visually, trellis graphics look like equal-bin-size plots with another plot drawn inside each bin.  Figure \ref{fig:cat-examples} illustrates some of these plots.

\begin{figure}[htbp]
  \begin{center}
  %  \includegraphics[scale=1]{file}
  \end{center}
  \caption{Area plot examples.}
  \label{fig:cat-examples}
\end{figure}

Area plots are the graphical equivalent to contingency tables.  And like contingency tables, they can be used to display raw counts, or to display proportions or percentages.  Similarly, it may be useful to augment area plots with marginal displays as we augment contingency tables with marginal sums.

Area plots different to other types of graphics, because must be able to flexible change level of aggregation.  Values are typically counts, but can also be sums of observation level weights.

These techniques can also be used with continuous data, if the data is binned to create a categorical variable.  This gives rise to the histogram and spinogram, analogues of the bar chart and spine chart respectively.  A long standing tradition is that no gaps are displayed between adjacent rectangles when used for originally continuous data.  Two examples are shown in Figure \ref{fig:cont-examples}.

\begin{figure}[htbp]
  \begin{center}
  \end{center}
  \caption{Area plots of originally continuous data}
  \label{fig:cont-examples}
\end{figure}

These techniques are also closely related to TreeMaps and dimensional stacking \citep{leblanc:1990}

What are the common components of all these plots, and how can be describe them in a succinct way that also allows us to describe what tasks each display is good for?  We propose that there is one key feature that makes these plots special.  That is there their coordinate system, which is hierarchical.

In most coordinate systems that we are used to, such as Cartesian or polar coordinates, are symmetric in the sense that you can locate a point by starting from either dimension.  For example, in Cartesian coordinates you can identify either the x- or y-coordinate first, and the other next.  This also implies a distance symmetry when projected onto 2D Euclidean geometry. When you vary one coordinate by a small amount (holding all others constant) points in the projective geometry will be close.

Neither of these properties hold for hierarchical coordinate systems.

We will discuss hierachical coordinate systems in general, and then present a specific hierachical coordinate system which gives rise to the area plots described previously.  We also present an R implementation, in the R package {\tt ggplot}.

Related work:

\citep{vliegen:2006}
\citep{ahlberg:1996}

\citep{hartigan:1984,hartigan:1981}
\citep{friendly:1999,friendly:1994}
\citep{theus:1999}
\citep{hofmann:2000,hofmann:2001,hofmann:2003}

Goals:

\begin{itemize}
  \item Put all area plots (bar charts, spine plots, mosaic plots, fluctuation diagrams) on a common footing
  \item Continue to build link between graphics and log-linear models
  \item Provide tools for ``residuals and re-iteration''
\end{itemize}

Basic idea is factorising table of probabilities:

\begin{itemize}
  \item as products of marginal and conditional distributions
  \item as areas in plots
  \item as sums of parameters in log-linear models
\end{itemize}

\section{Data}
\label{sec:data}

\subsection{Factorising probabilities}

Let $f(x, y, z)$ be the 2d pmf function generated.  Can write this pmf as product of marginals and conditionals

\begin{itemize}
  \item $f(x, y, z) = f(x, y | z) f(z)$
  \item $f(x, y, z) = f(x | y, z) f(y, z)$
  \item $f(x, y, z) = f(x | y, z) f(y | z) f(z)$
\end{itemize}

Compare to area of mosaic plot.

One other special 2d case is the equal bin size plot - which all 1d x 1d primitive plots collapse down to when counts are equal.  The important difference is the formatting - a grid rather than nested boxes.

\subsection{Weighting}
\label{sub:weighting}

We have assumed the probabilities represent counts, but WLOG we can use any set of non-negative additive weights instead.  Some of the applications of such weighted plots are described in \citet{unwin:2007}.

\subsection{Continuous data}
\label{sub:continuous_data}

Binning.

Convention when using binned continuous data is to remove the gap between regions.

\subsection{Nested data}
\label{sub:nested_data}

\section{Display}
\label{sec:display}

The three plots above are special cases of $a(i, j) = w(i, j)^x * h(i, j)^y$.  $x + y = 1$.

Perceptual constraints.  Best at comparing area when:

\begin{itemize}
  \item Simple shape (i.e. square vs polygon)
  \item Aspect ratios similar
  \item Areas are large
  \item Areas are close
  \item One border is constant.
\end{itemize}

Impossible to simultaneously satisfy all constraints.  Every product plot requires the creator to trade off between these.

Restrictions on partitioning

\begin{itemize}
  \item Area should be proportional to weight
  \item Containment
  \item Rectangular
  \item Non-overlapping
\end{itemize}

Pseudo 2d, where you take a long 1d structure and wrap it into 2d.

Pseduo 1d, where you make a new variable from multiple variables.

\subsection{1d primitives}

Space-filling vs. not space-filling.  Read length on common scale, vs. put more information in the display.

Basic principle of all area plots is area should be proportional to probability. For the plots we are interested in, the shape used to represent the count is a rectangle, so that $a(i, j) = w(i, j) * h(i, j)$. Each plot type constrains this relationship in a different way:

\begin{itemize}
  \item {\bf bar}: height is proportional to value, width equally divides space. Bars allow comparison between absolute numbers.

  \item {\bf spine}: width is proportional to value, height equally divides space. Spines not useful at top level, but allow comparison of proportions at next level (ie. conditioned on top level values)

  \item {\bf treemap}: no restrictions on height and width.

\end{itemize}

\subsection{2d primitives}

\begin{itemize}
  \item {\bf fluct}: width and height proportional to square root of value. Flucts allow comparisons of (approximately) absolute numbers in two directions.

\end{itemize}

\subsection{Plot templates}

\begin{itemize}
  \item {\bf Barchart}: (bars, v/h, width)
  \item {\bf Histogram}: (bars, v/h, 0)
  \item {\bf Stacked} barchart + (bars, h/v, zero)
  \item {\bf Side by side}: (bins, v/h, zero) + barchart  (or possibly spines with width prop to number of bars inside) - that would  give different scales, or use nesting operator with standard barchart
  \item {\bf Spineplot}: (spines, v/h, width)
  \item {\bf Stacked} spineplot: spineplot + (bars, h/v, 0)
  \item {\bf Spinogram}: (spines, v/h, 0)
  \item {\bf Fluctuation} diagram: (flucts, both, total)
  \item {\bf Stacked} fluctuation diagram: fluctuation + (spines, v/h, zero)
  \item {\bf Mosaic} plots: (spines, v/h, width) + (bar, h/v, width) + ... (alternating)
  \item {\bf Double-decker} plot: 
\end{itemize}

\subsection{Violation of constraints}

\subsubsection{Small values and zeros}
\subsubsection{Cascading}

The cascaded treemaps of \citet{lu:2008} is an idea that illustrates how the violation of containment can be productive.  In the cascaded treemap, each level is slightly offset from the one above to create a pseudo-3d perspective.

\subsubsection{Non-rectangular partitions}

Pie charts fall out naturally, as bars in polar coordinates, with angle and radius instead of height and width. Corrections to calculation of area.  .  This is related to the infoslices of \citet{andrews:1998}, which only use half of the disk.  Figure~\ref{fig:polar} shows some examples of product plots in polar coordinates.

Circular treemaps of \citet{wetzel:2008} which use circles instead of rectangles, but is not space filling because you can not tile a circular region with circles. The radial displays of \citet{stasko:2000} deliberately keep the radius constant in order to relatively highlight the lower levels.  Similarly so does the fanlens of \citet{lou:2007}

Other non-rectangular treemaps have been proposed, such as the space-filling curve approach of \citep{wattenberg:2005}, or the voronoi treemaps of \citep{wattenberg:2005}, but because of the perceptual difficultly of comparing areas of arbitrary polygons, these approaches tend to be attractive rather than useful.

\section{Connection with log-linear models}
\label{sec:models}

which imply the obvious mapping between d(i, j) and a(i, j).  For >2d,
the definitions generalise in the obvious, but the constraints become
more complicated.

$log(d(i,j)) = 1 + d_{i \cdot} + d_{j|i}$
$log(d(i,j)) = 1 + d_{\cdot j} + d_{i|j}$

\section{Labelling}
\label{sec:legends}

\begin{itemize}
  \item Additional info 
  \begin{itemise}
    \item colour (map of the market)
    \item texture (sieve plots)
    \item embedded plots (time series in lab escape)
    \item photographs
    \item text (tables)
  \end{itemise}

  \item Display of hierarchy
  \begin{itemize}
    \item Spacing / borders
    \item Shading
    \item Cascading
    \item Labelling
  \end{itemize}
\end{itemize}

% bibtool -x product-graphics -o references.bib
\bibliography{references}
\end{document}

### 2009-05-13 product-graphics.tex
Update data, add description to paper

@@
 \label{sec:data}
+
+To illustrate the techniques in this paper, we'll use a small sample from the general social survey ({\sc gss}) focussing on happiness.
+51\,020 observations from a yearly survey from 1972 to 2006.
+Happy, age, sex, marital status, terminal education level, relative financial status and health.
+Also contains a weighting variable wtssall, which is a probability weight from the survey.
+
+The survey also contains a respondent id variable.  4510 people in sample,  1271 people interviewed only once, 1107 interviewed every year.
 

### 2009-05-13 product-graphics.tex
The id variable is not the same across years :(
Add some examples of polar plots.  Tweak template names

@@
 \DeclareGraphicsExtensions{.png,.pdf}
+\graphicspath{{plots/}}
 \usepackage{hyperref}
-\usepackage{color}
-\definecolor{slateblue}{rgb}{0.07,0.07,0.488}
-\hypersetup{colorlinks=true,linkcolor=slateblue,anchorcolor=slateblue,citecolor=slateblue,filecolor=slateblue,urlcolor=slateblue,bookmarksnumbered=true,pdfview=FitB}
 \usepackage{url}
@@
 
-The survey also contains a respondent id variable.  4510 people in sample,  1271 people interviewed only once, 1107 interviewed every year.
-
 \subsection{Factorising probabilities}
@@
 \begin{itemize}
-  \item {\bf Barchart}: (bars, v/h, width)
-  \item {\bf Histogram}: (bars, v/h, 0)
-  \item {\bf Stacked} barchart + (bars, h/v, zero)
-  \item {\bf Side by side}: (bins, v/h, zero) + barchart  (or possibly spines with width prop to number of bars inside) - that would  give different scales, or use nesting operator with standard barchart
-  \item {\bf Spineplot}: (spines, v/h, width)
-  \item {\bf Stacked} spineplot: spineplot + (bars, h/v, 0)
-  \item {\bf Spinogram}: (spines, v/h, 0)
-  \item {\bf Fluctuation} diagram: (flucts, both, total)
-  \item {\bf Stacked} fluctuation diagram: fluctuation + (spines, v/h, zero)
-  \item {\bf Mosaic} plots: (spines, v/h, width) + (bar, h/v, width) + ... (alternating)
-  \item {\bf Double-decker} plot: 
+  \item {\bf Stacked} barchart: 1 bar + $n-1$ spines in opposite direction.
+  \item {\bf Nested} barchart: $n$ bars.
+  \item {\bf Mosaic} plot: spines in alternating directions.  The 1d case has a special name, the spineplot.
+  \item {\bf Double-decker} plot: $n-1$ spines + 1 spine in opposite direction.
+
+  \item {\bf Fluctuation} diagram: 
 \end{itemize}
@@
 Pie charts fall out naturally, as bars in polar coordinates, with angle and radius instead of height and width.
-Corrections to calculation of area.
-.
+And to ensure that counts stay proportional to areas, the square-root is taken of the y-axis.
 This is related to the infoslices of \citet{andrews:1998}, which only use half of the disk.
 Figure~\ref{fig:polar} shows some examples of product plots in polar coordinates.
+
+\begin{itemize}
+  \item Pie charts
+  \item Fourfold displays \citep{friendly:1995}
+  \item Bullseye chart
+\end{itemize}
+
+\begin{figure}[htbp]
+  \centering
+    \includegraphics[width=0.25\linewidth]{hb-vb-cartesian}%
+    \includegraphics[width=0.25\linewidth]{hb-vs-cartesian}%
+    \includegraphics[width=0.25\linewidth]{hs-hs-cartesian}%
+    \includegraphics[width=0.25\linewidth]{hs-vs-cartesian}
+
+    \includegraphics[width=0.25\linewidth]{hb-vb-polar}%
+    \includegraphics[width=0.25\linewidth]{hb-vs-polar}%
+    \includegraphics[width=0.25\linewidth]{hs-hs-polar}%
+    \includegraphics[width=0.25\linewidth]{hs-vs-polar}
+  \caption{}
+  \label{fig:polar}
+\end{figure}
 
@@
 
+Labelling these plots is particularly challenging.
+
 \begin{itemize}
   \item Additional info 
-  \begin{itemise}
+  \begin{itemize}
     \item colour (map of the market)
@@
     \item text (tables)
-  \end{itemise}
+  \end{itemize}
 

### 2009-05-13 plots.r

@@ -1,0 +1,35 @@
+# Polar coordinates
+
+h <- read.csv("../happy.csv")
+setwd("../code"); source("divide.r"); setwd("../paper")
+
+shade <- list(
+  aes(fill = sex),
+  opts(legend.position = "none", 
+    axis.text.x = theme_blank(),
+    axis.text.y = theme_blank()
+  ), 
+  xlab(NULL),
+  ylab(NULL),
+  scale_y_sqrt()
+)
+
+prodplot(h, ~ happy + sex) + shade
+ggsave("plots/hs-vs-cartesian.pdf", width = 4, height = 4)
+last_plot() + coord_polar()
+ggsave("plots/hs-vs-polar.pdf", width = 4, height = 4)
+
+prodplot(h, ~ happy + sex, divider = nested()) + shade
+ggsave("plots/hb-vb-cartesian.pdf", width = 4, height = 4)
+last_plot() + coord_polar()
+ggsave("plots/hb-vb-polar.pdf", width = 4, height = 4)
+
+prodplot(h, ~ happy + sex, divider = list(hspline, hspline)) + shade
+ggsave("plots/hs-hs-cartesian.pdf", width = 4, height = 4)
+last_plot() + coord_polar()
+ggsave("plots/hs-hs-polar.pdf", width = 4, height = 4)
+
+prodplot(h, ~ happy + sex, divider = stacked()) + shade
+ggsave("plots/hb-vs-cartesian.pdf", width = 4, height = 4)
+last_plot() + coord_polar()
+ggsave("plots/hb-vs-polar.pdf", width = 4, height = 4)

### 2009-05-14 product-graphics.tex
Some ideas for continuous plots

@@
 
+Give a few examples of non-traditional continuous-categorical plots.  Fluctuation diagram for exploring joint 2d distribution.  Mosaic for conditional.  Age and year distribution.
+
+Cat + continuous.
+
 \subsection{Nested data}
@@
 \label{sec:display}
-
-The three plots above are special cases of $a(i, j) = w(i, j)^x * h(i, j)^y$.  $x + y = 1$.
 
@@
 
-Impossible to simultaneously satisfy all constraints.  Every product plot requires the creator to trade off between these.
+Impossible to simultaneously satisfy all constraints.
+Every product plot requires the creator to trade off between these.
+We will focus our attention on rectangular partitions.
+Special cases of $a(i, j) = w(i, j)^x * h(i, j)^y$,   $x + y = 1$.
 
@@
 \section{Connection with log-linear models}
-\label{sec:models}
-
-which imply the obvious mapping between d(i, j) and a(i, j).  For >2d,
-the definitions generalise in the obvious, but the constraints become
-more complicated.
-
-$log(d(i,j)) = 1 + d_{i \cdot} + d_{j|i}$
-$log(d(i,j)) = 1 + d_{\cdot j} + d_{i|j}$
 

### 2009-05-14 plots-cont.r

@@ -1,0 +1,14 @@
+h <- read.csv("../happy.csv")
+setwd("../code"); source("divide.r"); setwd("../paper")
+
+qplot(age, year, data = h)
+ggsave("plots/age-year.pdf", width = 4, height = 4)
+
+h$age_b <- round_any(h$age, 5)
+h$year_b <- round_any(h$year, 5)
+
+prodplot(h, ~ year_b + age_b) + aes(fill = age_b)
+prodplot(h, ~  age_b, ~ year_b) + aes(fill = age_b)
+prodplot(h, ~  age_b, ~ year_b, divider = list(hspline, vbar)) + aes(fill = age_b)
+
+prodplot(h, ~  age_b, divider = list(vbar)) + aes(fill = age_b)

### 2009-05-14 plots-cont.r
Better loading system + rearrangement

@@ -1,5 +1,4 @@
-h <- read.csv("../happy.csv")
-setwd("../code"); source("divide.r"); setwd("../paper")
+source("../load.r")
 
 qplot(age, year, data = h)
 ggsave("plots/age-year.pdf", width = 4, height = 4)

### 2009-05-14 plots-polar.r

@@ -1,7 +1,6 @@
 # Polar coordinates
 
-h <- read.csv("../happy.csv")
-setwd("../code"); source("divide.r"); setwd("../paper")
+source("../load.r")
 
 shade <- list(
   aes(fill = sex),

### 2009-11-25 plots-cont.r
More work, including start on unit tests.

@@ -10,4 +10,9 @@ prodplot(h, ~ year_b + age_b) + aes(fill = age_b)
 prodplot(h, ~  age_b, ~ year_b) + aes(fill = age_b)
 prodplot(h, ~  age_b, ~ year_b, divider = list(hspline, vbar)) + aes(fill = age_b)
 
+
+prodplot(h, ~ year_b + age_b, divider = list(fluct))
+prodplot(h, ~ age_b, ~ year_b, divider = list(fluct))
+prodplot(h, ~ year_b, ~ age_b, divider = list(fluct))
+
 prodplot(h, ~  age_b, divider = list(vbar)) + aes(fill = age_b)

### 2009-11-26 product-graphics.tex
More work, including start on unit tests.
Starting to thrash the paper into some sort of shape

@@
 
-The Grammar of Graphics [reference] is a grammar describing the components of statistical graphics and how they can be combined together.
-These components are very flexible and can be used to describe any type of plot.
-However, the book (Grammar with a capital G) does not present many examples of graphics for categorical data, and in those examples the use of the grammar gives little insight into plot.
-This paper describes common categorical plots, and presents a specialisation of the Grammar of Graphics specifically designed to describe these types of plots and give insight into the types of relationships that they can show.
+% JASA methods
+%
+
+% The Grammar of Graphics [reference] describes the independent components of
+% statistical graphics and how they can be combined together. These components
+% are very flexible and can be used to describe any type of plot. However, the
+% book (Grammar with a capital G) does not present many examples of graphics
+% for categorical data, and in those examples the use of the grammar gives
+% little insight into plot. This paper describes common categorical plots, and
+% presents a specialisation of the Grammar of Graphics specifically designed
+% to describe these types of plots and give insight into the types of
+% relationships that they can show.
 
@@
 
-Area plots are the graphical equivalent to contingency tables.
-And like contingency tables, they can be used to display raw counts, or to display proportions or percentages.
-Similarly, it may be useful to augment area plots with marginal displays as we augment contingency tables with marginal sums.
-
-Area plots different to other types of graphics, because must be able to flexible change level of aggregation.  Values are typically counts, but can also be sums of observation level weights.
-
-These techniques can also be used with continuous data, if the data is binned to create a categorical variable.
-This gives rise to the histogram and spinogram, analogues of the bar chart and spine chart respectively.
-A long standing tradition is that no gaps are displayed between adjacent rectangles when used for originally continuous data.
-Two examples are shown in Figure \ref{fig:cont-examples}.
-
-\begin{figure}[htbp]
-  \begin{center}
-  \end{center}
-  \caption{Area plots of originally continuous data}
-  \label{fig:cont-examples}
-\end{figure}
-
-These techniques are also closely related to TreeMaps and dimensional stacking \citep{leblanc:1990}
-
-What are the common components of all these plots, and how can be describe them in a succinct way that also allows us to describe what tasks each display is good for?
-We propose that there is one key feature that makes these plots special.
-That is there their coordinate system, which is hierarchical.
-
-In most coordinate systems that we are used to, such as Cartesian or polar coordinates, are symmetric in the sense that you can locate a point by starting from either dimension.
-For example, in Cartesian coordinates you can identify either the x- or y-coordinate first, and the other next.
-This also implies a distance symmetry when projected onto 2D Euclidean geometry.
-When you vary one coordinate by a small amount (holding all others constant) points in the projective geometry will be close.
-
-Neither of these properties hold for hierarchical coordinate systems.
-
-We will discuss hierachical coordinate systems in general, and then present a specific hierachical coordinate system which gives rise to the area plots described previously.
-We also present an R implementation, in the R package {\tt ggplot}.
-
-Related work:
-
-\citep{vliegen:2006}
-\citep{ahlberg:1996}
-
-\citep{hartigan:1984,hartigan:1981}
-\citep{friendly:1999,friendly:1994}
-\citep{theus:1999}
-\citep{hofmann:2000,hofmann:2001,hofmann:2003}
-
-Goals:
-
 \begin{itemize}
@@
 
-\section{Data}
-\label{sec:data}
-
-To illustrate the techniques in this paper, we'll use a small sample from the general social survey ({\sc gss}) focussing on happiness.
-51\,020 observations from a yearly survey from 1972 to 2006.
-Happy, age, sex, marital status, terminal education level, relative financial status and health.
-Also contains a weighting variable wtssall, which is a probability weight from the survey.
-
-\subsection{Factorising probabilities}
+Area plots are the graphical equivalent of contingency tables.
+Like contingency tables, they can be used to display raw counts, or conditioned and marginal breakdowns.
+Similarly, it may be useful to augment area plots with marginal displays as we augment contingency tables with marginal sums.
+We can decompose these graphics in a similar way to the way we decompose products of probabilities.
 
@@
 
-Compare to area of mosaic plot.
+These techniques are also closely related to dimensional stacking \citep{leblanc:1990}
 
-One other special 2d case is the equal bin size plot - which all 1d x 1d primitive plots collapse down to when counts are equal.
-The important difference is the formatting - a grid rather than nested boxes.
+Generalised treemaps: \citep{vliegen:2006}
+Cocktailmaps: \citep{ahlberg:1996}
+Mosaic plots: \citep{hartigan:1984,hartigan:1981,friendly:1994,hofmann:2003}
+Extension to mosaic: \citep{friendly:1999,hofmann:2000}
+Connection to models: \citep{theus:1999,hofmann:2001}
 
-\subsection{Weighting}
-\label{sub:weighting}
+% Hierarchical coordinate systems 
+%
+% What are the common components of all these plots, and how can be describe
+% them in a succinct way that also allows us to describe what tasks each
+% display is good for?  We propose that there is one key feature that makes
+% these plots special.  That is there their coordinate system, which is
+% hierarchical.  
+%
+% In most coordinate systems that we are used to, such as Cartesian or polar
+% coordinates, are symmetric in the sense that you can locate a point by
+% starting from either dimension.  For example, in Cartesian coordinates you
+% can identify either the x- or y-coordinate first, and the other next.  This
+% also implies a distance symmetry when projected onto 2D Euclidean geometry.
+% When you vary one coordinate by a small amount (holding all others constant)
+% points in the projective geometry will be close.
+%
+% Neither of these properties hold for hierarchical coordinate systems. 
+%
+% We will discuss hierachical coordinate systems in general, and then present
+% a specific hierachical coordinate system which gives rise to the area plots
+% described previously.  We also present an R implementation, in the R package
+% {\tt ggplot}.
 
-We have assumed the probabilities represent counts, but WLOG we can use any set of non-negative additive weights instead.
-Some of the applications of such weighted plots are described in \citet{unwin:2007}.
-
-\subsection{Continuous data}
-\label{sub:continuous_data}
-
-Binning.
-
-Convention when using binned continuous data is to remove the gap between regions.
-
-Give a few examples of non-traditional continuous-categorical plots.  Fluctuation diagram for exploring joint 2d distribution.  Mosaic for conditional.  Age and year distribution.
-
-Cat + continuous.
-
-\subsection{Nested data}
-\label{sub:nested_data}
+To illustrate the techniques in this paper, we'll use a small sample from the general social survey ({\sc gss}) focussing on happiness.
+51\,020 observations from a yearly survey from 1972 to 2006.
+Happy, age, sex, marital status, terminal education level, relative financial status and health.
+Also contains a weighting variable wtssall, which is a probability weight from the survey.
 
@@
 
-Impossible to simultaneously satisfy all constraints.
-Every product plot requires the creator to trade off between these.
-We will focus our attention on rectangular partitions.
-Special cases of $a(i, j) = w(i, j)^x * h(i, j)^y$,   $x + y = 1$.
+Impossible to simultaneously satisfy all constraints.  Every product plot requires the creator to trade off between these.
 
-Restrictions on partitioning
+Constraints on partitioning
 
@@
   \item Area should be proportional to weight
-  \item Containment
-  \item Rectangular
-  \item Non-overlapping
+  \item Containment / Non-overlapping
 \end{itemize}
 
-Pseudo 2d, where you take a long 1d structure and wrap it into 2d.
+We will add one more constraint to reduce the space of possible
 
-Pseduo 1d, where you make a new variable from multiple variables.
+\begin{itemize}
+  \item Rectangular partitions
+\end{itemize}
+
+We will focus our attention on rectangular partitions.
+In general,  $h(i, j) = c(i, j)^x$ $w(i, j) = c(i, j)^y$, $x + y = 1$.
+Basic principle of all area plots is area should be proportional to probability.
+For the plots we are interested in, the shape used to represent the count is a rectangle, so that $a(i, j) = w(i, j) * h(i, j)$.
 
@@
 
-Basic principle of all area plots is area should be proportional to probability.
-For the plots we are interested in, the shape used to represent the count is a rectangle, so that $a(i, j) = w(i, j) * h(i, j)$.
 Each plot type constrains this relationship in a different way:
@@
   \item {\bf treemap}: no restrictions on height and width.
+At first glance, it may seem strange to include the treemap as a 1d primitive, but it lays out a single variable in 2d dimensions, just like the other approaches.
 
@@
 \begin{itemize}
-  \item {\bf fluct}: width and height proportional to square root of value. Flucts allow comparisons of (approximately) absolute numbers in two directions.
+  \item {\bf fluct}: width and height proportional to square root of value. Flucts allow comparisons of absolute numbers in two directions.
 
@@
 
+One special 2d case is the equal bin size plot - which all 1d x 1d primitives and the fluct collapse to when counts are equal.
+
 \subsection{Plot templates}
+
+Plot templates combine the simple primitives into well-known graphics:
 
@@
 
-\subsubsection{Small values and zeros}
-\subsubsection{Cascading}
+\subsubsection{Area not proportional to weight}
+
+It can be useful to violate the constraint that area should be proportional to weight to distinguish between zeros and very small values.
+A zero weight should have zero area, but giving it positive area can be useful so it's actually possible to see!
+Similarly for very small values which would otherwise occupy less than a pixel, useful to constain to minimal perceptible size.
+
+Antony's zooming, where you also truncate the maximum size.
+
+Image.
+
+\subsubsection{Overlapping rectangles}
 
@@
 In the cascaded treemap, each level is slightly offset from the one above to create a pseudo-3d perspective.
+
+Example
 
@@
 
+These displays generalise the fourfold displays of \citet{friendly:1995}.
+
 \begin{itemize}
   \item Pie charts
-  \item Fourfold displays \citep{friendly:1995}
+  \item Nightingale
   \item Bullseye chart 
@@
 
+\section{Margining}
+\label{sec:margining}
+
+Mathematically defining margining operation.
+
 \section{Connection with log-linear models}
+
+\section{Variations}
+\label{sec:variations}
+
+% Pseudo 2d, where you take a long 1d structure and wrap it into 2d.
+% 
+% Pseduo 1d, where you make a new variable from multiple variables.
+
+\subsection{Weighting}
+\label{sub:weighting}
+
+We have assumed the probabilities represent counts, but without loss of generality we can use any set of non-negative, additive weights instead.
+Some of the applications of such weighted plots are described in \citet{unwin:2007}.
+Include a brief summary here.
+
+\subsection{Continuous data}
+\label{sub:continuous_data}
+
+These techniques can also be used with continuous data, if the data is binned to create a categorical variable.
+This gives rise to the histogram and spinogram, analogues of the bar chart and spine chart respectively.
+A long standing tradition is that no gaps are displayed between adjacent rectangles when used for originally continuous data.
+Two examples are shown in Figure \ref{fig:cont-examples}.
+
+\begin{figure}[htbp]
+  \begin{center}
+  \end{center}
+  \caption{Area plots of originally continuous data}
+  \label{fig:cont-examples}
+\end{figure}
+
+Give a few examples of non-traditional continuous-categorical plots.  Fluctuation diagram for exploring joint 2d distribution.  Mosaic for conditional.  Age and year distribution.
+
+Cat + continuous.
+
+\subsection{Nested data}
+\label{sub:nested_data}
+
+We distinguish nested data from crossed data by the number of interactions with missing values.  Completely crossed data contains every combination of all variables.  Nested data contains
+
+Aka sparse vs. dense.
+
+It is not always possible to determine nested data from inspection of the data alone, but may need knowledge of the data collection.
+For example, if a survey of teachers within schools was performed, and each teacher was identified by a identifier unique within a school, it might look like the data was only missing a few combinations, when in fact it is meaningless.
+
+Or when you have a hierarchy.
+
+Example dataset
+
+Possibilities: create new variable which is interaction of nested vars, or drop missings/zeros.  Treemaps deal almost exclusively with nested data.
 
@@
 
-Labelling these plots is particularly challenging.
+Labelling these plots is particularly challenging.  Future work!
 
@@
     \item texture (sieve plots)
-    \item embedded plots (time series in lab escape)
     \item photographs
     \item text (tables)
+    \item embedded plots (time series in lab escape)
   \end{itemize}
@@
 
+\section{Conclusions} % (fold)
+\label{sec:conclusions}
+
+% section conclusions (end)
+
 % bibtool -x product-graphics -o references.bib

### 2009-11-26 product-graphics.tex
More work on paper

@@
 \title{Product graphics}
-\author{Heike Hofmann and Hadley Wickham}
+\author{Hadley Wickham and Heike Hofmann}
 \begin{document}
@@
 % JASA methods
+% Infovis?
+
+% Aims of this paper:
 %
+%  * emphasise the connection between area products and factorising 
+%    probabilities
+% 
+%  * pull all area plots on a common footing - provide a grammar for 
+%    categorical graphics
 
-% The Grammar of Graphics [reference] describes the independent components of
-% statistical graphics and how they can be combined together. These components
-% are very flexible and can be used to describe any type of plot. However, the
-% book (Grammar with a capital G) does not present many examples of graphics
-% for categorical data, and in those examples the use of the grammar gives
-% little insight into plot. This paper describes common categorical plots, and
-% presents a specialisation of the Grammar of Graphics specifically designed
-% to describe these types of plots and give insight into the types of
-% relationships that they can show.
+Area plots are the graphical equivalent of contingency tables, displaying how a total is broken down into components.
 
-Area plots [reference] display categorical data with a rectangle for each combination of the factors of interest, with the area of the rectangle proportional to the number of observations in that combination.
-The layout of the rectangles give rise to the difference categorical plots that we are familiar with: bar charts, spine charts, mosaic plots, equal-bin-size plots, and fluctuation diagrams.
+display categorical data with a rectangle for each combination of the factors of interest, with the area of the rectangle proportional to the number of observations in that combination.
+
+The layout of the rectangles give rise to the difference categorical plots that we are familiar with: bar charts, spine charts, mosaic plots \citep{hartigan:1984,hartigan:1981,friendly:1994,hofmann:2003},  equal-bin-size plots, and fluctuation diagrams.
 Trellis graphics are another type of display that uses categorical variables to create small multiples for different subsets of the data.
 Visually, trellis graphics look like equal-bin-size plots with another plot drawn inside each bin.
+
 Figure \ref{fig:cat-examples} illustrates some of these plots.
@@
 
-\begin{itemize}
-  \item Put all area plots (bar charts, spine plots, mosaic plots, fluctuation diagrams) on a common footing
-  \item Continue to build link between graphics and log-linear models
-  \item Provide tools for ``residuals and re-iteration''
-\end{itemize}
+Hierarchical partitioning of space.
 
@@
 
-Area plots are the graphical equivalent of contingency tables.
-Like contingency tables, they can be used to display raw counts, or conditioned and marginal breakdowns.
-Similarly, it may be useful to augment area plots with marginal displays as we augment contingency tables with marginal sums.
-We can decompose these graphics in a similar way to the way we decompose products of probabilities.
-
-Let $f(x, y, z)$ be the 2d pmf function generated.  Can write this pmf as product of marginals and conditionals
+Let $f(x, y, z)$ be the 3d pmf function.  We can can write this pmf as product of marginals and conditionals in many different ways.  We can decompose product graphics in a similar way.
 
@@
 
-Generalised treemaps: \citep{vliegen:2006}
-Cocktailmaps: \citep{ahlberg:1996}
-Mosaic plots: \citep{hartigan:1984,hartigan:1981,friendly:1994,hofmann:2003}
-Extension to mosaic: \citep{friendly:1999,hofmann:2000}
-Connection to models: \citep{theus:1999,hofmann:2001}
+% Other papers to cite:
+% 
+% Generalised treemaps: \citep{vliegen:2006}
+% Cocktailmaps: \citep{ahlberg:1996}
+% Extension to mosaic: \citep{friendly:1999,hofmann:2000}
+% Connection to models: \citep{theus:1999,hofmann:2001}
 
@@
 % a specific hierachical coordinate system which gives rise to the area plots
-% described previously.  We also present an R implementation, in the R package
-% {\tt ggplot}.
+% described previously.
 
-To illustrate the techniques in this paper, we'll use a small sample from the general social survey ({\sc gss}) focussing on happiness.
-51\,020 observations from a yearly survey from 1972 to 2006.
-Happy, age, sex, marital status, terminal education level, relative financial status and health.
-Also contains a weighting variable wtssall, which is a probability weight from the survey.
+To illustrate these ideas, we'll use a small sample from the general social survey ({\sc gss}) focussing on happiness. 51\,020 observations from cross-section survey given yearly from 1972 to 2006.
+
+\begin{itemize}
+  \item Happy.  Discrete: ``very happy'', ``pretty happy'', ``not too happy''
+  \item Age (in years), 18--89.
+  \item Sex: ``female'', ``male''
+  \item Degree: ``lt high school'', ``high school'', ``junior college'', ``bachelor'', ``gradudate''
+  \item Relative financial status {\sf finrela}: ``far above average'', ``above average'', ``average'', ``below average'', ``far below average''
+  \item Health: ``excellent'', ``good'', ``fair'', ``poor''
+  \item Year: 1972--2006.
+  \item Probability weight ({\sf wtsall}): 0.43--6.42.
+\end{itemize}
 

### 2009-11-26 happy.r
Reshuffle to get all paper related material in one directory

@@ -1,0 +1,35 @@
+h <- read.csv(file.path(PATH, "happy.csv"))
+h$degree <- factor(h$degree, 
+  levels = c("lt high school","high school","junior college", "bachelor",
+             "graduate"))
+
+# Mosaic plot of happiness and education
+prodplot(h, ~ degree + happy)
+# = 
+prodplot(h, ~ degree | happy) 
+# * 
+prodplot(h, ~ happy)
+
+# Mosiac plot of happiness, conditional on education
+prodplot(h, ~ happy + degree)
+# = 
+prodplot(h, ~ happy | degree)
+# * 
+prodplot(h, ~ degree)
+
+# Add some colour
+prodplot(h, ~ happy | degree) + aes(fill = happy)
+prodplot(h, ~ happy | degree) + aes(fill = degree)
+prodplot(h, ~ happy | degree) + aes(fill = .wt)
+prodplot(h, ~ happy | degree) + aes(fill = factor(level))
+
+# Use stacked bars instead of mosaic plot
+prodplot(h, ~ degree + happy, divider = stacked())
+prodplot(h, ~ happy, ~ degree, divider = stacked())
+
+# Use nested bars instead of mosaic plot
+prodplot(h, ~ degree + happy, divider = nested())
+prodplot(h, ~ happy, ~ degree, divider = nested())
+
+prodplot(h, ~ degree + happy, ~ sex)
+prodplot(h, ~ degree + happy, ~ sex, divider = ddecker())

### 2009-11-28 product-graphics.tex
Fleshing out more details.

@@
 %
-%  * emphasise the connection between area products and factorising 
+%  * emphasise the connection between area graphics and factorising 
 %    probabilities
@@
 
-display categorical data with a rectangle for each combination of the factors of interest, with the area of the rectangle proportional to the number of observations in that combination.
-
-The layout of the rectangles give rise to the difference categorical plots that we are familiar with: bar charts, spine charts, mosaic plots \citep{hartigan:1984,hartigan:1981,friendly:1994,hofmann:2003},  equal-bin-size plots, and fluctuation diagrams.
-Trellis graphics are another type of display that uses categorical variables to create small multiples for different subsets of the data.
-Visually, trellis graphics look like equal-bin-size plots with another plot drawn inside each bin.
+display categorical data with a polygon, typically a rectangle, for each combination of the factors of interest, with the area proportional to the number of observations in that combination.
+Because the total area for a graphic is usually constrained, an area plot serves to partition up this total, displaying proportions rather than counts.
 
@@
 \end{itemize}
+
+It can be useful to display partial decompositions, i.e. just $f(x, y | z)$, which is still a 3 variable pmf, but each category of z sums to one.
+
+Because of their hierarchical nature, an area plot of $f(x, y, z)$ also shows $f(y, z)$ and $f(z)$.
+
+By convention we will write the top most level of the product plot on the left, and lower levels to the right.
+While $f(x, y, z)$ and $f(z, y, z)$ display the same counts, their interpretation is different because of which marginal tables you see.
 
@@
 
-To illustrate these ideas, we'll use a small sample from the general social survey ({\sc gss}) focussing on happiness. 51\,020 observations from cross-section survey given yearly from 1972 to 2006.
+\section{Data}
+\label{sec:data}
+
+Two ways of representing data:
+
+\begin{itemize}
+  \item Multi-dimensional contingency table
+  \item Data frame, with column of counts
+\end{itemize}
+
+The two are equivalent, but it's usually easier to work with the data frame version, because variables are explicit.  Also deals better with sparsity and is common input to modelling algorithms.
+
+Define the conditioning operator which makes the distribution uniform.
+
+In this paper, we'll use a small sample from the general social survey ({\sc gss}) focussing on happiness. 51\,020 observations from cross-section survey given yearly from 1972 to 2006.
 
@@
 
+For recursive layout, need to be able to tile plane with space, and tile shape with shape.
+
 Impossible to simultaneously satisfy all constraints.  Every product plot requires the creator to trade off between these.
 
-Constraints on partitioning
+To simplify the problem, we'll initially work with three constraints on partitioning:
 
@@
   \item Containment / Non-overlapping
+  \item Rectangular partitions
 \end{itemize}
 
-We will add one more constraint to reduce the space of possible
+Section~\ref{sec:relax} discusses what happens when some of these constraints are relaxed.
 
 \begin{itemize}
-  \item Rectangular partitions
 \end{itemize}
@@
 
+\begin{figure}[htbp]
+  \centering
+    \includegraphics[width=0.25\linewidth]{part-vbar-vr}%
+    \includegraphics[width=0.25\linewidth]{part-hbar-vr}%
+    \includegraphics[width=0.25\linewidth]{part-vspline-vr}%
+    \includegraphics[width=0.25\linewidth]{part-hspline-vr}
+
+  \caption{1d partitions showing proportions 0.5, 0.2, 0.2 and 0.1. From left to right: vertical bars, horizontal bars, vertical splines and horizontal splines. }
+  \label{fig:part-1d}
+\end{figure}
+
 \begin{itemize}
-  \item {\bf bar}: height is proportional to value, width equally divides space. Bars allow comparison between absolute numbers.
+  \item {\bf bar}: height is proportional to value, width equally divides space.
+Bars allow comparison between absolute numbers, but are not space filling.
+They occupy sum(values / max(values)) of the total area.
 
@@
 
-Plot templates combine the simple primitives into well-known graphics:
+The sequence of partitioning choices defines the different named area plots:
 
@@
   \item {\bf Nested} barchart: $n$ bars.
-  \item {\bf Mosaic} plot: spines in alternating directions.  The 1d case has a special name, the spineplot.
+  \item {\bf Mosaic} plot: spines in alternating directions.  The 1d case has a special name, the spineplot. \citep{hartigan:1984,hartigan:1981,friendly:1994,hofmann:2003}
   \item {\bf Double-decker} plot: $n-1$ spines + 1 spine in opposite direction.
 
-  \item {\bf Fluctuation} diagram: 
+  \item {\bf Fluctuation} diagram: fluct, then spline. 
+  \item {\bf Equal bin size} plot: a fluctuation diagram, when the 2 variables have uniform distribution.
 \end{itemize}
 
-\subsection{Violation of constraints}
+Trellis graphics are another type of display that uses categorical variables to create small multiples for different subsets of the data.
+Visually, trellis graphics look like equal-bin-size plots with another plot drawn inside each bin.
 
-\subsubsection{Area not proportional to weight}
+\section{Relaxing constraints}
+\label{sec:relax}
+
+\subsection{Area not proportional to weight}
 
@@
 
-\subsubsection{Overlapping rectangles}
+\subsection{Overlapping rectangles}
 
@@
 In the cascaded treemap, each level is slightly offset from the one above to create a pseudo-3d perspective.
+This makes it easier to see all the levels of the hierarchy, not just the lowest level.
 
-Example
+Example.
 
-\subsubsection{Non-rectangular partitions}
+\subsection{Non-rectangular partitions}
 
@@
 This is related to the infoslices of \citet{andrews:1998}, which only use half of the disk.
-Figure~\ref{fig:polar} shows some examples of product plots in polar coordinates.
+Figure~\ref{fig:polar} illustrates a few product plots in polar coordinates.
+Variables are happiness (top level in four categories) and sex (nested, two levels).
 
@@
     \includegraphics[width=0.25\linewidth]{hs-vs-polar}
-  \caption{}
+  \caption{(Top) Area graphics in Cartesian coordinates.
+(Bottom) Area graphics in polar coordinates.
+From left to right: horizontal bar nested inside horizontal bar, vertical spline nested inside horizontal bar (stacked bar chart), horizontal spline nested in horizontal spline, horizontal spline nested in vertical spline (mosaic plot).}
   \label{fig:polar}
@@
 
-\section{Margining}
-\label{sec:margining}
-
-Mathematically defining margining operation.
-
-\section{Connection with log-linear models}
+% \section{Connection with log-linear models}
 
@@
 
-Example dataset
+Example dataset - Doug Bates teachers/schools?
 
 Possibilities: create new variable which is interaction of nested vars, or drop missings/zeros.  Treemaps deal almost exclusively with nested data.
+
+\section{Data analysis} % (fold)
+\label{sec:data_analysis}
+
+% section data_analysis (end)
 

### 2009-11-28 plots-partition.r

@@ -1,0 +1,20 @@
+eq <- data.frame(trt = letters[1:4], .wt = rep(0.25, 4))
+vr <- data.frame(trt = letters[1:4], .wt = c(0.5, 0.2, 0.2, 0.1))
+
+prodplot(eq, ~ trt, "vbar")
+ggsave("plots/part-vbar-eq.pdf", width = 4, height = 4)
+prodplot(vr, ~ trt, "vbar")
+ggsave("plots/part-vbar-vr.pdf", width = 4, height = 4)
+prodplot(eq, ~ trt, "hbar")
+ggsave("plots/part-hbar-eq.pdf", width = 4, height = 4)
+prodplot(vr, ~ trt, "hbar")
+ggsave("plots/part-hbar-vr.pdf", width = 4, height = 4)
+
+prodplot(eq, ~ trt, "vspline")
+ggsave("plots/part-vspline-eq.pdf", width = 4, height = 4)
+prodplot(vr, ~ trt, "vspline")
+ggsave("plots/part-vspline-vr.pdf", width = 4, height = 4)
+prodplot(eq, ~ trt, "hspline")
+ggsave("plots/part-hspline-eq.pdf", width = 4, height = 4)
+prodplot(vr, ~ trt, "hspline")
+ggsave("plots/part-hspline-vr.pdf", width = 4, height = 4)

### 2009-11-28 product-graphics.tex
Fleshing out factorising & partitions sections

@@
 
+\section{Factorising}
+
 Basic idea is factorising table of probabilities:
@@
 
+Based on the modelling specification of \citep{wilkinson:1973} we define an algebra for describing these tables.
+
+\begin{itemize}
+  \item \verb|~ a| represents $f(a)$,
+  \item \verb|~ a + b| represents $f(a, b)$
+  \item \verb!~ a | b! represents $f(a | b)$ which is equivalent to $f(a, b)$ but with the constraint that $\forall b \sum_a f(a | b) = 1$.
+\end{itemize}
+
 By convention we will write the top most level of the product plot on the left, and lower levels to the right.
-While $f(x, y, z)$ and $f(z, y, z)$ display the same counts, their interpretation is different because of which marginal tables you see.
+While $f(x, y, z)$ and $f(z, y, x)$ display the same counts, their interpretation is different because of which marginal tables you see.
+
+Figure~\ref{fig:fact-simple} illustrates these ideas with the distribution of happiness and sex.
+We start with a plot where area broken down by happiness.
+We can then continue to factor this plot to also display sex - the area of each rectangle is proportional to the number of people who are happy with that sex, and the construction allows us to see the conditional distribution of sex within happy.
+In this case the breaks line up, suggesting that there is no difference in sex, given happiness.
+This could be followed up by a formula hypothesis test.
+
+Removing the marginal effect of happiness, as in the column on the far right, makes each of the top-level categories the same size, and makes it easier to compare the conditional distributions.
+It's not that important in this case, but in the bar chart below, it allows us to see the conditional distribution, and in the case when there are many levels with very different probabilities it can make the plot much easier to read.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.33\linewidth]{fact-happy}%
+  \includegraphics[width=0.33\linewidth]{fact-happy-sex}%
+  \includegraphics[width=0.33\linewidth]{fact-happy|sex}
+
+  \includegraphics[width=0.33\linewidth]{fact-happy-2}%
+  \includegraphics[width=0.33\linewidth]{fact-happy-sex-2}%
+  \includegraphics[width=0.33\linewidth]{fact-happy|sex-2}
+
+  \caption{Plots of the distribution of happiness and sex.}
+  \label{fig:fact-simple}
+\end{figure}
 
@@
 
-The two are equivalent, but it's usually easier to work with the data frame version, because variables are explicit.  Also deals better with sparsity and is common input to modelling algorithms.
+The two are equivalent, but it's usually computational easier to work with the data frame version, because variables are explicit, and mathematically easier to work with the multidimensional array because multiplication is nicely defined.
+Also deals better with sparsity and is common input to modelling algorithms.
 
@@
 
-Perceptual constraints.  Best at comparing area when:
+Once we have made the decision to map area to proportion, then we need to impose some constraints on the set of possible tilings of the plane.
+Some constraints are implied by our perceptual capabilities and others by simple geometry.
+(In general, we could also think of partitions of higher-dimensional spaces (e.g.
+3d or 4d), but we will need to project these down to two for viewing anyway, so we might as well work directly in the final space.)
 
 \begin{itemize}
-  \item Simple shape (i.e. square vs polygon)
-  \item Aspect ratios similar
-  \item Areas are large
-  \item Areas are close
-  \item One border is constant.
+  \item Area should be proportional to weight.  Imposed by problem statement.
+
+  \item Rectangular partitions.
+It is easier to compare the areas of simple shape (i.e.
+rectangle vs polygon), and even better when one the length of one border is fixed, so we only need compare the length of the other \citep{cleveland:1984}.
+Rectangular partitions are also recursive - a rectangular partition of a rectangular area produces rectangles.
+There are a limited number of shapes that do this - triangles yes, circles, no.
+
+  \item Containment / Non-overlapping.  To be able to see the complete area of each rectangle they must be non-overlapping, and each split must be nested within the shapes of the previous level.
+
 \end{itemize}
 
-For recursive layout, need to be able to tile plane with space, and tile shape with shape.
+Note, however, that each of these constraints can be profitably relaxed.  See Section~\ref{sec:relax} for details.
 
-Impossible to simultaneously satisfy all constraints.  Every product plot requires the creator to trade off between these.
+Let $a(i)$ give the area assigned to category $i$, which has proportion $f(i)$.
+$a(i) \propto f(i)$ or $a(i) = k \dot a(i)$, and $a(i) = w(i) \times h(i)$ where $w$ and $h$ represent width and height respectively.
+Most of the the partitions described below use additional constraints: $h(i) = \sqrt{k} f(i)^x$ and $w(i) = \sqrt{k} f(i)^y$ where $x + y = 1$.
+This ensures that it is only necessary to inspect one border to compare magnitudes.
 
-To simplify the problem, we'll initially work with three constraints on partitioning:
+The other chief difference between different tiling methods is whether or not they are space filling.
+Space filling methods will ensure that areas are larger, but typically make it harder to compare magnitudes because scales no longer align.
 
-\begin{itemize}
-  \item Area should be proportional to weight
-  \item Containment / Non-overlapping
-  \item Rectangular partitions
-\end{itemize}
-
-Section~\ref{sec:relax} discusses what happens when some of these constraints are relaxed.
-
-\begin{itemize}
-\end{itemize}
-
-We will focus our attention on rectangular partitions.
-In general,  $h(i, j) = c(i, j)^x$ $w(i, j) = c(i, j)^y$, $x + y = 1$.
-Basic principle of all area plots is area should be proportional to probability.
-For the plots we are interested in, the shape used to represent the count is a rectangle, so that $a(i, j) = w(i, j) * h(i, j)$.
+The following two sections discuss particular partitions of the plane based on either a 1d or 2d pmf.  Displays of higher dimensional pmfs must be made by recursive combinations of these.
 
@@
 
-Space-filling vs. not space-filling.  Read length on common scale, vs. put more information in the display.
-
-Each plot type constrains this relationship in a different way:
-
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=0.25\linewidth]{part-vbar-vr}%
-    \includegraphics[width=0.25\linewidth]{part-hbar-vr}%
-    \includegraphics[width=0.25\linewidth]{part-vspline-vr}%
-    \includegraphics[width=0.25\linewidth]{part-hspline-vr}
-
-  \caption{1d partitions showing proportions 0.5, 0.2, 0.2 and 0.1. From left to right: vertical bars, horizontal bars, vertical splines and horizontal splines. }
-  \label{fig:part-1d}
-\end{figure}
+There are three 1d primitives, as displayed in Figure~\ref{fig:part-1d}:
 
@@
 They occupy sum(values / max(values)) of the total area.
+Bars can be arranged horizontally (``hbar'') or vertically (``vbar'').
 
   \item {\bf spine}: width is proportional to value, height equally divides space.
-Spines not useful at top level, but allow comparison of proportions at next level (ie.
-conditioned on top level values)
+Splines are space filling, but make it more difficult to compare absolute counts.
+Splines can be arranged horizontally (``hspline'') or vertically (``vspline'').
 
   \item {\bf treemap}: no restrictions on height and width.
-At first glance, it may seem strange to include the treemap as a 1d primitive, but it lays out a single variable in 2d dimensions, just like the other approaches.
+Comparing the areas is the most difficult perceptually, and are most useful when there are a large number of categories.
+There are a many variations of the treemap; this paper uses the squarified treemap \citep{bruls:1999}.
 
 \end{itemize}
+
+\begin{figure}[htbp]
+  \centering
+    \includegraphics[width=0.33\linewidth]{part-hbar-vr}%
+    \includegraphics[width=0.33\linewidth]{part-hspline-vr}
+    % \includegraphics[width=0.25\linewidth]{part-hspline-vr}
+
+  \caption{1d partitions showing proportions 0.5, 0.2, 0.2 and 0.1. From left to right: vertical bars, horizontal bars, vertical splines and horizontal splines. }
+  \label{fig:part-1d}
+\end{figure}
 
@@
 
-\subsection{Overlapping rectangles}
+\subsection{Levels not contained}
 

### 2009-11-28 plots-factorising.r

@@ -1,0 +1,16 @@
+library(productplots)
+
+prodplot(happy, ~ happy, "hspline") + aes(fill = happy)
+ggsave("plots/fact-happy.pdf", width = 4, height = 4)
+prodplot(happy, ~ sex + happy, c("vspline", "hspline")) + aes(fill = sex)
+ggsave("plots/fact-happy-sex.pdf", width = 4, height = 4)
+prodplot(happy, ~ sex | happy, c("vspline", "hspline")) + aes(fill = sex)
+ggsave("plots/fact-happy|sex.pdf", width = 4, height = 4)
+
+
+prodplot(happy, ~ happy, "hbar") + aes(fill = happy)
+ggsave("plots/fact-happy-2.pdf", width = 4, height = 4)
+prodplot(happy, ~ sex + happy, c("vspline", "hbar")) + aes(fill = sex)
+ggsave("plots/fact-happy-sex-2.pdf", width = 4, height = 4)
+prodplot(happy, ~ sex | happy, c("vspline", "hbar")) + aes(fill = sex)
+ggsave("plots/fact-happy|sex-2.pdf", width = 4, height = 4)

### 2009-11-28 product-graphics.tex
More working

@@
 Splines are space filling, but make it more difficult to compare absolute counts.
-Splines can be arranged horizontally (``hspline'') or vertically (``vspline'').
+Splines can be arranged horizontally (``hspline''),  vertically (``vspline''), or can automatically pick their orientation (``spline'').
 
@@
   \centering
-    \includegraphics[width=0.33\linewidth]{part-hbar-vr}%
-    \includegraphics[width=0.33\linewidth]{part-hspline-vr}
-    % \includegraphics[width=0.25\linewidth]{part-hspline-vr}
+    \includegraphics[width=0.33\linewidth]{part-hbar}%
+    \includegraphics[width=0.33\linewidth]{part-hspline}
+    % \includegraphics[width=0.25\linewidth]{part-treemap}
 
-  \caption{1d partitions showing proportions 0.5, 0.2, 0.2 and 0.1. From left to right: vertical bars, horizontal bars, vertical splines and horizontal splines. }
+  \caption{1d partitions showing proportions 0.5, 0.2, 0.2 and 0.1. From left to right: horizontal bars, horizontal splines, and a treemap.}
   \label{fig:part-1d}
@@
 
-One special 2d case is the equal bin size plot - which all 1d x 1d primitives and the fluct collapse to when counts are equal.
+One special 2d case is the equal bin size plot - which all 1d x 1d primitives and the fluct collapse to when proportions are equal, as when conditioned.
+
+Figure~\ref{fig:fluct} shows these two behaviours, looking at the relationship between happiness, health and financial status.
+On the left, a display of the raw proportion shows that most people are in good health and average financial standing, but it is difficult to see how happiness varies because the size of the squares is variable.
+On the right, conditioning on financial status and health makes all squares the same size and makes it easier to see the conditional distribution of happiness - people tend to be happier the healthier and the richer they are
+
+\begin{figure}[htbp]
+  \centering
+    \includegraphics[width=0.5\linewidth]{part-fluct}%
+    \includegraphics[width=0.5\linewidth]{part-fluct-cond}
+  \caption{caption}
+  \label{fig:fluct}
+\end{figure}
 
@@
 
-The sequence of partitioning choices defines the different named area plots:
+The sequence of partitioning choices defines different named area plots:
 
 \begin{itemize}
-  \item {\bf Stacked} barchart: 1 bar + $n-1$ spines in opposite direction.
-  \item {\bf Nested} barchart: $n$ bars.
-  \item {\bf Mosaic} plot: spines in alternating directions.  The 1d case has a special name, the spineplot. \citep{hartigan:1984,hartigan:1981,friendly:1994,hofmann:2003}
-  \item {\bf Double-decker} plot: $n-1$ spines + 1 spine in opposite direction.
+  \item {\bf Barchart} (1d). 1 hbar.
+  \item {\bf Column chart} (1d). 1 vbar.
+  \item {\bf Spineplot} (1d). 1 vspline.
 
-  \item {\bf Fluctuation} diagram: fluct, then spline. 
-  \item {\bf Equal bin size} plot: a fluctuation diagram, when the 2 variables have uniform distribution.
+  \item {\bf Stacked} barchart (2d). 1 hbar and 1 vspine.
+  \item {\bf Nested} barchart (2d).  2 hbars. \citep{peltier:2009}
+  \item {\bf Fluctuation} diagram (2d): fluct.
+
+  \item {\bf Equal bin size} plot (3d): fluct and vspline, provided the final two variables are uniform.
+
+  \item {\bf Mosaic} plot (nd).  Alternating hspines and vspines.  \citep{hartigan:1984,hartigan:1981,friendly:1994,hofmann:2003}
+  \item {\bf Double-decker} plot (nd).  $n-1$ hspines and 1 vspine. \citep{hofmann:2001}
+  \item {\bf Treemap} (nd): n spines. \citep{shneiderman:1992}
+  \item {\bf Squarified treemap} (nd): n squares. \citep{bruls:1999}
+
+\end{itemize}
+
+% http://peltiertech.com/Utility/ClusterStackUtility.html
+
+\begin{itemize}
+
 \end{itemize}

### 2009-11-28 colours.r

@@ -1,0 +1,9 @@
+# Plots should use a consistent colour scheme for happiness so the legend
+# only needs to be displayed once.  Currently there seems to be a bug in 
+# scale_fill_brewer which is messing things up.
+
+happy_fill <- list(
+  aes(fill = happy), 
+  scale_fill_brewer("Happiness", pal = "Blues", 
+    limits = c("not too happy", "happy", "very happy")),
+  opts(legend.position = "none"))

### 2009-11-28 plots-partition.r

@@ -1,20 +1,27 @@
-eq <- data.frame(trt = letters[1:4], .wt = rep(0.25, 4))
 vr <- data.frame(trt = letters[1:4], .wt = c(0.5, 0.2, 0.2, 0.1))
 
-prodplot(eq, ~ trt, "vbar")
-ggsave("plots/part-vbar-eq.pdf", width = 4, height = 4)
-prodplot(vr, ~ trt, "vbar")
-ggsave("plots/part-vbar-vr.pdf", width = 4, height = 4)
-prodplot(eq, ~ trt, "hbar")
-ggsave("plots/part-hbar-eq.pdf", width = 4, height = 4)
 prodplot(vr, ~ trt, "hbar")
-ggsave("plots/part-hbar-vr.pdf", width = 4, height = 4)
+ggsave("plots/part-hbar.pdf", width = 4, height = 4)
 
-prodplot(eq, ~ trt, "vspline")
-ggsave("plots/part-vspline-eq.pdf", width = 4, height = 4)
-prodplot(vr, ~ trt, "vspline")
-ggsave("plots/part-vspline-vr.pdf", width = 4, height = 4)
-prodplot(eq, ~ trt, "hspline")
-ggsave("plots/part-hspline-eq.pdf", width = 4, height = 4)
 prodplot(vr, ~ trt, "hspline")
-ggsave("plots/part-hspline-vr.pdf", width = 4, height = 4)
+ggsave("plots/part-hspline.pdf", width = 4, height = 4)
+
+prodplot(vr, ~ trt, "treemap")
+ggsave("plots/part-treemap.pdf", width = 4, height = 4)
+
+
+# 2d partition
+prodplot(happy, ~ happy + finrela, c("vspline", "hspline")) + 
+  aes(fill = finrela)
+prodplot(happy, ~ happy + health, c("vspline", "hspline")) + 
+  aes(fill = health)
+prodplot(happy, ~ happy + marital, c("vspline", "hspline")) + 
+  aes(fill = marital)
+
+prodplot(happy, ~ happy + finrela + health, c("vspline", "fluct"), 
+  colour = NA, na.rm = TRUE) + aes(fill = happy)
+ggsave("plots/part-fluct.pdf", width = 6, height = 6)
+prodplot(happy, ~ happy | finrela + health, c("vspline", "fluct"), 
+  colour = NA, na.rm = TRUE) + aes(fill = happy)
+ggsave("plots/part-fluct-cond.pdf", width = 6, height = 6)
+

### 2009-11-30 plots-cont.r
Convert spline to spine

@@ -8,7 +8,7 @@ h$year_b <- round_any(h$year, 5)
 
 prodplot(h, ~ year_b + age_b) + aes(fill = age_b)
 prodplot(h, ~  age_b, ~ year_b) + aes(fill = age_b)
-prodplot(h, ~  age_b, ~ year_b, divider = list(hspline, vbar)) + aes(fill = age_b)
+prodplot(h, ~  age_b, ~ year_b, divider = list(hspine, vbar)) + aes(fill = age_b)
 
 
 prodplot(h, ~ year_b + age_b, divider = list(fluct))

### 2009-11-30 plots-factorising.r

@@ -1,16 +1,16 @@
 library(productplots)
 
-prodplot(happy, ~ happy, "hspline") + aes(fill = happy)
+prodplot(happy, ~ happy, "hspine") + aes(fill = happy)
 ggsave("plots/fact-happy.pdf", width = 4, height = 4)
-prodplot(happy, ~ sex + happy, c("vspline", "hspline")) + aes(fill = sex)
+prodplot(happy, ~ sex + happy, c("vspine", "hspine")) + aes(fill = sex)
 ggsave("plots/fact-happy-sex.pdf", width = 4, height = 4)
-prodplot(happy, ~ sex | happy, c("vspline", "hspline")) + aes(fill = sex)
+prodplot(happy, ~ sex | happy, c("vspine", "hspine")) + aes(fill = sex)
 ggsave("plots/fact-happy|sex.pdf", width = 4, height = 4)
 
 
 prodplot(happy, ~ happy, "hbar") + aes(fill = happy)
 ggsave("plots/fact-happy-2.pdf", width = 4, height = 4)
-prodplot(happy, ~ sex + happy, c("vspline", "hbar")) + aes(fill = sex)
+prodplot(happy, ~ sex + happy, c("vspine", "hbar")) + aes(fill = sex)
 ggsave("plots/fact-happy-sex-2.pdf", width = 4, height = 4)
-prodplot(happy, ~ sex | happy, c("vspline", "hbar")) + aes(fill = sex)
+prodplot(happy, ~ sex | happy, c("vspine", "hbar")) + aes(fill = sex)
 ggsave("plots/fact-happy|sex-2.pdf", width = 4, height = 4)

### 2009-11-30 plots-partition.r

@@ -3,25 +3,25 @@ vr <- data.frame(trt = letters[1:4], .wt = c(0.5, 0.2, 0.2, 0.1))
 prodplot(vr, ~ trt, "hbar")
 ggsave("plots/part-hbar.pdf", width = 4, height = 4)
 
-prodplot(vr, ~ trt, "hspline")
-ggsave("plots/part-hspline.pdf", width = 4, height = 4)
+prodplot(vr, ~ trt, "hspine")
+ggsave("plots/part-hspine.pdf", width = 4, height = 4)
 
 prodplot(vr, ~ trt, "treemap")
 ggsave("plots/part-treemap.pdf", width = 4, height = 4)
 
 
 # 2d partition
-prodplot(happy, ~ happy + finrela, c("vspline", "hspline")) + 
+prodplot(happy, ~ happy + finrela, c("vspine", "hspine")) + 
   aes(fill = finrela)
-prodplot(happy, ~ happy + health, c("vspline", "hspline")) + 
+prodplot(happy, ~ happy + health, c("vspine", "hspine")) + 
   aes(fill = health)
-prodplot(happy, ~ happy + marital, c("vspline", "hspline")) + 
+prodplot(happy, ~ happy + marital, c("vspine", "hspine")) + 
   aes(fill = marital)
 
-prodplot(happy, ~ happy + finrela + health, c("vspline", "fluct"), 
+prodplot(happy, ~ happy + finrela + health, c("vspine", "fluct"), 
   colour = NA, na.rm = TRUE) + aes(fill = happy)
 ggsave("plots/part-fluct.pdf", width = 6, height = 6)
-prodplot(happy, ~ happy | finrela + health, c("vspline", "fluct"), 
+prodplot(happy, ~ happy | finrela + health, c("vspine", "fluct"), 
   colour = NA, na.rm = TRUE) + aes(fill = happy)
 ggsave("plots/part-fluct-cond.pdf", width = 6, height = 6)
 

### 2009-11-30 plots-polar.r

@@ -23,7 +23,7 @@ ggsave("plots/hb-vb-cartesian.pdf", width = 4, height = 4)
 last_plot() + coord_polar()
 ggsave("plots/hb-vb-polar.pdf", width = 4, height = 4)
 
-prodplot(h, ~ happy + sex, divider = list(hspline, hspline)) + shade
+prodplot(h, ~ happy + sex, divider = list(hspine, hspine)) + shade
 ggsave("plots/hs-hs-cartesian.pdf", width = 4, height = 4)
 last_plot() + coord_polar()
 ggsave("plots/hs-hs-polar.pdf", width = 4, height = 4)

### 2009-11-30 product-graphics.tex
Minor updates

@@
 Splines are space filling, but make it more difficult to compare absolute counts.
-Splines can be arranged horizontally (``hspline''),  vertically (``vspline''), or can automatically pick their orientation (``spline'').
+Splines can be arranged horizontally (``hspine''),  vertically (``vspine''), or can automatically pick their orientation (``spine'').
 
@@
     \includegraphics[width=0.33\linewidth]{part-hbar}%
-    \includegraphics[width=0.33\linewidth]{part-hspline}
+    \includegraphics[width=0.33\linewidth]{part-hspine}
     % \includegraphics[width=0.25\linewidth]{part-treemap}
 
-  \caption{1d partitions showing proportions 0.5, 0.2, 0.2 and 0.1. From left to right: horizontal bars, horizontal splines, and a treemap.}
+  \caption{1d partitions showing proportions 0.5, 0.2, 0.2 and 0.1. From left to right: horizontal bars, horizontal spines, and a treemap.}
   \label{fig:part-1d}
@@
   \item {\bf Column chart} (1d). 1 vbar.
-  \item {\bf Spineplot} (1d). 1 vspline.
+  \item {\bf Spineplot} (1d). 1 vspine.
 
@@
 
-  \item {\bf Equal bin size} plot (3d): fluct and vspline, provided the final two variables are uniform.
+  \item {\bf Equal bin size} plot (3d): fluct and vspine, provided the final two variables are uniform.
 
@@
 % http://peltiertech.com/Utility/ClusterStackUtility.html
-
-\begin{itemize}
-
-\end{itemize}
 
@@
 (Bottom) Area graphics in polar coordinates.
-From left to right: horizontal bar nested inside horizontal bar, vertical spline nested inside horizontal bar (stacked bar chart), horizontal spline nested in horizontal spline, horizontal spline nested in vertical spline (mosaic plot).}
+From left to right: horizontal bar nested inside horizontal bar, vertical spine nested inside horizontal bar (stacked bar chart), horizontal spine nested in horizontal spine, horizontal spine nested in vertical spine (mosaic plot).}
   \label{fig:polar}

### 2009-12-01 product-graphics.tex
Improving display section

@@
 
-It can be useful to display partial decompositions, i.e. just $f(x, y | z)$, which is still a 3 variable pmf, but each category of z sums to one.
+It can be useful to display partial decompositions, i.e. just $f(x, y | z)$, which is still a 3 variable pmf, and is defined as
+
+\[ f(x, y | z) = \frac{f(x, y, z)}{\sum_{x = a, y = b} f(a, b, z)} \]
+
+so that each category of z sums to one, and for any fixed value of z, $f(x, y | z)$ is a valid pmf.
+
+\[ \forall b \sum_a f(a | b) = 1 \]
 
@@
 
-Based on the modelling specification of \citep{wilkinson:1973} we define an algebra for describing these tables.
+Based on the modelling specification of \citet{wilkinson:1973} we define an algebra for describing these tables.
 
@@
   \item \verb|~ a + b| represents $f(a, b)$
-  \item \verb!~ a | b! represents $f(a | b)$ which is equivalent to $f(a, b)$ but with the constraint that $\forall b \sum_a f(a | b) = 1$.
+  \item \verb!~ a | b! represents $f(a | b)$
 \end{itemize}
@@
 
-Once we have made the decision to map area to proportion, then we need to impose some constraints on the set of possible tilings of the plane.
-Some constraints are implied by our perceptual capabilities and others by simple geometry.
-(In general, we could also think of partitions of higher-dimensional spaces (e.g.
-3d or 4d), but we will need to project these down to two for viewing anyway, so we might as well work directly in the final space.)
+Once we have made the decision to map area to proportion, we need to impose some constraints on the set of possible tilings of the plane.
+In general, we could also think of partitions of higher-dimensional spaces (e.g.
+3d or 4d), but we will need to project these down to two for viewing anyway, so there is little disadvantage to working in 2d.
+
+Some constraints are implied by our perceptual capabilities and others by geometry. On top of the constraint that area must be proportional to value, we add two additional constraints:
 
 \begin{itemize}
-  \item Area should be proportional to weight.  Imposed by problem statement.
 
-  \item Rectangular partitions.
+  \item Partitions must be rectangular.
 It is easier to compare the areas of simple shape (i.e.
 rectangle vs polygon), and even better when one the length of one border is fixed, so we only need compare the length of the other \citep{cleveland:1984}.
-Rectangular partitions are also recursive - a rectangular partition of a rectangular area produces rectangles.
-There are a limited number of shapes that do this - triangles yes, circles, no.
+Rectangular partitions are also recursive in the sense that we can always tile a rectangle with smaller rectangles.
+This is not true for most shapes.
 
-  \item Containment / Non-overlapping.  To be able to see the complete area of each rectangle they must be non-overlapping, and each split must be nested within the shapes of the previous level.
+  \item Partitions must be disjoint.
+To be able to see the complete area of each rectangle they must be non-overlapping, and each split must be nested within the shapes of the previous level.
+Note that we do not impose the constraint that the tiling be space-filling.
+Space filling methods use will ensure that areas are larger, but make it harder to compare magnitudes because scales no longer align.
 
@@
 Let $a(i)$ give the area assigned to category $i$, which has proportion $f(i)$.
-$a(i) \propto f(i)$ or $a(i) = k \dot a(i)$, and $a(i) = w(i) \times h(i)$ where $w$ and $h$ represent width and height respectively.
-Most of the the partitions described below use additional constraints: $h(i) = \sqrt{k} f(i)^x$ and $w(i) = \sqrt{k} f(i)^y$ where $x + y = 1$.
-This ensures that it is only necessary to inspect one border to compare magnitudes.
+Since we are using rectangles, $a(i) = k \dot a(i) = k \dot w(i) \times h(i)$ where $w$ and $h$ represent width and height respectively.
+Certain partitions also satisfy the constraint that $h(i) = \sqrt{k} f(i)^x$ and $w(i) = \sqrt{k} f(i)^y$ where $x + y = 1$.
+This is useful because it is only necessary to inspect one border to compare magnitudes.
 
-The other chief difference between different tiling methods is whether or not they are space filling.
-Space filling methods will ensure that areas are larger, but typically make it harder to compare magnitudes because scales no longer align.
-
-The following two sections discuss particular partitions of the plane based on either a 1d or 2d pmf.  Displays of higher dimensional pmfs must be made by recursive combinations of these.
+The following two sections describe the partitions currently used to produce area graphics.
+Section~\ref{sub:part-1d} discusses partitions for 1d pmfs: bars, spines and tiles, Section~\ref{sub:part-2d} discusses one partition for 2d pmfs and Section~\ref{sub:part-nd} describes how the 1d and 2d primitives can be recursively combined to generate a partitioning for any higher dimension pmf and discusses how existing area graphics fall into this framework.
 
 \subsection{1d primitives}
+\label{sub:part-1d}
 
-There are three 1d primitives, as displayed in Figure~\ref{fig:part-1d}:
+There are three 1d primitives, as shown in Figure~\ref{fig:part-1d}:
 
 \begin{itemize}
-  \item {\bf bar}: height is proportional to value, width equally divides space.
+  \item {\bf bar}s: height is proportional to value, width equally divides space.
 Bars allow comparison between absolute numbers, but are not space filling.
@@
 Bars can be arranged horizontally (``hbar'') or vertically (``vbar'').
+Not space filling.
 
-  \item {\bf spine}: width is proportional to value, height equally divides space.
+  \item {\bf spine}s: width is proportional to value, height equally divides space.
 Splines are space filling, but make it more difficult to compare absolute counts.
 Splines can be arranged horizontally (``hspine''),  vertically (``vspine''), or can automatically pick their orientation (``spine'').
+Space filling.
 
-  \item {\bf treemap}: no restrictions on height and width.
+  \item {\bf tile}s: tile the plane with rectangles, trying to keep the aspect ratio of each rectangle close to 1.
 Comparing the areas is the most difficult perceptually, and are most useful when there are a large number of categories.
-There are a many variations of the treemap; this paper uses the squarified treemap \citep{bruls:1999}.
+The is partitioning used by the squarified treemap \citep{bruls:1999}.
+Not space filling.
 
@@
     \includegraphics[width=0.33\linewidth]{part-hbar}%
-    \includegraphics[width=0.33\linewidth]{part-hspine}
-    % \includegraphics[width=0.25\linewidth]{part-treemap}
+    \includegraphics[width=0.33\linewidth]{part-hspine}%
+    \includegraphics[width=0.33\linewidth]{part-treemap}
 
-  \caption{1d partitions showing proportions 0.5, 0.2, 0.2 and 0.1. From left to right: horizontal bars, horizontal spines, and a treemap.}
+  \caption{1d partitions showing the distribution of happiness.  From left to right: bars, spines and tiles.}
   \label{fig:part-1d}
@@
 
+The appearance of bars and splines is identical for uniformly distributed data.
+
 \subsection{2d primitives}
+\label{sub:part-2d}
 
-\begin{itemize}
-  \item {\bf fluct}: width and height proportional to square root of value. Flucts allow comparisons of absolute numbers in two directions.
+We are currently aware of one 2d primitive, the {\bf fluct}: where width and height are proportional to the square root of the proportion.
+Flucts allow comparisons of relative magnitudes numbers in two directions.
+Each rectangle is arranged on a regular grid formed by the levels of the two variables, and thus the fluct is not space filling.
+A fluct displays a joint 2d distribution, unlike the combination of any two 1d partitions which display a conditional distribution.
 
-\end{itemize}
-
-One special 2d case is the equal bin size plot - which all 1d x 1d primitives and the fluct collapse to when proportions are equal, as when conditioned.
-
-Figure~\ref{fig:fluct} shows these two behaviours, looking at the relationship between happiness, health and financial status.
-On the left, a display of the raw proportion shows that most people are in good health and average financial standing, but it is difficult to see how happiness varies because the size of the squares is variable.
-On the right, conditioning on financial status and health makes all squares the same size and makes it easier to see the conditional distribution of happiness - people tend to be happier the healthier and the richer they are
+One special case of the fluct is the equal bin size plot which occurs when the two variables are jointly uniformly distributed, usually as a result of conditioning.
+The equal bin size plot is useful to visualise missing combinations, or when supplemented with a third variable, to visualise the distribution conditional on two other variables.
 
@@
     \includegraphics[width=0.5\linewidth]{part-fluct-cond}
-  \caption{caption}
+  \caption{Distribution of age (in decades) and marital status.  All combinations of the two variables are present.}
   \label{fig:fluct}
@@
 
-\subsection{Plot templates}
+\subsection{Combinations}
+\label{sub:part-nd}
+
+To visualise higher-dimensional pmfs, we partition recursively using the primitives described above.
+That is, to display $f(x, y, z)$ we first create a partition based on $f(z)$ then partition each of those pieces based on $f(y | z)$ and again by $f(x | y, z)$.
+Figure \ref{fig:recursive} illustrates this process.
+
+Figure~\ref{fig:fluct} uses a fluct and a vspine to explore the relationship between happiness, health and financial status.
+On the left, a fluctuation diagram shows that most people are in good health and average financial standing, but it is difficult to see how happiness varies because the size of the squares is variable.
+Conditioning on financial status and health produces the equal bin size plot on the right and makes it easier to see the conditional distribution of happiness - people tend to be happier the healthier and richer they are.
+
+\begin{figure}[htbp]
+  \centering
+    \includegraphics[width=0.5\linewidth]{part-nd}%
+    \includegraphics[width=0.5\linewidth]{part-nd-1}
+  \caption{caption}
+  \label{fig:part-nd}
+\end{figure}
 
@@
 
-  \item {\bf Equal bin size} plot (3d): fluct and vspine, provided the final two variables are uniform.
+  \item {\bf Equal bin size} plot (3d): fluct and vspine, provided the first two variables are uniform.
 
@@
   \item {\bf Treemap} (nd): n spines. \citep{shneiderman:1992}
-  \item {\bf Squarified treemap} (nd): n squares. \citep{bruls:1999}
+  \item {\bf Squarified treemap} (nd): n tiles. \citep{bruls:1999}
 
@@
 
-Trellis graphics are another type of display that uses categorical variables to create small multiples for different subsets of the data.
-Visually, trellis graphics look like equal-bin-size plots with another plot drawn inside each bin.
+Trellis graphics \citep{becker:1996}, as known as lattice or facetted or
+conditioned graphics are another related display. They uses categorical
+variables to generate multiple panels, each containing a plot of the subset of
+the data. Trellised area plots also fall into our framework and can be
+created by conditioning on the trellising variables.
 
@@
 \label{sec:relax}
+
+There are a number of cases in which it is useful to relax the constraints that we defined above.  These are described in sequence below.
 
@@
 A zero weight should have zero area, but giving it positive area can be useful so it's actually possible to see!
-Similarly for very small values which would otherwise occupy less than a pixel, useful to constain to minimal perceptible size.
+Similarly for very small values which would otherwise occupy less than a pixel, useful to constrain to minimal perceptible size.
 
@@
 % Pseduo 1d, where you make a new variable from multiple variables.
+
+\subsection{Missing values}
+\label{sub:missing_values}
 

### 2009-12-01 plots-partition.r

@@ -1,16 +1,33 @@
-vr <- data.frame(trt = letters[1:4], .wt = c(0.5, 0.2, 0.2, 0.1))
+l(productplots)
+fill_happy <- list(
+  aes(fill = happy),
+  opts(legend.position = "none"),
+  scale_fill_manual(
+    values = c("not too happy" = "#FC8D59", "pretty happy" = "#FFFFBF", "very happy" = "#91CF60", "NA" = "grey50"))
+)
 
-prodplot(vr, ~ trt, "hbar")
+prodplot(happy, ~ happy, "hbar", colour = "grey50", na.rm = TRUE) + fill_happy
 ggsave("plots/part-hbar.pdf", width = 4, height = 4)
 
-prodplot(vr, ~ trt, "hspine")
+prodplot(happy, ~ happy, "hspine", na.rm = TRUE) + fill_happy
 ggsave("plots/part-hspine.pdf", width = 4, height = 4)
 
-prodplot(vr, ~ trt, "treemap")
+prodplot(happy, ~ happy, "treemap", na.rm = TRUE) + fill_happy
 ggsave("plots/part-treemap.pdf", width = 4, height = 4)
 
-
 # 2d partition
+
+happy$decade <- cut(happy$age, seq(10, 90, by = 10))
+prodplot(happy, ~  decade + marital, "fluct", na.rm = T) + 
+  aes(fill = marital)
+ggsave("plots/part-fluct.pdf", width = 6, height = 6)
+prodplot(happy, ~ . |  decade + marital, "fluct", na.rm = T) + 
+  aes(fill = marital)
+ggsave("plots/part-fluct-cond.pdf", width = 6, height = 6)
+
+
+# nd partition
+
 prodplot(happy, ~ happy + finrela, c("vspine", "hspine")) + 
   aes(fill = finrela)
 prodplot(happy, ~ happy + health, c("vspine", "hspine")) + 
@@ -19,9 +36,9 @@ prodplot(happy, ~ happy + marital, c("vspine", "hspine")) +
   aes(fill = marital)
 
 prodplot(happy, ~ happy + finrela + health, c("vspine", "fluct"), 
-  colour = NA, na.rm = TRUE) + aes(fill = happy)
-ggsave("plots/part-fluct.pdf", width = 6, height = 6)
-prodplot(happy, ~ happy | finrela + health, c("vspine", "fluct"), 
[…]

### 2009-12-01 product-graphics.tex
Note about conditional densities
Better equal bin size example

@@
     \includegraphics[width=0.5\linewidth]{part-fluct-cond}
-  \caption{Distribution of age (in decades) and marital status.  All combinations of the two variables are present.}
+  \caption{(Left) Distribution of age (in decades) and marital status.
+All combinations of the two variables are present.
+(Right) Distribution of health status and year of survey.
+Health status was not recorded for three years.}
   \label{fig:fluct}
@@
 
+An alternative approach is use density estimates (or conditional density estimates) and plot with continuous area plots.
+As in: Hofmann, H., Theus, M.
+(2005), Interactive graphics for visualizing conditional distributions, Unpublished Manuscript.
+
 \subsection{Nested data}

### 2009-12-01 plots-partition.r

@@ -17,12 +17,22 @@ ggsave("plots/part-treemap.pdf", width = 4, height = 4)
 
 # 2d partition
 
+# Find combinations with most zeros
+comb <- subset(expand.grid(x = 2:9, y = 2:9), x > y)
+comb$x <- names(happy)[comb$x]
+comb$y <- names(happy)[comb$y]
+subset(mdply(comb, function(x, y) {
+  sum(table(happy[[x]], happy[[y]]) == 0)
+}), V1 > 0)
+
+
+
 happy$decade <- cut(happy$age, seq(10, 90, by = 10))
 prodplot(happy, ~  decade + marital, "fluct", na.rm = T) + 
   aes(fill = marital)
 ggsave("plots/part-fluct.pdf", width = 6, height = 6)
-prodplot(happy, ~ . |  decade + marital, "fluct", na.rm = T) + 
-  aes(fill = marital)
+prodplot(happy, ~ . | year + health, "fluct") + 
+  aes(fill = health)
 ggsave("plots/part-fluct-cond.pdf", width = 6, height = 6)
 
 

### 2009-12-01 colour.r
Update polar plots

@@ -1,0 +1,6 @@
+fill_happy <- list(
+  aes(fill = happy),
+  opts(legend.position = "none"),
+  scale_fill_manual(
+    values = c("not too happy" = "#FC8D59", "pretty happy" = "#FFFFBF", "very happy" = "#91CF60", "NA" = "grey50"))
+)

### 2009-12-01 plots-partition.r

@@ -1,10 +1,4 @@
 l(productplots)
-fill_happy <- list(
-  aes(fill = happy),
-  opts(legend.position = "none"),
-  scale_fill_manual(
-    values = c("not too happy" = "#FC8D59", "pretty happy" = "#FFFFBF", "very happy" = "#91CF60", "NA" = "grey50"))
-)
 
 prodplot(happy, ~ happy, "hbar", colour = "grey50", na.rm = TRUE) + fill_happy
 ggsave("plots/part-hbar.pdf", width = 4, height = 4)

### 2009-12-01 plots-polar.r

@@ -1,6 +1,5 @@
 # Polar coordinates
-
-source("../load.r")
+source("colour.r")
 
 shade <- list(
   aes(fill = sex),
@@ -13,22 +12,22 @@ shade <- list(
   scale_y_sqrt()
 )
 
-prodplot(h, ~ happy + sex) + shade
+prodplot(happy, ~ sex + happy, set_offset(c("vspine", "hspine")), na.rm = T) + shade
 ggsave("plots/hs-vs-cartesian.pdf", width = 4, height = 4)
 last_plot() + coord_polar()
 ggsave("plots/hs-vs-polar.pdf", width = 4, height = 4)
 
-prodplot(h, ~ happy + sex, divider = nested()) + shade
+prodplot(happy, ~ sex + happy, set_offset(c("hbar", "hbar")), na.rm = T) + shade
 ggsave("plots/hb-vb-cartesian.pdf", width = 4, height = 4)
 last_plot() + coord_polar()
 ggsave("plots/hb-vb-polar.pdf", width = 4, height = 4)
 
-prodplot(h, ~ happy + sex, divider = list(hspine, hspine)) + shade
+prodplot(happy, ~ sex + happy, divider = set_offset(c("hspine", "hspine")), na.rm = T) + shade
 ggsave("plots/hs-hs-cartesian.pdf", width = 4, height = 4)
 last_plot() + coord_polar()
 ggsave("plots/hs-hs-polar.pdf", width = 4, height = 4)
 
-prodplot(h, ~ happy + sex, divider = stacked()) + shade
+prodplot(happy, ~ sex + happy, divider = set_offset(c("vspine", "hbar")), na.rm = T) + shade
 ggsave("plots/hb-vs-cartesian.pdf", width = 4, height = 4)
 last_plot() + coord_polar()
 ggsave("plots/hb-vs-polar.pdf", width = 4, height = 4)

### 2009-12-01 product-graphics.tex
More improvements

@@
 
+For example, instead of using flucts to explore the joint distribution of age and marital status, we could use vertical and horizontal splines, as shown in Figure~\ref{fig:marital}.  Th
+
+\begin{figure}[htbp]
+  \centering
+    \includegraphics[width=0.5\linewidth]{part-marital-1}%
+    \includegraphics[width=0.5\linewidth]{part-marital-2}
+  \caption{The joint distribution of marital status and age (in decades) with a vspine and hspine.  On the right, conditional distribution of age has been fixed.}
+  \label{fig:marital}
+\end{figure}
+
 Figure~\ref{fig:fluct} uses a fluct and a vspine to explore the relationship between happiness, health and financial status.

### 2009-12-01 plots-partition.r

@@ -32,6 +32,13 @@ ggsave("plots/part-fluct-cond.pdf", width = 6, height = 6)
 
 # nd partition
 
+prodplot(happy, ~ marital + decade, set_offset(c("vspine", "hspine"), c(0, 0.02)), na.rm = T) + aes(fill = marital)
+ggsave("plots/part-marital-1.pdf", width = 6, height = 6)
+
+prodplot(happy, ~ marital | decade, set_offset(c("vspine", "hspine"), c(0, 0.02)), na.rm = T) + aes(fill = marital)
+ggsave("plots/part-marital-2.pdf", width = 6, height = 6)
+
+
 prodplot(happy, ~ happy + finrela, c("vspine", "hspine")) + 
   aes(fill = finrela)
 prodplot(happy, ~ happy + health, c("vspine", "hspine")) + 

### 2009-12-01 product-graphics.tex
Working on violations of constraints

@@
 It can be useful to violate the constraint that area should be proportional to weight to distinguish between zeros and very small values.
-A zero weight should have zero area, but giving it positive area can be useful so it's actually possible to see!
-Similarly for very small values which would otherwise occupy less than a pixel, useful to constrain to minimal perceptible size.
+A zero weight should have zero area, but giving it positive area can be useful, so you can actually see it!
+In general, it's useful to constrain all areas to be above a certain minimal perceptible size.
+Areas which are constrained in such a way should have a visual flag (such as a difference colour) so ensure that the reader knows that the relationship between area and value has been violated.
 
-Antony's zooming, where you also truncate the maximum size.
+On the other end of the spectrum, it can be useful to constrain put a cap on the largest possible values \citetext{Antony Unwin, priv.
+comm.}.
+By controlling this cap it is possible to zoom in on the smallest values.
 
-Image.
+Other non-linear transformations may also be useful.
+For example, we could display the square root of the weights, to stabilise the variance of the areas.
+If values span multiple orders of magnitude it may be useful to map log value to area.
 
-\subsection{Levels not contained}
+\subsection{Non-disjoint partitions}
 
@@
 This makes it easier to see all the levels of the hierarchy, not just the lowest level.
+Figure~\ref{fig:cascade} shows an example of how cascading can help illuminate the structure of a complex mosaic plot.
 
-Example.
+\begin{figure}[htbp]
+  \centering
+    \includegraphics[width=0.5\linewidth]{sex-marital-happy}%
+    \includegraphics[width=0.5\linewidth]{cascade}
+  \caption{A mosaic plot of happiness by martial status and sex. (Left) Coloured by happiness.  (Right) A cascaded view helps show how the plot is built up.}
+  \label{fig:cascade}
+\end{figure}
 
@@
 
-Pie charts fall out naturally, as bars in polar coordinates, with angle and radius instead of height and width.
-And to ensure that counts stay proportional to areas, the square-root is taken of the y-axis.
-This is related to the infoslices of \citet{andrews:1998}, which only use half of the disk.
-Figure~\ref{fig:polar} illustrates a few product plots in polar coordinates.
-Variables are happiness (top level in four categories) and sex (nested, two levels).
-
-These displays generalise the fourfold displays of \citet{friendly:1995}.
+A pie chart is a popular method of displaying proportions, but is not a rectangular partition.
+However, it turns out that there is a simple relationship between the bar charts and pie charts: an pie chart is an hspine drawn in polar coordinates with the x coordinate mapped to angle and the y coordinate to radius.
+Many other displays that use circulate arcs turn out to be special cases of product plots, drawn in polar coordinates:
 
 \begin{itemize}
-  \item Pie charts
-  \item Nightingale
-  \item Bullseye chart 
+  \item Fourfold displays \citep{friendly:1995}, wind rose, sector graphic: hbar.
+
+  \item Bullseye chart, hspine.
+
+  \item Doughnut plot, vspine + hspine.
+
+  \item Circular bar chart or racetrack plot, vbar.
+
+  \item Infoslices \citep{andrews:1998}, nested vbar.  But only use half of the disk, and are specialised for highly nested data.
+
+  \item Nightingales's coxcomb, is very similar to vspine / hbar in polar coordinates, except that the pie slices are overlapping and so violates the constraint of disjoint area.
 \end{itemize}
+
+\noindent Eight variants are displayed Figure~\ref{fig:polar}.
+
+To make this work more generally, the y-variable (mapped to radius) should be square-root transformed to ensure that that counts stay proportional to areas.
+The radial displays of \citet{stasko:2000} and fanlens of \citet{lou:2007} deliberately choose not to do this to highlight the outer levels.
 
@@
     \includegraphics[width=0.25\linewidth]{hs-vs-polar}
+
+    \includegraphics[width=0.25\linewidth]{hb-vb-polar-2}%
+    \includegraphics[width=0.25\linewidth]{hb-vs-polar-2}%
+    \includegraphics[width=0.25\linewidth]{hs-hs-polar-2}%
+    \includegraphics[width=0.25\linewidth]{hs-vs-polar-2}
   \caption{(Top) Area graphics in Cartesian coordinates.
-(Bottom) Area graphics in polar coordinates.
-From left to right: horizontal bar nested inside horizontal bar, vertical spine nested inside horizontal bar (stacked bar chart), horizontal spine nested in horizontal spine, horizontal spine nested in vertical spine (mosaic plot).}
+(Mid) Area graphics in polar coordinates.
+(Bottom) Product graphics rotated 90 degrees before being converted to polar coordinates.
+From left to right: hbar + hbar, , vspline + hbar, hspine + hspine, hspine + vspine.}
   \label{fig:polar}
@@
 
-Circular treemaps of \citet{wetzel:2008} which use circles instead of rectangles, but is not space filling because you can not tile a circular region with circles.
-The radial displays of \citet{stasko:2000} deliberately keep the radius constant in order to relatively highlight the lower levels.
-Similarly so does the fanlens of \citet{lou:2007}
-
-Other non-rectangular treemaps have been proposed, such as the space-filling curve approach of \citep{wattenberg:2005}, or the voronoi treemaps of \citep{wattenberg:2005}, but because of the perceptual difficultly of comparing areas of arbitrary polygons, these approaches tend to be attractive rather than useful.
-
-% \section{Connection with log-linear models}
+A number of non-rectangular treemaps have been proposed, such as the circular treemaps of \citet{wetzel:2008}, the space-filling curves of \citet{wattenberg:2005} and the voronoi treemaps of \citet{balzer:2005}.
+However, because it is such much more difficult to compare the areas of arbitrary polygons, these approaches tend to be attractive rather than useful.
 

### 2009-12-01 plot-constraints.r

@@ -1,0 +1,16 @@
+l(productplots)
+source("colour.r")
+
+prodplot(happy, ~ happy + marital | sex, na.rm = T, 
+  set_offset(c("vspine", "hspine", "hspine"), c(0, 0.01, 0.02))) + fill_happy
+ggsave("plots/sex-marital-happy.pdf", width = 6, height = 4)
+
+df <- prodcalc(happy, ~ happy + marital | sex, na.rm = T, set_offset(c("vspine", "hspine", "hspine"), c(0, 0.01, 0.02)), cascade = 0.1)
+
+ggplot(df, aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, order = level)) + 
+  geom_rect(aes(colour = level, size = -level), fill = alpha("white", 0.9)) +
+  xlab(NULL) + ylab(NULL) + 
+  scale_colour_gradient("Level", low = "black", high = "grey60", breaks = 1:3) + 
+  scale_size(to = c(1, 2)) +
+  opts(legend.position = "none")
+ggsave("plots/cascade.pdf", width = 6, height = 4)

### 2009-12-01 plots-polar.r

@@ -16,18 +16,30 @@ prodplot(happy, ~ sex + happy, set_offset(c("vspine", "hspine")), na.rm = T) + s
 ggsave("plots/hs-vs-cartesian.pdf", width = 4, height = 4)
 last_plot() + coord_polar()
 ggsave("plots/hs-vs-polar.pdf", width = 4, height = 4)
+last_plot() + coord_polar("y")
+ggsave("plots/hs-vs-polar-2.pdf", width = 4, height = 4)
+
+prodplot(happy, ~ sex + happy, set_offset(c("hspine", "vspine")), na.rm = T) + shade
+last_plot() + coord_polar()
+
 
 prodplot(happy, ~ sex + happy, set_offset(c("hbar", "hbar")), na.rm = T) + shade
 ggsave("plots/hb-vb-cartesian.pdf", width = 4, height = 4)
 last_plot() + coord_polar()
 ggsave("plots/hb-vb-polar.pdf", width = 4, height = 4)
+last_plot() + coord_polar("y")
+ggsave("plots/hb-vb-polar-2.pdf", width = 4, height = 4)
 
 prodplot(happy, ~ sex + happy, divider = set_offset(c("hspine", "hspine")), na.rm = T) + shade
 ggsave("plots/hs-hs-cartesian.pdf", width = 4, height = 4)
 last_plot() + coord_polar()
 ggsave("plots/hs-hs-polar.pdf", width = 4, height = 4)
+last_plot() + coord_polar("y")
+ggsave("plots/hs-hs-polar-2.pdf", width = 4, height = 4)
 
 prodplot(happy, ~ sex + happy, divider = set_offset(c("vspine", "hbar")), na.rm = T) + shade
 ggsave("plots/hb-vs-cartesian.pdf", width = 4, height = 4)
 last_plot() + coord_polar()
 ggsave("plots/hb-vs-polar.pdf", width = 4, height = 4)
+last_plot() + coord_polar("y")
+ggsave("plots/hb-vs-polar-2.pdf", width = 4, height = 4)

### 2009-12-04 product-graphics.tex
nd partition section

@@
 \begin{itemize}
+
   \item {\bf bar}s: height is proportional to value, width equally divides space.
@@
 
-To visualise higher-dimensional pmfs, we partition recursively using the primitives described above.
+To visualise higher-dimensional pmfs, we recursively partition using the previously described primitives.
 That is, to display $f(x, y, z)$ we first create a partition based on $f(z)$ then partition each of those pieces based on $f(y | z)$ and again by $f(x | y, z)$.
-Figure \ref{fig:recursive} illustrates this process.
+Figure \ref{fig:recursive} illustrates this process with a graphic that explores whether, conditional on marital status, there is a difference in happiness between men and women.
 
-For example, instead of using flucts to explore the joint distribution of age and marital status, we could use vertical and horizontal splines, as shown in Figure~\ref{fig:marital}.  Th
+% Heike: may be this would be a good spot for some of your Kronecker
+% product theory?
 
@@
   \centering
-    \includegraphics[width=0.5\linewidth]{part-marital-1}%
-    \includegraphics[width=0.5\linewidth]{part-marital-2}
-  \caption{The joint distribution of marital status and age (in decades) with a vspine and hspine.  On the right, conditional distribution of age has been fixed.}
-  \label{fig:marital}
-\end{figure}
-
-Figure~\ref{fig:fluct} uses a fluct and a vspine to explore the relationship between happiness, health and financial status.
-On the left, a fluctuation diagram shows that most people are in good health and average financial standing, but it is difficult to see how happiness varies because the size of the squares is variable.
-Conditioning on financial status and health produces the equal bin size plot on the right and makes it easier to see the conditional distribution of happiness - people tend to be happier the healthier and richer they are.
-
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=0.5\linewidth]{part-nd}%
-    \includegraphics[width=0.5\linewidth]{part-nd-1}
-  \caption{caption}
-  \label{fig:part-nd}
+    \includegraphics[width=0.33\linewidth]{part-comb-1}%
+    \includegraphics[width=0.33\linewidth]{part-comb-2}%
+    \includegraphics[width=0.33\linewidth]{part-comb-3}
+  \caption{Conditional on marital status, are men or women happier?
+This figure shows the construction of $\sim happy + sex + marital$ from (left to right) vspines by marital status, vspines by marital status and sex, and hspines by marital status, sex and happiness.
+For all levels of marital status, men are slightly less happy.}
+  \label{fig:recursive}
 \end{figure}
@@
 \end{itemize}
+
+Different combinations of partitions reveal different features of the data.
+Take for example, the distribution of age and marital status, as shown in Figure~\ref{fig:fluct}.
+Instead of visualisation the joint distribution with a fluct, we could focus on the conditional distribution of marital status, given age, or vice versa.
+Figure~\ref{fig:marital} shows two ways to do this.
+The left plot shows $\sim marital + age$ with a vspine nested in a hspine, and the right plot $\sim age + martial$ with a hbar nested in a vspine.
+These displays show the same data but are arranged in such a way to make different comparisons easier.
+
+\begin{figure}[htbp]
+  \centering
+    \includegraphics[width=0.5\linewidth]{part-marital-1}%
+    \includegraphics[width=0.5\linewidth]{part-marital-2}
+  \caption{(Left) The joint distribution of marital status and age (in decades) partitioned by a vspine and hspine.
+(Right) The joint distribution of age and marital status partitioned by a vspine and hbar.}
+  \label{fig:marital}
+\end{figure}
+
+Conditioning is also an important tool, because it allows one to remove relationships that are well-known or uninteresting.
+Figure~\ref{fig:part-cond} uses a fluct and a vspine to explore the relationship between happiness, health and financial status.
+The left plot displays raw proportions, showing that most people are in good health and average financial standing.
+However, it is difficult to see how happiness varies within these conditions because the perceptual task is comparing areas.
+Conditioning on financial status and health produces the plot on the right (known as an equal bin size plot) and makes it easier to see the conditional distribution of happiness, because the perceptual task is simpler: comparing length on a common non-aligned scales.
+Here we can see people tend to be happier the healthier and richer they are.
+
+\begin{figure}[htbp]
+  \centering
+    \includegraphics[width=0.5\linewidth]{part-fluctuation}%
+    \includegraphics[width=0.5\linewidth]{part-equal-area}
+  \caption{Two displays of $\sim happy + health + finrela$.
+(Left) Unconditional values partitioned with a vspine and fluct.
+(Right) Same partitions, but health and relative financial status conditioned to uniform distribution.
+We can no longer see the joint distribution of health and financial status, but it is much easier to see the conditional distribution of happiness: healthier and richer people are happier.
+(Maybe money does buy happiness?)}
+  \label{fig:part-cond}
+\end{figure}
 

### 2009-12-04 plots-partition.r

@@ -1,4 +1,7 @@
 l(productplots)
+source("colour.r")
+
+# 1d partitions --------------------------------------------------------------
 
 prodplot(happy, ~ happy, "hbar", colour = "grey50", na.rm = TRUE) + fill_happy
 ggsave("plots/part-hbar.pdf", width = 4, height = 4)
@@ -9,7 +12,7 @@ ggsave("plots/part-hspine.pdf", width = 4, height = 4)
 prodplot(happy, ~ happy, "treemap", na.rm = TRUE) + fill_happy
 ggsave("plots/part-treemap.pdf", width = 4, height = 4)
 
-# 2d partition
+# 2d partitions --------------------------------------------------------------
 
 # Find combinations with most zeros
 comb <- subset(expand.grid(x = 2:9, y = 2:9), x > y)
@@ -20,24 +23,40 @@ subset(mdply(comb, function(x, y) {
 }), V1 > 0)
 
 
-
+# age & marital status
 happy$decade <- cut(happy$age, seq(10, 90, by = 10))
 prodplot(happy, ~  decade + marital, "fluct", na.rm = T) + 
   aes(fill = marital)
 ggsave("plots/part-fluct.pdf", width = 6, height = 6)
-prodplot(happy, ~ . | year + health, "fluct") + 
+
+# year & health
+prodplot(happy, ~ . | year + health, "fluct", na.rm = TRUE) + 
   aes(fill = health)
 ggsave("plots/part-fluct-cond.pdf", width = 6, height = 6)
 
 
-# nd partition
+# nd partitions --------------------------------------------------------------
 
+# show partitioning process
+df <- prodcalc(happy, ~ happy + sex + marital, na.rm = TRUE, 
+  set_offset(c("vspine", "hspine", "hspine"), c(0, 0, 0.01)))
+
+draw(subset(df, level == 1)) + aes(fill = marital)
+ggsave("plots/part-comb-1.pdf", width = 6, height = 6)
+
+draw(subset(df, level == 2)) + aes(fill = sex)
+ggsave("plots/part-comb-2.pdf", width = 6, height = 6)
+
+draw(subset(df, level == 3)) + fill_happy
+ggsave("plots/part-comb-3.pdf", width = 6, height = 6)
+
+
+# alternative displays of marital status & age
[…]

### 2009-12-04 colour.r
Nicer colour schemes

@@ -1,6 +1,36 @@
 fill_happy <- list(
   aes(fill = happy),
   opts(legend.position = "none"),
+  scale_fill_manual(values = c(
+    "not too happy" = "#FC8D59", 
+    "pretty happy" = "#FFFFBF", 
+    "very happy" = "#91CF60", 
+    "NA" = "grey50")))
+
+fill_sex <- list(
+  aes(fill = sex),
+  scale_fill_manual(values = c(
+    "male" = "#67A9CF", 
+    "female" = "#EF8A62", 
+    "NA" = "grey50")))
+
+fill_marital <- list(
+  aes(fill = marital),
+  scale_fill_manual(values = c(
+    "married" = "#F1EEF6", 
+    "never married" = "#BDC9E1", 
+    "divorced" = "#74A9CF", 
+    "widowed" = "#2B8CBE", 
+    "separated" = "#045A8D", 
+    "NA" = "grey50")))
+
+fill_health <- list(
+  aes(fill = health),
   scale_fill_manual(
-    values = c("not too happy" = "#FC8D59", "pretty happy" = "#FFFFBF", "very happy" = "#91CF60", "NA" = "grey50"))
+    values = c(
+      "poor" = "#EDF8E9", 
+      "fair" = "#BAE4B3", 
+      "good" = "#74C476", 
+      "excellent" = "#238B45", 
+      "NA" = "grey50"))
 )

### 2009-12-04 colours.r

@@ -1,9 +1,0 @@
-# Plots should use a consistent colour scheme for happiness so the legend
-# only needs to be displayed once.  Currently there seems to be a bug in 
-# scale_fill_brewer which is messing things up.
-
-happy_fill <- list(
-  aes(fill = happy), 
-  scale_fill_brewer("Happiness", pal = "Blues", 
-    limits = c("not too happy", "happy", "very happy")),
-  opts(legend.position = "none"))

### 2009-12-04 plots-partition.r

@@ -15,23 +15,21 @@ ggsave("plots/part-treemap.pdf", width = 4, height = 4)
 # 2d partitions --------------------------------------------------------------
 
 # Find combinations with most zeros
-comb <- subset(expand.grid(x = 2:9, y = 2:9), x > y)
-comb$x <- names(happy)[comb$x]
-comb$y <- names(happy)[comb$y]
-subset(mdply(comb, function(x, y) {
-  sum(table(happy[[x]], happy[[y]]) == 0)
-}), V1 > 0)
+# comb <- subset(expand.grid(x = 2:9, y = 2:9), x > y)
+# comb$x <- names(happy)[comb$x]
+# comb$y <- names(happy)[comb$y]
+# subset(mdply(comb, function(x, y) {
+#   sum(table(happy[[x]], happy[[y]]) == 0)
+# }), V1 > 0)
 
 
 # age & marital status
 happy$decade <- cut(happy$age, seq(10, 90, by = 10))
-prodplot(happy, ~  decade + marital, "fluct", na.rm = T) + 
-  aes(fill = marital)
+prodplot(happy, ~  decade + marital, "fluct", na.rm = T) + fill_marital
 ggsave("plots/part-fluct.pdf", width = 6, height = 6)
 
 # year & health
-prodplot(happy, ~ . | year + health, "fluct", na.rm = TRUE) + 
-  aes(fill = health)
+prodplot(happy, ~ . | year + health, "fluct", na.rm = TRUE) + fill_health
 ggsave("plots/part-fluct-cond.pdf", width = 6, height = 6)
 
 
@@ -41,10 +39,10 @@ ggsave("plots/part-fluct-cond.pdf", width = 6, height = 6)
 df <- prodcalc(happy, ~ happy + sex + marital, na.rm = TRUE, 
   set_offset(c("vspine", "hspine", "hspine"), c(0, 0, 0.01)))
 
-draw(subset(df, level == 1)) + aes(fill = marital)
+draw(subset(df, level == 1)) + fill_marital
 ggsave("plots/part-comb-1.pdf", width = 6, height = 6)
 
-draw(subset(df, level == 2)) + aes(fill = sex)
+draw(subset(df, level == 2)) + fill_sex
 ggsave("plots/part-comb-2.pdf", width = 6, height = 6)
 
 draw(subset(df, level == 3)) + fill_happy
[…]

### 2009-12-04 plots-polar.r

@@ -1,8 +1,7 @@
 # Polar coordinates
 source("colour.r")
 
-shade <- list(
-  aes(fill = sex),
+shade <- c(fill_sex, list(
   opts(legend.position = "none", 
     axis.text.x = theme_blank(),
     axis.text.y = theme_blank()
@@ -10,7 +9,7 @@ shade <- list(
   xlab(NULL),
   ylab(NULL),
   scale_y_sqrt()
-)
+))
 
 prodplot(happy, ~ sex + happy, set_offset(c("vspine", "hspine")), na.rm = T) + shade
 ggsave("plots/hs-vs-cartesian.pdf", width = 4, height = 4)

### 2009-12-04 product-graphics.tex
Various tweaks

@@
 
+Some notation: let capital letters refer to the sample space, and lower case particular values.
+
 Let $f(x, y, z)$ be the 3d pmf function. We can can write this pmf as product of marginals and conditionals in many different ways.  We can decompose product graphics in a similar way.
@@
   \item \verb!~ a | b! represents $f(a | b)$
+  \item \verb!~ . | a! is a special case of $f(a) = \frac{1}{\#A}$. 
 \end{itemize}
@@
     \includegraphics[width=0.25\linewidth]{hs-vs-polar-2}
+
   \caption{(Top) Area graphics in Cartesian coordinates.
 (Mid) Area graphics in polar coordinates.
-(Bottom) Product graphics rotated 90 degrees before being converted to polar coordinates.
-From left to right: hbar + hbar, , vspline + hbar, hspine + hspine, hspine + vspine.}
+From left to right: hbar + hbar, vspline + hbar, hspine + hspine, hspine + vspine.
+(Bottom) Product graphics rotated 90 degrees before being converted to polar coordinates: vbar + vbar, hspline + vbar, vspine + vspine, vspine + hspine.}
+
   \label{fig:polar}
@@
 Some of the applications of such weighted plots are described in \citet{unwin:2007}.
-Include a brief summary here.
+In the happy dataset, the {\sf wtssall} variable gives analytic survey weights.
+Figure~\ref{fig:weighting} shows the difference between the weighted and unweighted distributions of age and sex.
+The distribution is subtly different.
+
+To express that we are displaying weights, rather than counts, we extend the formula syntax by putting the weighting variable on the left-hand side: $wtssall \sim sex + age$ vs. $ \sim sex + age$.
+
+\begin{figure}[htbp]
+  \centering
+    \includegraphics[width=0.5\linewidth]{wt-count}%
+    \includegraphics[width=0.5\linewidth]{wt-wtssall}%
+  \caption{(Left) Unweighted distribution (counts).  (Right) Weighted by probability weights.}
+  \label{fig:weighting}
+\end{figure}
 

### 2009-12-04 plots-partition.r

@@ -64,10 +64,10 @@ prodplot(happy, ~ happy + marital, c("vspine", "hspine")) +
   aes(fill = marital)
 
 # fluctuation & equal bin size
-prodplot(happy, ~ happy + finrela + health, c("vspine", "fluct"), 
-  colour = NA, na.rm = TRUE) + fill_happy
+prodplot(happy, ~ happy + finrela + health, na.rm = TRUE,
+  set_offset(c("vspine", "fluct"), c(0, 0.05))) + fill_happy
 ggsave("plots/part-fluctuation.pdf", width = 6, height = 6)
 
-prodplot(happy, ~ happy | finrela + health, c("vspine", "fluct"), 
-  colour = NA, na.rm = TRUE) + fill_happy
+prodplot(happy, ~ happy | finrela + health, na.rm = TRUE,
+  set_offset(c("vspine", "fluct"), c(0, 0.05))) + fill_happy
 ggsave("plots/part-equal-area.pdf", width = 6, height = 6)

### 2009-12-04 plots-variations.r

@@ -1,0 +1,15 @@
+library(productplots)
+source("colour.r")
+
+happy2 <- transform(happy, .wt = wtssall)
+
+# prodplot(happy, ~ sex, "hbar") + fill_sex
+# prodplot(happy2, ~ sex, "hbar") + fill_sex
+# prodplot(happy, ~ finrela + health, "fluct")
+# prodplot(happy2, ~ finrela + health, "fluct")
+
+prodplot(happy, ~ sex  + age, set_offset(c("vspine", "hbar")), na.rm = TRUE) + fill_sex
+ggsave("plots/wt-count.pdf", width = 6, height = 4)
+
+prodplot(happy2, ~ sex + age, set_offset(c("vspine", "hbar")), na.rm = TRUE) + fill_sex
+ggsave("plots/wt-wtssall.pdf", width = 6, height = 4)

### 2009-12-05 colour.r
Tweaking plot appearance, now that we have basic axis labels

@@ -9,6 +9,7 @@ fill_happy <- list(
 
 fill_sex <- list(
   aes(fill = sex),
+  opts(legend.position = "none"),
   scale_fill_manual(values = c(
     "male" = "#67A9CF", 
     "female" = "#EF8A62", 
@@ -16,6 +17,7 @@ fill_sex <- list(
 
 fill_marital <- list(
   aes(fill = marital),
+  opts(legend.position = "none"),
   scale_fill_manual(values = c(
     "married" = "#F1EEF6", 
     "never married" = "#BDC9E1", 
@@ -26,6 +28,7 @@ fill_marital <- list(
 
 fill_health <- list(
   aes(fill = health),
+  opts(legend.position = "none"),
   scale_fill_manual(
     values = c(
       "poor" = "#EDF8E9", 

### 2009-12-05 plots-factorising.r

@@ -1,16 +1,17 @@
 library(productplots)
+source("colour.r")
 
-prodplot(happy, ~ happy, "hspine") + aes(fill = happy)
+prodplot(happy, ~ happy, "hspine", na.rm = TRUE) + fill_happy
 ggsave("plots/fact-happy.pdf", width = 4, height = 4)
-prodplot(happy, ~ sex + happy, c("vspine", "hspine")) + aes(fill = sex)
+prodplot(happy, ~ sex + happy, c("vspine", "hspine"), na.rm = TRUE) + fill_sex
 ggsave("plots/fact-happy-sex.pdf", width = 4, height = 4)
-prodplot(happy, ~ sex | happy, c("vspine", "hspine")) + aes(fill = sex)
+prodplot(happy, ~ sex | happy, c("vspine", "hspine"), na.rm = TRUE) + fill_sex
 ggsave("plots/fact-happy|sex.pdf", width = 4, height = 4)
 
 
-prodplot(happy, ~ happy, "hbar") + aes(fill = happy)
+prodplot(happy, ~ happy, "hbar", na.rm = TRUE) + fill_happy
 ggsave("plots/fact-happy-2.pdf", width = 4, height = 4)
-prodplot(happy, ~ sex + happy, c("vspine", "hbar")) + aes(fill = sex)
+prodplot(happy, ~ sex + happy, c("vspine", "hbar"), na.rm = TRUE) + fill_sex
 ggsave("plots/fact-happy-sex-2.pdf", width = 4, height = 4)
-prodplot(happy, ~ sex | happy, c("vspine", "hbar")) + aes(fill = sex)
+prodplot(happy, ~ sex | happy, c("vspine", "hbar"), na.rm = TRUE) + fill_sex
 ggsave("plots/fact-happy|sex-2.pdf", width = 4, height = 4)

### 2009-12-05 plots-partition.r

@@ -39,13 +39,13 @@ ggsave("plots/part-fluct-cond.pdf", width = 6, height = 6)
 df <- prodcalc(happy, ~ happy + sex + marital, na.rm = TRUE, 
   set_offset(c("vspine", "hspine", "hspine"), c(0, 0, 0.01)))
 
-draw(subset(df, level == 1)) + fill_marital
+draw(subset(df, level <= 1)) + fill_marital
 ggsave("plots/part-comb-1.pdf", width = 6, height = 6)
 
-draw(subset(df, level == 2)) + fill_sex
+draw(subset(df, level <= 2)) + fill_sex
 ggsave("plots/part-comb-2.pdf", width = 6, height = 6)
 
-draw(subset(df, level == 3)) + fill_happy
+draw(subset(df, level <= 3)) + fill_happy
 ggsave("plots/part-comb-3.pdf", width = 6, height = 6)
 
 

### 2009-12-05 plots-variations.r

@@ -1,6 +1,8 @@
 library(productplots)
 source("colour.r")
 
+# Weighting ------------------------------------------------------------------
+
 happy2 <- transform(happy, .wt = wtssall)
 
 # prodplot(happy, ~ sex, "hbar") + fill_sex
@@ -8,8 +10,22 @@ happy2 <- transform(happy, .wt = wtssall)
 # prodplot(happy, ~ finrela + health, "fluct")
 # prodplot(happy2, ~ finrela + health, "fluct")
 
-prodplot(happy, ~ sex  + age, set_offset(c("vspine", "hbar")), na.rm = TRUE) + fill_sex
+prodplot(happy, ~ sex  + age, set_offset(c("vspine", "hbar")), na.rm = TRUE) +
+  fill_sex
 ggsave("plots/wt-count.pdf", width = 6, height = 4)
 
-prodplot(happy2, ~ sex + age, set_offset(c("vspine", "hbar")), na.rm = TRUE) + fill_sex
+prodplot(happy2, ~ sex + age, set_offset(c("vspine", "hbar")), na.rm = TRUE) +
+ fill_sex
 ggsave("plots/wt-wtssall.pdf", width = 6, height = 4)
+
+
+# Continuous data ------------------------------------------------------------
+
+happy$age_n <- cut_number(happy$age, 10)
+happy$age_i <- cut_interval(happy$age, 10)
+
+prodplot(happy, ~ age_n, set_offset("hbar"))
+prodplot(happy, ~ age_i, set_offset("hbar"))
+
+prodplot(happy, ~ age_n, set_offset("hspine"))
+prodplot(happy, ~ age_i, set_offset("hspine"))

### 2009-12-05 product-graphics.tex
Initial work on variations

@@
 These techniques can also be used with continuous data, if the data is binned to create a categorical variable.
+There are many different possible choices of binning algorithms, but we will focus on two:
+
+\begin{itemize}
+  \item Intervals of equal width.
+  \item Intervals containing equal numbers of points.
+\end{itemize}
+
 This gives rise to the histogram and spinogram, analogues of the bar chart and spine chart respectively.
@@
 
-\subsection{Nested data}
-\label{sub:nested_data}
-
-We distinguish nested data from crossed data by the number of interactions with missing values.  Completely crossed data contains every combination of all variables.  Nested data contains
-
-Aka sparse vs. dense.
-
-It is not always possible to determine nested data from inspection of the data alone, but may need knowledge of the data collection.
+% \subsection{Nested data}
+% \label{sub:nested_data}
+% 
+% Use space-filling!
+% 
+% We distinguish nested data from crossed data by the number of interactions with missing values.  Completely crossed data contains every combination of all variables.  Nested data contains 
+% 
+% 
+% Aka sparse vs. dense.
+% 
+% It is not always possible to determine nested data from inspection of the data alone, but may need knowledge of the data collection.
 For example, if a survey of teachers within schools was performed, and each teacher was identified by a identifier unique within a school, it might look like the data was only missing a few combinations, when in fact it is meaningless.
-
-Or when you have a hierarchy.
-
-Example dataset - Doug Bates teachers/schools?
-
-Possibilities: create new variable which is interaction of nested vars, or drop missings/zeros.  Treemaps deal almost exclusively with nested data.
-
-\section{Data analysis} % (fold)
-\label{sec:data_analysis}
-
-% section data_analysis (end)
+% 
+% 
+% 
+% Or when you have a hierarchy.
+% 
+% Example dataset - Doug Bates teachers/schools?
+% 
+% Possibilities: create new variable which is interaction of nested vars, or drop missings/zeros.  Treemaps deal almost exclusively with nested data.
 

### 2009-12-05 product-graphics.tex
A bit more on labelling

@@
 
-Labelling these plots is particularly challenging.  Future work!
+Due to the hierarchical nature of these plots, it's difficult to label them in an informative way.
+We take a pragmatic approach, and label a single horizontal and single vertical split, where possible.
+Understanding the composition of these plots is most easy when:
 
 \begin{itemize}
-  \item Additional info 
+
+  \item Exploration: You have interactive control over the creation of the plot, and can interactively query regions of interest
+
+  \item Communication: The plot is built up step-by-step, and shows the minimum number of splits.  It may be easier to understand a single complex plot when factored into two or more simpler plots.
+
+\end{itemize}
+
+To create the labels in used in this paper, we inspect each level of the factorisation looking for the first level in which there are rows or columns.
+We do this rather inspecting the partitions that make up the plot, because there are many possible combinations that create columns: hsplines, hbars, vsplines, vbars for uniform data, flucts, ...
+and then can occur at any level in the hierarchy depending on what's come before.
+
+Apart from textual labels on the x-axis, other work has labelled individual cells with attributes of the data:
+
 \begin{itemize}
@@
 
-  \item Display of hierarchy
+Or attributes of the hierarchy:
+
 \begin{itemize}
@@
   \item Labelling
-  \end{itemize}
 \end{itemize}

### 2009-12-05 product-graphics.tex
Add colour keys to captions where needed

@@
 \usepackage[pdftex]{graphicx}
+\usepackage{tikz}
 \DeclareGraphicsExtensions{.png,.pdf}
@@
 \bibliographystyle{abbrvnat}
-\usepackage[small]{caption2}
+\usepackage[small]{caption}
 \usepackage[small]{titlesec}
 \renewcommand\familydefault{bch}
+
+\definecolor{female}{HTML}{EF8A62}
+\definecolor{male}{HTML}{67A9CF}
+\definecolor{not-too-happy}{HTML}{FC8D59}
+\definecolor{pretty-happy}{HTML}{FFFFBF}
+\definecolor{very-happy}{HTML}{91CF60}
+\definecolor{grey30}{HTML}{4D4D4D}
+\newcommand{\key}[1]
+  {\protect \tikz{\fill[#1] rectangle (1ex,1ex);}}
 
@@
 
-  \caption{Plots of the distribution of happiness and sex.}
+  \caption{Plots of the distribution of happiness and sex (\key{male} male, \key{female} female).}
   \label{fig:fact-simple}
@@
 We can no longer see the joint distribution of health and financial status, but it is much easier to see the conditional distribution of happiness: healthier and richer people are happier.
-(Maybe money does buy happiness?)}
+(Maybe money does buy happiness?).
+(\key{not-too-happy} Not too happy, \key{pretty-happy} pretty happy, \key{very-happy} very happy)}
   \label{fig:part-cond}
@@
     \includegraphics[width=0.5\linewidth]{cascade}
-  \caption{A mosaic plot of happiness by martial status and sex. (Left) Coloured by happiness.  (Right) A cascaded view helps show how the plot is built up.}
+  \caption{A mosaic plot of happiness by martial status and sex.
+(Left) Coloured by happiness (\key{not-too-happy} Not too happy, \key{pretty-happy} pretty happy, \key{very-happy} very happy).
+(Right) A cascaded view helps show how the plot is built up.}
   \label{fig:cascade}
@@
 From left to right: hbar + hbar, vspline + hbar, hspine + hspine, hspine + vspine.
-(Bottom) Product graphics rotated 90 degrees before being converted to polar coordinates: vbar + vbar, hspline + vbar, vspine + vspine, vspine + hspine.}
+(Bottom) Product graphics rotated 90 degrees before being converted to polar coordinates: vbar + vbar, hspline + vbar, vspine + vspine, vspine + hspine.
+(\key{male} male, \key{female} female)}
 

### 2009-12-06 product-graphics.tex
Minor tweaking

@@
 A pie chart is a popular method of displaying proportions, but is not a rectangular partition.
-However, it turns out that there is a simple relationship between the bar charts and pie charts: an pie chart is an hspine drawn in polar coordinates with the x coordinate mapped to angle and the y coordinate to radius.
-Many other displays that use circulate arcs turn out to be special cases of product plots, drawn in polar coordinates:
+However, it turns out that there is a simple relationship between product plots and pie charts: an pie chart is an hspine drawn in polar coordinates with the x coordinate mapped to angle and the y coordinate to radius.
+Many other displays that use circular arcs turn out to be special cases of product plots, drawn in polar coordinates:
 
@@
 
-  \item Bullseye chart, hspine.
+  \item Bullseye chart: hspine.
 
-  \item Doughnut plot, vspine + hspine.
+  \item Doughnut plot: vspine + hspine.
 
-  \item Circular bar chart or racetrack plot, vbar.
+  \item Circular bar chart or racetrack plot: vbar.
 
-  \item Infoslices \citep{andrews:1998}, nested vbar.  But only use half of the disk, and are specialised for highly nested data.
+  \item Infoslices \citep{andrews:1998}, nested vbar. But they only use half of the disk, and are specialised for highly nested data.
 
@@
 
-\noindent Eight variants are displayed Figure~\ref{fig:polar}.
+\noindent Eight variants are displayed in Figure~\ref{fig:polar}.
 
@@
 A number of non-rectangular treemaps have been proposed, such as the circular treemaps of \citet{wetzel:2008}, the space-filling curves of \citet{wattenberg:2005} and the voronoi treemaps of \citet{balzer:2005}.
-However, because it is such much more difficult to compare the areas of arbitrary polygons, these approaches tend to be attractive rather than useful.
+However, because it is so much more difficult to compare the areas of arbitrary polygons, these approaches tend to be attractive rather than useful.
 
@@
 
-% Pseudo 2d, where you take a long 1d structure and wrap it into 2d.
-% 
-% Pseduo 1d, where you make a new variable from multiple variables.
+On top of this basic framework, there are a few variations designed for dealing with special types of data: missing data, weighted data, and continuous data.  These variations are described below.
 
@@
 
-We have assumed the probabilities represent counts, but without loss of generality we can use any set of non-negative, additive weights instead.
-Some of the applications of such weighted plots are described in \citet{unwin:2007}.
-In the happy dataset, the {\sf wtssall} variable gives analytic survey weights.
+We have assumed that the probabilities represent counts, but without loss of generality, we can use any set of non-negative, additive weights.
+For example, in the happy dataset, the {\sf wtssall} variable gives analytic survey weights.
 Figure~\ref{fig:weighting} shows the difference between the weighted and unweighted distributions of age and sex.
 The distribution is subtly different.
+\citep{unwin:2007} describes the applications of weighted plots in more detail.
 
@@
     \includegraphics[width=0.5\linewidth]{wt-wtssall}%
-  \caption{(Left) Unweighted distribution (counts).  (Right) Weighted by probability weights.}
+  \caption{Joint distribution of age and sex (\key{male} male, \key{female} female). (Left) counts and (right) probability weights.}
   \label{fig:weighting}
@@
 Due to the hierarchical nature of these plots, it's difficult to label them in an informative way.
-We take a pragmatic approach, and label a single horizontal and single vertical split, where possible.
+We take a pragmatic approach, and label a single horizontal and single vertical split, where possible, using colour to identify a third variable, if necessary.
+
 Understanding the composition of these plots is most easy when:
@@
   \item colour (map of the market)
-  \item texture (sieve plots)
-  \item photographs
+  \item texture, such as in sieve plots \citep{friendly:2000}
+  \item photographs \citep{bederson:2001}
   \item text (tables)

### 2010-04-25 labels.tex
Thrashing paper into shape

--- 2009-12-05 product-graphics.tex
@@
-\documentclass[letterpaper,oneside]{scrartcl}
-\usepackage{fullpage}
-\usepackage[utf8]{inputenc}
-\usepackage[pdftex]{graphicx}
-\DeclareGraphicsExtensions{.png,.pdf}
-\graphicspath{{plots/}}
-\usepackage{hyperref}
-\usepackage{url}
-\usepackage[round,sectionbib]{natbib}
-\bibliographystyle{abbrvnat}
-\usepackage[small]{caption2}
-\usepackage[small]{titlesec}
-\renewcommand\familydefault{bch}
-
-\title{Product graphics}
-\author{Hadley Wickham and Heike Hofmann}
-\begin{document}
-\maketitle
-
-\section{Introduction}
-
-% JASA methods
-% Infovis?
-
-% Aims of this paper:
-%
-%  * emphasise the connection between area graphics and factorising 
-%    probabilities
-% 
-%  * pull all area plots on a common footing - provide a grammar for 
-%    categorical graphics
-
-Area plots are the graphical equivalent of contingency tables, displaying how a total is broken down into components.
-
-display categorical data with a polygon, typically a rectangle, for each combination of the factors of interest, with the area proportional to the number of observations in that combination.
-Because the total area for a graphic is usually constrained, an area plot serves to partition up this total, displaying proportions rather than counts.
-
-Figure \ref{fig:cat-examples} illustrates some of these plots.
-
-\begin{figure}[htbp]
-  \begin{center}
-  %  \includegraphics[scale=1]{file}
-  \end{center}
-  \caption{Area plot examples.}
-  \label{fig:cat-examples}
-\end{figure}
-
-Hierarchical partitioning of space.
-
-\section{Factorising}
-
-Basic idea is factorising table of probabilities:
-
-\begin{itemize}
-  \item as products of marginal and conditional distributions
-  \item as areas in plots
-  \item as sums of parameters in log-linear models
-\end{itemize}
-
-Some notation: let capital letters refer to the sample space, and lower case particular values.
-
-Let $f(x, y, z)$ be the 3d pmf function. We can can write this pmf as product of marginals and conditionals in many different ways.  We can decompose product graphics in a similar way.
-
-\begin{itemize}
-  \item $f(x, y, z) = f(x, y | z) f(z)$
-  \item $f(x, y, z) = f(x | y, z) f(y, z)$
-  \item $f(x, y, z) = f(x | y, z) f(y | z) f(z)$
-\end{itemize}
-
-It can be useful to display partial decompositions, i.e. just $f(x, y | z)$, which is still a 3 variable pmf, and is defined as
-
-\[ f(x, y | z) = \frac{f(x, y, z)}{\sum_{x = a, y = b} f(a, b, z)} \]
-
-so that each category of z sums to one, and for any fixed value of z, $f(x, y | z)$ is a valid pmf.
-
-\[ \forall b \sum_a f(a | b) = 1 \]
-
-Because of their hierarchical nature, an area plot of $f(x, y, z)$ also shows $f(y, z)$ and $f(z)$.
-
-Based on the modelling specification of \citet{wilkinson:1973} we define an algebra for describing these tables.
-
-\begin{itemize}
-  \item \verb|~ a| represents $f(a)$,
-  \item \verb|~ a + b| represents $f(a, b)$
-  \item \verb!~ a | b! represents $f(a | b)$
-  \item \verb!~ . | a! is a special case of $f(a) = \frac{1}{\#A}$. 
-\end{itemize}
-
-By convention we will write the top most level of the product plot on the left, and lower levels to the right.
-While $f(x, y, z)$ and $f(z, y, x)$ display the same counts, their interpretation is different because of which marginal tables you see.
-
-Figure~\ref{fig:fact-simple} illustrates these ideas with the distribution of happiness and sex.
-We start with a plot where area broken down by happiness.
-We can then continue to factor this plot to also display sex - the area of each rectangle is proportional to the number of people who are happy with that sex, and the construction allows us to see the conditional distribution of sex within happy.
-In this case the breaks line up, suggesting that there is no difference in sex, given happiness.
-This could be followed up by a formula hypothesis test.
-
-Removing the marginal effect of happiness, as in the column on the far right, makes each of the top-level categories the same size, and makes it easier to compare the conditional distributions.
-It's not that important in this case, but in the bar chart below, it allows us to see the conditional distribution, and in the case when there are many levels with very different probabilities it can make the plot much easier to read.
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=0.33\linewidth]{fact-happy}%
-  \includegraphics[width=0.33\linewidth]{fact-happy-sex}%
-  \includegraphics[width=0.33\linewidth]{fact-happy|sex}
-
-  \includegraphics[width=0.33\linewidth]{fact-happy-2}%
-  \includegraphics[width=0.33\linewidth]{fact-happy-sex-2}%
-  \includegraphics[width=0.33\linewidth]{fact-happy|sex-2}
-
-  \caption{Plots of the distribution of happiness and sex.}
-  \label{fig:fact-simple}
-\end{figure}
-
-These techniques are also closely related to dimensional stacking \citep{leblanc:1990}
-
-% Other papers to cite:
-% 
-% Generalised treemaps: \citep{vliegen:2006}
-% Cocktailmaps: \citep{ahlberg:1996}
-% Extension to mosaic: \citep{friendly:1999,hofmann:2000}
-% Connection to models: \citep{theus:1999,hofmann:2001}
-
-% Hierarchical coordinate systems 
-%
-% What are the common components of all these plots, and how can be describe
-% them in a succinct way that also allows us to describe what tasks each
-% display is good for?  We propose that there is one key feature that makes
-% these plots special.  That is there their coordinate system, which is
-% hierarchical.  
-%
-% In most coordinate systems that we are used to, such as Cartesian or polar
-% coordinates, are symmetric in the sense that you can locate a point by
-% starting from either dimension.  For example, in Cartesian coordinates you
-% can identify either the x- or y-coordinate first, and the other next.  This
-% also implies a distance symmetry when projected onto 2D Euclidean geometry.
-% When you vary one coordinate by a small amount (holding all others constant)
-% points in the projective geometry will be close.
-%
-% Neither of these properties hold for hierarchical coordinate systems. 
-%
-% We will discuss hierachical coordinate systems in general, and then present
-% a specific hierachical coordinate system which gives rise to the area plots
-% described previously.
-
-\section{Data}
-\label{sec:data}
-
-Two ways of representing data:
-
-\begin{itemize}
-  \item Multi-dimensional contingency table
-  \item Data frame, with column of counts
-\end{itemize}
-
-The two are equivalent, but it's usually computational easier to work with the data frame version, because variables are explicit, and mathematically easier to work with the multidimensional array because multiplication is nicely defined.
-Also deals better with sparsity and is common input to modelling algorithms.
-
-Define the conditioning operator which makes the distribution uniform.
-
-In this paper, we'll use a small sample from the general social survey ({\sc gss}) focussing on happiness. 51\,020 observations from cross-section survey given yearly from 1972 to 2006.
-
-\begin{itemize}
-  \item Happy.  Discrete: ``very happy'', ``pretty happy'', ``not too happy''
-  \item Age (in years), 18--89.
-  \item Sex: ``female'', ``male''
-  \item Degree: ``lt high school'', ``high school'', ``junior college'', ``bachelor'', ``gradudate''
-  \item Relative financial status {\sf finrela}: ``far above average'', ``above average'', ``average'', ``below average'', ``far below average''
-  \item Health: ``excellent'', ``good'', ``fair'', ``poor''
-  \item Year: 1972--2006.
-  \item Probability weight ({\sf wtsall}): 0.43--6.42.
-\end{itemize}
-
-\section{Display}
-\label{sec:display}
-
-Once we have made the decision to map area to proportion, we need to impose some constraints on the set of possible tilings of the plane.
-In general, we could also think of partitions of higher-dimensional spaces (e.g.
-3d or 4d), but we will need to project these down to two for viewing anyway, so there is little disadvantage to working in 2d.
-
-Some constraints are implied by our perceptual capabilities and others by geometry. On top of the constraint that area must be proportional to value, we add two additional constraints:
-
-\begin{itemize}
-
-  \item Partitions must be rectangular.
-It is easier to compare the areas of simple shape (i.e.
-rectangle vs polygon), and even better when one the length of one border is fixed, so we only need compare the length of the other \citep{cleveland:1984}.
-Rectangular partitions are also recursive in the sense that we can always tile a rectangle with smaller rectangles.
-This is not true for most shapes.
-
-  \item Partitions must be disjoint.
-To be able to see the complete area of each rectangle they must be non-overlapping, and each split must be nested within the shapes of the previous level.
-Note that we do not impose the constraint that the tiling be space-filling.
-Space filling methods use will ensure that areas are larger, but make it harder to compare magnitudes because scales no longer align.
-
-\end{itemize}
-
-Note, however, that each of these constraints can be profitably relaxed. See Section~\ref{sec:relax} for details.
-
-Let $a(i)$ give the area assigned to category $i$, which has proportion $f(i)$.
-Since we are using rectangles, $a(i) = k \dot a(i) = k \dot w(i) \times h(i)$ where $w$ and $h$ represent width and height respectively.
-Certain partitions also satisfy the constraint that $h(i) = \sqrt{k} f(i)^x$ and $w(i) = \sqrt{k} f(i)^y$ where $x + y = 1$.
-This is useful because it is only necessary to inspect one border to compare magnitudes.
-
-The following two sections describe the partitions currently used to produce area graphics.
-Section~\ref{sub:part-1d} discusses partitions for 1d pmfs: bars, spines and tiles, Section~\ref{sub:part-2d} discusses one partition for 2d pmfs and Section~\ref{sub:part-nd} describes how the 1d and 2d primitives can be recursively combined to generate a partitioning for any higher dimension pmf and discusses how existing area graphics fall into this framework.
-
-\subsection{1d primitives}
-\label{sub:part-1d}
-
-There are three 1d primitives, as shown in Figure~\ref{fig:part-1d}:
-
-\begin{itemize}
-
-  \item {\bf bar}s: height is proportional to value, width equally divides space.
-Bars allow comparison between absolute numbers, but are not space filling.
-They occupy sum(values / max(values)) of the total area.
-Bars can be arranged horizontally (``hbar'') or vertically (``vbar'').
-Not space filling.
-
-  \item {\bf spine}s: width is proportional to value, height equally divides space.
-Splines are space filling, but make it more difficult to compare absolute counts.
-Splines can be arranged horizontally (``hspine''),  vertically (``vspine''), or can automatically pick their orientation (``spine'').
-Space filling.
-
-  \item {\bf tile}s: tile the plane with rectangles, trying to keep the aspect ratio of each rectangle close to 1.
-Comparing the areas is the most difficult perceptually, and are most useful when there are a large number of categories.
-The is partitioning used by the squarified treemap \citep{bruls:1999}.
-Not space filling.
-
-\end{itemize}
-
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=0.33\linewidth]{part-hbar}%
-    \includegraphics[width=0.33\linewidth]{part-hspine}%
-    \includegraphics[width=0.33\linewidth]{part-treemap}
-
-  \caption{1d partitions showing the distribution of happiness.  From left to right: bars, spines and tiles.}
-  \label{fig:part-1d}
-\end{figure}
-
-The appearance of bars and splines is identical for uniformly distributed data.
-
-\subsection{2d primitives}
-\label{sub:part-2d}
-
-We are currently aware of one 2d primitive, the {\bf fluct}: where width and height are proportional to the square root of the proportion.
-Flucts allow comparisons of relative magnitudes numbers in two directions.
-Each rectangle is arranged on a regular grid formed by the levels of the two variables, and thus the fluct is not space filling.
-A fluct displays a joint 2d distribution, unlike the combination of any two 1d partitions which display a conditional distribution.
-
-One special case of the fluct is the equal bin size plot which occurs when the two variables are jointly uniformly distributed, usually as a result of conditioning.
-The equal bin size plot is useful to visualise missing combinations, or when supplemented with a third variable, to visualise the distribution conditional on two other variables.
-
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=0.5\linewidth]{part-fluct}%
-    \includegraphics[width=0.5\linewidth]{part-fluct-cond}
-  \caption{(Left) Distribution of age (in decades) and marital status.
-All combinations of the two variables are present.
-(Right) Distribution of health status and year of survey.
-Health status was not recorded for three years.}
-  \label{fig:fluct}
-\end{figure}
-
-\subsection{Combinations}
-\label{sub:part-nd}
-
-To visualise higher-dimensional pmfs, we recursively partition using the previously described primitives.
-That is, to display $f(x, y, z)$ we first create a partition based on $f(z)$ then partition each of those pieces based on $f(y | z)$ and again by $f(x | y, z)$.
-Figure \ref{fig:recursive} illustrates this process with a graphic that explores whether, conditional on marital status, there is a difference in happiness between men and women.
-
-% Heike: may be this would be a good spot for some of your Kronecker
-% product theory?
-
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=0.33\linewidth]{part-comb-1}%
-    \includegraphics[width=0.33\linewidth]{part-comb-2}%
-    \includegraphics[width=0.33\linewidth]{part-comb-3}
-  \caption{Conditional on marital status, are men or women happier?
-This figure shows the construction of $\sim happy + sex + marital$ from (left to right) vspines by marital status, vspines by marital status and sex, and hspines by marital status, sex and happiness.
-For all levels of marital status, men are slightly less happy.}
-  \label{fig:recursive}
-\end{figure}
-
-The sequence of partitioning choices defines different named area plots:
-
-\begin{itemize}
-  \item {\bf Barchart} (1d). 1 hbar.
-  \item {\bf Column chart} (1d). 1 vbar.
-  \item {\bf Spineplot} (1d). 1 vspine.
-
-  \item {\bf Stacked} barchart (2d). 1 hbar and 1 vspine.
-  \item {\bf Nested} barchart (2d).  2 hbars. \citep{peltier:2009}
-  \item {\bf Fluctuation} diagram (2d): fluct.
-
-  \item {\bf Equal bin size} plot (3d): fluct and vspine, provided the first two variables are uniform.
-
-  \item {\bf Mosaic} plot (nd).  Alternating hspines and vspines.  \citep{hartigan:1984,hartigan:1981,friendly:1994,hofmann:2003}
-  \item {\bf Double-decker} plot (nd).  $n-1$ hspines and 1 vspine. \citep{hofmann:2001}
-  \item {\bf Treemap} (nd): n spines. \citep{shneiderman:1992}
-  \item {\bf Squarified treemap} (nd): n tiles. \citep{bruls:1999}
-
-\end{itemize}
-
-Different combinations of partitions reveal different features of the data.
-Take for example, the distribution of age and marital status, as shown in Figure~\ref{fig:fluct}.
-Instead of visualisation the joint distribution with a fluct, we could focus on the conditional distribution of marital status, given age, or vice versa.
-Figure~\ref{fig:marital} shows two ways to do this.
-The left plot shows $\sim marital + age$ with a vspine nested in a hspine, and the right plot $\sim age + martial$ with a hbar nested in a vspine.
-These displays show the same data but are arranged in such a way to make different comparisons easier.
-
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=0.5\linewidth]{part-marital-1}%
-    \includegraphics[width=0.5\linewidth]{part-marital-2}
-  \caption{(Left) The joint distribution of marital status and age (in decades) partitioned by a vspine and hspine.
-(Right) The joint distribution of age and marital status partitioned by a vspine and hbar.}
-  \label{fig:marital}
-\end{figure}
-
-Conditioning is also an important tool, because it allows one to remove relationships that are well-known or uninteresting.
-Figure~\ref{fig:part-cond} uses a fluct and a vspine to explore the relationship between happiness, health and financial status.
-The left plot displays raw proportions, showing that most people are in good health and average financial standing.
-However, it is difficult to see how happiness varies within these conditions because the perceptual task is comparing areas.
-Conditioning on financial status and health produces the plot on the right (known as an equal bin size plot) and makes it easier to see the conditional distribution of happiness, because the perceptual task is simpler: comparing length on a common non-aligned scales.
-Here we can see people tend to be happier the healthier and richer they are.
-
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=0.5\linewidth]{part-fluctuation}%
-    \includegraphics[width=0.5\linewidth]{part-equal-area}
-  \caption{Two displays of $\sim happy + health + finrela$.
-(Left) Unconditional values partitioned with a vspine and fluct.
-(Right) Same partitions, but health and relative financial status conditioned to uniform distribution.
-We can no longer see the joint distribution of health and financial status, but it is much easier to see the conditional distribution of happiness: healthier and richer people are happier.
-(Maybe money does buy happiness?)}
-  \label{fig:part-cond}
-\end{figure}
-
-% http://peltiertech.com/Utility/ClusterStackUtility.html
-
-Trellis graphics \citep{becker:1996}, as known as lattice or facetted or
-conditioned graphics are another related display. They uses categorical
-variables to generate multiple panels, each containing a plot of the subset of
-the data. Trellised area plots also fall into our framework and can be
-created by conditioning on the trellising variables.
-
-\section{Relaxing constraints}
-\label{sec:relax}
-
-There are a number of cases in which it is useful to relax the constraints that we defined above.  These are described in sequence below.
-
-\subsection{Area not proportional to weight}
-
-It can be useful to violate the constraint that area should be proportional to weight to distinguish between zeros and very small values.
-A zero weight should have zero area, but giving it positive area can be useful, so you can actually see it!
-In general, it's useful to constrain all areas to be above a certain minimal perceptible size.
-Areas which are constrained in such a way should have a visual flag (such as a difference colour) so ensure that the reader knows that the relationship between area and value has been violated.
-
-On the other end of the spectrum, it can be useful to constrain put a cap on the largest possible values \citetext{Antony Unwin, priv.
-comm.}.
-By controlling this cap it is possible to zoom in on the smallest values.
-
-Other non-linear transformations may also be useful.
-For example, we could display the square root of the weights, to stabilise the variance of the areas.
-If values span multiple orders of magnitude it may be useful to map log value to area.
-
-\subsection{Non-disjoint partitions}
-
-The cascaded treemaps of \citet{lu:2008} is an idea that illustrates how the violation of containment can be productive.
-In the cascaded treemap, each level is slightly offset from the one above to create a pseudo-3d perspective.
-This makes it easier to see all the levels of the hierarchy, not just the lowest level.
-Figure~\ref{fig:cascade} shows an example of how cascading can help illuminate the structure of a complex mosaic plot.
-
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=0.5\linewidth]{sex-marital-happy}%
-    \includegraphics[width=0.5\linewidth]{cascade}
-  \caption{A mosaic plot of happiness by martial status and sex. (Left) Coloured by happiness.  (Right) A cascaded view helps show how the plot is built up.}
-  \label{fig:cascade}
-\end{figure}
-
-\subsection{Non-rectangular partitions}
-
-A pie chart is a popular method of displaying proportions, but is not a rectangular partition.
-However, it turns out that there is a simple relationship between the bar charts and pie charts: an pie chart is an hspine drawn in polar coordinates with the x coordinate mapped to angle and the y coordinate to radius.
-Many other displays that use circulate arcs turn out to be special cases of product plots, drawn in polar coordinates:
-
-\begin{itemize}
-  \item Fourfold displays \citep{friendly:1995}, wind rose, sector graphic: hbar.
-
-  \item Bullseye chart, hspine.
-
-  \item Doughnut plot, vspine + hspine.
-
-  \item Circular bar chart or racetrack plot, vbar.
-
-  \item Infoslices \citep{andrews:1998}, nested vbar.  But only use half of the disk, and are specialised for highly nested data.
-
-  \item Nightingales's coxcomb, is very similar to vspine / hbar in polar coordinates, except that the pie slices are overlapping and so violates the constraint of disjoint area.
-\end{itemize}
-
-\noindent Eight variants are displayed Figure~\ref{fig:polar}.
-
-To make this work more generally, the y-variable (mapped to radius) should be square-root transformed to ensure that that counts stay proportional to areas.
-The radial displays of \citet{stasko:2000} and fanlens of \citet{lou:2007} deliberately choose not to do this to highlight the outer levels.
-
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=0.25\linewidth]{hb-vb-cartesian}%
-    \includegraphics[width=0.25\linewidth]{hb-vs-cartesian}%
-    \includegraphics[width=0.25\linewidth]{hs-hs-cartesian}%
-    \includegraphics[width=0.25\linewidth]{hs-vs-cartesian}
-
-    \includegraphics[width=0.25\linewidth]{hb-vb-polar}%
-    \includegraphics[width=0.25\linewidth]{hb-vs-polar}%
-    \includegraphics[width=0.25\linewidth]{hs-hs-polar}%
-    \includegraphics[width=0.25\linewidth]{hs-vs-polar}
-
-    \includegraphics[width=0.25\linewidth]{hb-vb-polar-2}%
-    \includegraphics[width=0.25\linewidth]{hb-vs-polar-2}%
-    \includegraphics[width=0.25\linewidth]{hs-hs-polar-2}%
-    \includegraphics[width=0.25\linewidth]{hs-vs-polar-2}
-
-  \caption{(Top) Area graphics in Cartesian coordinates.
-(Mid) Area graphics in polar coordinates.
-From left to right: hbar + hbar, vspline + hbar, hspine + hspine, hspine + vspine.
-(Bottom) Product graphics rotated 90 degrees before being converted to polar coordinates: vbar + vbar, hspline + vbar, vspine + vspine, vspine + hspine.}
-
-  \label{fig:polar}
-\end{figure}
-
-A number of non-rectangular treemaps have been proposed, such as the circular treemaps of \citet{wetzel:2008}, the space-filling curves of \citet{wattenberg:2005} and the voronoi treemaps of \citet{balzer:2005}.
-However, because it is such much more difficult to compare the areas of arbitrary polygons, these approaches tend to be attractive rather than useful.
-
-\section{Variations}
-\label{sec:variations}
-
-% Pseudo 2d, where you take a long 1d structure and wrap it into 2d.
-% 
-% Pseduo 1d, where you make a new variable from multiple variables.
-
-\subsection{Missing values}
-\label{sub:missing_values}
-
-\subsection{Weighting}
-\label{sub:weighting}
-
-We have assumed the probabilities represent counts, but without loss of generality we can use any set of non-negative, additive weights instead.
-Some of the applications of such weighted plots are described in \citet{unwin:2007}.
-In the happy dataset, the {\sf wtssall} variable gives analytic survey weights.
-Figure~\ref{fig:weighting} shows the difference between the weighted and unweighted distributions of age and sex.
-The distribution is subtly different.
-
-To express that we are displaying weights, rather than counts, we extend the formula syntax by putting the weighting variable on the left-hand side: $wtssall \sim sex + age$ vs. $ \sim sex + age$.
-
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=0.5\linewidth]{wt-count}%
-    \includegraphics[width=0.5\linewidth]{wt-wtssall}%
-  \caption{(Left) Unweighted distribution (counts).  (Right) Weighted by probability weights.}
-  \label{fig:weighting}
-\end{figure}
-
-\subsection{Continuous data}
-\label{sub:continuous_data}
-
-These techniques can also be used with continuous data, if the data is binned to create a categorical variable.
-There are many different possible choices of binning algorithms, but we will focus on two:
-
-\begin{itemize}
-  \item Intervals of equal width.
-  \item Intervals containing equal numbers of points.
-\end{itemize}
-
-This gives rise to the histogram and spinogram, analogues of the bar chart and spine chart respectively.
-A long standing tradition is that no gaps are displayed between adjacent rectangles when used for originally continuous data.
-Two examples are shown in Figure \ref{fig:cont-examples}.
-
-\begin{figure}[htbp]
-  \begin{center}
-  \end{center}
-  \caption{Area plots of originally continuous data}
-  \label{fig:cont-examples}
-\end{figure}
-
-Give a few examples of non-traditional continuous-categorical plots.  Fluctuation diagram for exploring joint 2d distribution.  Mosaic for conditional.  Age and year distribution.
-
-Cat + continuous.
-
-An alternative approach is use density estimates (or conditional density estimates) and plot with continuous area plots.
-As in: Hofmann, H., Theus, M.
-(2005), Interactive graphics for visualizing conditional distributions, Unpublished Manuscript.
-
-% \subsection{Nested data}
-% \label{sub:nested_data}
-% 
-% Use space-filling!
-% 
-% We distinguish nested data from crossed data by the number of interactions with missing values.  Completely crossed data contains every combination of all variables.  Nested data contains 
-% 
-% 
-% Aka sparse vs. dense.
-% 
-% It is not always possible to determine nested data from inspection of the data alone, but may need knowledge of the data collection.
-For example, if a survey of teachers within schools was performed, and each teacher was identified by a identifier unique within a school, it might look like the data was only missing a few combinations, when in fact it is meaningless.
-% 
-% 
-% 
-% Or when you have a hierarchy.
-% 
-% Example dataset - Doug Bates teachers/schools?
-% 
-% Possibilities: create new variable which is interaction of nested vars, or drop missings/zeros.  Treemaps deal almost exclusively with nested data.
-
 \section{Labelling}
@@
 \end{itemize}
-
-\section{Conclusions} % (fold)
-\label{sec:conclusions}
-
-% section conclusions (end)
-
-% bibtool -x product-graphics -o references.bib
-\bibliography{references}
-\end{document}

### 2010-04-25 product-graphics.tex
Thrashing paper into shape

@@
 \usepackage{hyperref}
+\usepackage{booktabs}
 \usepackage{url}
@@
 
-Area plots are the graphical equivalent of contingency tables, displaying how a total is broken down into components.
+Area plots, like bar charts and mosaic plots, are the graphical equivalent of contingency tables, displaying how a total is broken down.
 
-display categorical data with a polygon, typically a rectangle, for each combination of the factors of interest, with the area proportional to the number of observations in that combination.
+Display of discrete distribution, pmf, contingency table.
+
+This paper presents a unifying framework for area plots, showing how many existing plots are interrelated, and by drawing an
+
+The paper proceeds as follows.
+Section~\ref{sec:primitives} defines the simple one- and two-d primitives used for all more complicated graphics.
+Section~\ref{sec:combination} provides a mathematical framework that shows how these simple primitives can be combined to display higher-d tables.
+Section~\ref{sec:variations} shows how these plots can be adapted to deal with missing values, weights and continuous data.
+
+To illustrate these ideas, the same data for all examples. It's a small sample
+of variables related to happiness from the general social survey ({\sc gss}).
+The {\sc gss} is a yearly cross-sectional survey of Americans. Combining
+surveys from 1972 to 2006 gives 51\,020 observations. Of the over 5\,000
+variables, we select eight related to happiness, as described in
+Table~\ref{tbl:happy}. The data is included in the supplemental materials.
+
+\begin{table}
+  \begin{center}
+  \begin{tabular}{llp{8cm}}
+    \toprule
+    Variable name & Description & Values \\
+    \midrule
+    {\sf age} & age in years & 18--89 \\
+    {\sf degree} & highest education & lt high school, high school, junior college, bachelor, graduate \\
+    {\sf finrela} & relative financial status & far above average, above average, average, below average, far below average \\
+    {\sf happy} & happiness & \key{very-happy} very happy, \key{pretty-happy} pretty happy, \key{not-too-happy} not too happy \\
+    {\sf health} & health & excellent, good, fair, poor \\
+    {\sf sex} & sex & \key{female} female, \key{male} male\\
+    {\sf wtsall} & probability weight & 0.43--6.42 \\
+    {\sf year} & year of survey & 1972--2006 \\
+    \bottomrule
+
+  \end{tabular}
+  \end{center}
+  \caption{Description of sample data.  Common plot colours are shown next to respective levels.}
+  \label{tbl:happy}
+\end{table}
+
+\section{Graphical primitives}
+\label{sec:primitives}
+
+Tilings of the plane.
+In general, we could also think of partitions of higher-dimensional spaces (e.g.
+3d or 4d), but we will need to project these down to two for viewing anyway, so there is little disadvantage to working directly in 2d.
+To derive the graphical primitives we start with three constraints:
+
+\begin{itemize}
+  \item Area must be {\bf proportional} to value.
 Because the total area for a graphic is usually constrained, an area plot serves to partition up this total, displaying proportions rather than counts.
 
-Figure \ref{fig:cat-examples} illustrates some of these plots.
-
-\begin{figure}[htbp]
-  \begin{center}
-  %  \includegraphics[scale=1]{file}
-  \end{center}
-  \caption{Area plot examples.}
-  \label{fig:cat-examples}
-\end{figure}
-
-Hierarchical partitioning of space.
-
-\section{Factorising}
-
-Basic idea is factorising table of probabilities:
-
-\begin{itemize}
-  \item as products of marginal and conditional distributions
-  \item as areas in plots
-  \item as sums of parameters in log-linear models
-\end{itemize}
-
-Some notation: let capital letters refer to the sample space, and lower case particular values.
-
-Let $f(x, y, z)$ be the 3d pmf function. We can can write this pmf as product of marginals and conditionals in many different ways.  We can decompose product graphics in a similar way.
-
-\begin{itemize}
-  \item $f(x, y, z) = f(x, y | z) f(z)$
-  \item $f(x, y, z) = f(x | y, z) f(y, z)$
-  \item $f(x, y, z) = f(x | y, z) f(y | z) f(z)$
-\end{itemize}
-
-It can be useful to display partial decompositions, i.e. just $f(x, y | z)$, which is still a 3 variable pmf, and is defined as
-
-\[ f(x, y | z) = \frac{f(x, y, z)}{\sum_{x = a, y = b} f(a, b, z)} \]
-
-so that each category of z sums to one, and for any fixed value of z, $f(x, y | z)$ is a valid pmf.
-
-\[ \forall b \sum_a f(a | b) = 1 \]
-
-Because of their hierarchical nature, an area plot of $f(x, y, z)$ also shows $f(y, z)$ and $f(z)$.
-
-Based on the modelling specification of \citet{wilkinson:1973} we define an algebra for describing these tables.
-
-\begin{itemize}
-  \item \verb|~ a| represents $f(a)$,
-  \item \verb|~ a + b| represents $f(a, b)$
-  \item \verb!~ a | b! represents $f(a | b)$
-  \item \verb!~ . | a! is a special case of $f(a) = \frac{1}{\#A}$. 
-\end{itemize}
-
-By convention we will write the top most level of the product plot on the left, and lower levels to the right.
-While $f(x, y, z)$ and $f(z, y, x)$ display the same counts, their interpretation is different because of which marginal tables you see.
-
-Figure~\ref{fig:fact-simple} illustrates these ideas with the distribution of happiness and sex.
-We start with a plot where area broken down by happiness.
-We can then continue to factor this plot to also display sex - the area of each rectangle is proportional to the number of people who are happy with that sex, and the construction allows us to see the conditional distribution of sex within happy.
-In this case the breaks line up, suggesting that there is no difference in sex, given happiness.
-This could be followed up by a formula hypothesis test.
-
-Removing the marginal effect of happiness, as in the column on the far right, makes each of the top-level categories the same size, and makes it easier to compare the conditional distributions.
-It's not that important in this case, but in the bar chart below, it allows us to see the conditional distribution, and in the case when there are many levels with very different probabilities it can make the plot much easier to read.
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=0.33\linewidth]{fact-happy}%
-  \includegraphics[width=0.33\linewidth]{fact-happy-sex}%
-  \includegraphics[width=0.33\linewidth]{fact-happy|sex}
-
-  \includegraphics[width=0.33\linewidth]{fact-happy-2}%
-  \includegraphics[width=0.33\linewidth]{fact-happy-sex-2}%
-  \includegraphics[width=0.33\linewidth]{fact-happy|sex-2}
-
-  \caption{Plots of the distribution of happiness and sex (\key{male} male, \key{female} female).}
-  \label{fig:fact-simple}
-\end{figure}
-
-These techniques are also closely related to dimensional stacking \citep{leblanc:1990}
-
-% Other papers to cite:
-% 
-% Generalised treemaps: \citep{vliegen:2006}
-% Cocktailmaps: \citep{ahlberg:1996}
-% Extension to mosaic: \citep{friendly:1999,hofmann:2000}
-% Connection to models: \citep{theus:1999,hofmann:2001}
-
-% Hierarchical coordinate systems 
-%
-% What are the common components of all these plots, and how can be describe
-% them in a succinct way that also allows us to describe what tasks each
-% display is good for?  We propose that there is one key feature that makes
-% these plots special.  That is there their coordinate system, which is
-% hierarchical.  
-%
-% In most coordinate systems that we are used to, such as Cartesian or polar
-% coordinates, are symmetric in the sense that you can locate a point by
-% starting from either dimension.  For example, in Cartesian coordinates you
-% can identify either the x- or y-coordinate first, and the other next.  This
-% also implies a distance symmetry when projected onto 2D Euclidean geometry.
-% When you vary one coordinate by a small amount (holding all others constant)
-% points in the projective geometry will be close.
-%
-% Neither of these properties hold for hierarchical coordinate systems. 
-%
-% We will discuss hierachical coordinate systems in general, and then present
-% a specific hierachical coordinate system which gives rise to the area plots
-% described previously.
-
-\section{Data}
-\label{sec:data}
-
-Two ways of representing data:
-
-\begin{itemize}
-  \item Multi-dimensional contingency table
-  \item Data frame, with column of counts
-\end{itemize}
-
-The two are equivalent, but it's usually computational easier to work with the data frame version, because variables are explicit, and mathematically easier to work with the multidimensional array because multiplication is nicely defined.
-Also deals better with sparsity and is common input to modelling algorithms.
-
-Define the conditioning operator which makes the distribution uniform.
-
-In this paper, we'll use a small sample from the general social survey ({\sc gss}) focussing on happiness. 51\,020 observations from cross-section survey given yearly from 1972 to 2006.
-
-\begin{itemize}
-  \item Happy.  Discrete: ``very happy'', ``pretty happy'', ``not too happy''
-  \item Age (in years), 18--89.
-  \item Sex: ``female'', ``male''
-  \item Degree: ``lt high school'', ``high school'', ``junior college'', ``bachelor'', ``gradudate''
-  \item Relative financial status {\sf finrela}: ``far above average'', ``above average'', ``average'', ``below average'', ``far below average''
-  \item Health: ``excellent'', ``good'', ``fair'', ``poor''
-  \item Year: 1972--2006.
-  \item Probability weight ({\sf wtsall}): 0.43--6.42.
-\end{itemize}
-
-\section{Display}
-\label{sec:display}
-
-Once we have made the decision to map area to proportion, we need to impose some constraints on the set of possible tilings of the plane.
-In general, we could also think of partitions of higher-dimensional spaces (e.g.
-3d or 4d), but we will need to project these down to two for viewing anyway, so there is little disadvantage to working in 2d.
-
-Some constraints are implied by our perceptual capabilities and others by geometry. On top of the constraint that area must be proportional to value, we add two additional constraints:
-
-\begin{itemize}
-
-  \item Partitions must be rectangular.
+  \item Partitions must be {\bf rectangular}.
 It is easier to compare the areas of simple shape (i.e.
@@
 
-  \item Partitions must be disjoint.
+  \item Partitions must be {\bf disjoint}.
 To be able to see the complete area of each rectangle they must be non-overlapping, and each split must be nested within the shapes of the previous level.
@@
 
-Note, however, that each of these constraints can be profitably relaxed. See Section~\ref{sec:relax} for details.
+\noindent Each of these three constraints can be profitably relaxed, as detailed in Section~\ref{sec:variations}.
 
-Let $a(i)$ give the area assigned to category $i$, which has proportion $f(i)$.
+The following sections describe the partitions generated by these constraints.
+Section~\ref{sub:part-1d} discusses 1d partitions, bars, spines and tiles, and  Section~\ref{sub:part-2d} discusses the 2d partition, the fluct.
+
+% Let $a(i)$ give the area assigned to category $i$, which has proportion $f(i)$.
 Since we are using rectangles, $a(i) = k \dot a(i) = k \dot w(i) \times h(i)$ where $w$ and $h$ represent width and height respectively.
@@
 This is useful because it is only necessary to inspect one border to compare magnitudes.
-
-The following two sections describe the partitions currently used to produce area graphics.
-Section~\ref{sub:part-1d} discusses partitions for 1d pmfs: bars, spines and tiles, Section~\ref{sub:part-2d} discusses one partition for 2d pmfs and Section~\ref{sub:part-nd} describes how the 1d and 2d primitives can be recursively combined to generate a partitioning for any higher dimension pmf and discusses how existing area graphics fall into this framework.
 
@@
 Bars allow comparison between absolute numbers, but are not space filling.
-They occupy sum(values / max(values)) of the total area.
+Bars are not space filling, occupying $\sum(p_i / \max(p_i))$ of the total area.
 Bars can be arranged horizontally (``hbar'') or vertically (``vbar'').
-Not space filling.
 
-  \item {\bf spine}s: width is proportional to value, height equally divides space.
-Splines are space filling, but make it more difficult to compare absolute counts.
+  \item {\bf spine}s: width is proportional to value, height is fixed.
+Splines are space filling.
 Splines can be arranged horizontally (``hspine''), vertically (``vspine''), or can automatically pick their orientation (``spine'').
-Space filling.
 
   \item {\bf tile}s: tile the plane with rectangles, trying to keep the aspect ratio of each rectangle close to 1.
-Comparing the areas is the most difficult perceptually, and are most useful when there are a large number of categories.
-The is partitioning used by the squarified treemap \citep{bruls:1999}.
-Not space filling.
+There are no restrictions on height or width, just the overall area.
+This is partitioning used by the squarified treemap \citep{bruls:1999}.
 
 \end{itemize}
+
+Each of these three displays has different strengths and weaknesses.
+The perceptual task associated with bar charts is the simplest, but they do not occupy all the area.
+Spines and tiles are harder to read, but occupy the complete space and work better recursively \citep{heer:2010}.
 
@@
 
-We are currently aware of one 2d primitive, the {\bf fluct}: where width and height are proportional to the square root of the proportion.
-Flucts allow comparisons of relative magnitudes numbers in two directions.
-Each rectangle is arranged on a regular grid formed by the levels of the two variables, and thus the fluct is not space filling.
-A fluct displays a joint 2d distribution, unlike the combination of any two 1d partitions which display a conditional distribution.
+We are currently aware of one 2d primitive, the {\bf fluct}, which has both height and width proportional to the square root of the proportion.
+Flucts allow comparisons of relative magnitudes in two directions.
+Each rectangle is arranged on a regular grid formed by the levels of the two variables, and so is not space filling.
 
@@
     \includegraphics[width=0.5\linewidth]{part-fluct-cond}
-  \caption{(Left) Distribution of age (in decades) and marital status.
-All combinations of the two variables are present.
-(Right) Distribution of health status and year of survey.
-Health status was not recorded for three years.}
+  \caption{(Left) Distribution of age (in decades) and marital status. (Right) Equal bin-size plot showing health status and survey year. Health status was not recorded for three years.}
   \label{fig:fluct}
@@
 
-\subsection{Combinations}
-\label{sub:part-nd}
+\section{Factorising area plots}
+\label{sec:combination}
 
-To visualise higher-dimensional pmfs, we recursively partition using the previously described primitives.
+To construct plots for discrete pdfs with more than two dimensions, we need a mechanism to factorise them into low dimensional products. This factorisation has three roles:
+
+\begin{itemize}
+  \item as products of marginal and conditional distributions
+  \item as areas in plots
+  \item as sums of parameters in log-linear models
+\end{itemize}
+
+Let $f(x, y, z)$ be the 3d pmf function. We can can write this pmf as product of marginals and conditionals in many different ways.  We can decompose product graphics in a similar way.
+
+\begin{itemize}
+  \item $f(x, y, z) = f(x, y | z) f(z)$
+  \item $f(x, y, z) = f(x | y, z) f(y, z)$
+  \item $f(x, y, z) = f(x | y, z) f(y | z) f(z)$
+\end{itemize}
+
+% It can be useful to display partial decompositions, i.e. just $f(x, y | z)$, which is still a 3 variable pmf, and is defined as
+%
+% \[ f(x, y | z) = \frac{f(x, y, z)}{f(z)} \]
+%
+% so that each category of z sums to one, and for any fixed value of z, $f(x, y | z)$ is a valid pmf.
+%
+% \[ \forall b \sum_a f(a | b) = 1 \]
+%
+% Because of their hierarchical nature, an area plot of $f(x, y, z)$ also shows $f(y, z)$ and $f(z)$.
+
+Based on the modelling specification of \citet{wilkinson:1973} we define an algebra for describing these tables.
+
+\begin{itemize}
+  \item \verb|~ a| represents $f(a)$,
+  \item \verb|~ a + b| represents $f(a, b)$
+  \item \verb!~ a | b! represents $f(a | b)$
+  \item \verb!~ . | a! is a special case of $f(a) = \frac{1}{\#A}$.
+\end{itemize}
+
+By convention we will write the top most level of the product plot on the left, and lower levels to the right.
+While $f(x, y, z)$ and $f(z, y, x)$ display the same counts, their interpretation is different because of which marginal tables you see.
+
+Figure~\ref{fig:fact-simple} uses the distribution of sex and happiness to illustrates these ideas.
+From left to right, we have plots of \verb|~ happy|, \verb!~ happy | sex!
+and their product \verb|~ happy + sex|.
+Observe that the heights and widths of the rectangles are multiplied in the same way as the components of the pmf.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.3\linewidth]{fact-happy} $\times$ %
+  \includegraphics[width=0.3\linewidth]{fact-happy|sex} = %
+  \includegraphics[width=0.3\linewidth]{fact-happy-sex}
+
+  \includegraphics[width=0.3\linewidth]{fact-happy-2} $\times$ %
+  \includegraphics[width=0.3\linewidth]{fact-happy|sex-2} = %
+  \includegraphics[width=0.3\linewidth]{fact-happy-sex-2}
+
+  \caption{Plots of the distribution of happiness and sex (\key{male} male, \key{female} female).}
+  \label{fig:fact-simple}
+\end{figure}
+
+% To visualise higher-dimensional pmfs, we recursively partition using the previously described primitives.
 That is, to display $f(x, y, z)$ we first create a partition based on $f(z)$ then partition each of those pieces based on $f(y | z)$ and again by $f(x | y, z)$.
+
 Figure \ref{fig:recursive} illustrates this process with a graphic that explores whether, conditional on marital status, there is a difference in happiness between men and women.
-
-% Heike: may be this would be a good spot for some of your Kronecker
-% product theory?
 
@@
 \end{figure}
+
+Different combinations of partitions reveal different features of the data.
+Take for example, the distribution of age and marital status, as shown in Figure~\ref{fig:fluct}.
+Instead of visualisation the joint distribution with a fluct, we could focus on the conditional distribution of marital status, given age, or vice versa.
+Figure~\ref{fig:marital} shows two ways to do this.
+The left plot shows $\sim marital + age$ with a vspine nested in a hspine, and the right plot $\sim age + martial$ with a hbar nested in a vspine.
+These displays show the same data but are arranged in such a way to make different comparisons easier.
+
+\begin{figure}[htbp]
+  \centering
+    \includegraphics[width=0.5\linewidth]{part-marital-1}%
+    \includegraphics[width=0.5\linewidth]{part-marital-2}
+  \caption{(Left) The joint distribution of marital status and age (in decades) partitioned by a vspine and hspine.
+(Right) The joint distribution of age and marital status partitioned by a vspine and hbar.}
+  \label{fig:marital}
+\end{figure}
+
+Conditioning is also an important tool, because it allows one to remove relationships that are well-known or uninteresting.
+Figure~\ref{fig:part-cond} uses a fluct and a vspine to explore the relationship between happiness, health and financial status.
+The left plot displays raw proportions, showing that most people are in good health and average financial standing.
+However, it is difficult to see how happiness varies within these conditions because the perceptual task is comparing areas.
+Conditioning on financial status and health produces the plot on the right (an equal bin size plot) and makes it easier to see the conditional distribution of happiness, because the perceptual task is simpler: comparing length on a common non-aligned scales.
+Here we can see people tend to be happier the healthier and richer they are.
+
+\begin{figure}[htbp]
+  \centering
+    \includegraphics[width=0.5\linewidth]{part-fluctuation}%
+    \includegraphics[width=0.5\linewidth]{part-equal-area}
+  \caption{Two displays of $\sim happy + health + finrela$.
+(Left) Unconditional values partitioned with a vspine and fluct.
+(Right) Same partitions, but health and relative financial status conditioned to uniform distribution.
+We can no longer see the joint distribution of health and financial status, but it is much easier to see the conditional distribution of happiness: healthier and richer people are happier.
+(Maybe money does buy happiness?).
+(\key{not-too-happy} Not too happy, \key{pretty-happy} pretty happy, \key{very-happy} very happy)}
+  \label{fig:part-cond}
+\end{figure}
+
+% http://peltiertech.com/Utility/ClusterStackUtility.html
+
+\subsection{Existing plot types}
+
+% Other papers to cite:
+%
+% Generalised treemaps: \citep{vliegen:2006}
+% Cocktailmaps: \citep{ahlberg:1996}
+% Extension to mosaic: \citep{friendly:1999,hofmann:2000}
+% Connection to models: \citep{theus:1999,hofmann:2001}
 
@@
 
-Different combinations of partitions reveal different features of the data.
-Take for example, the distribution of age and marital status, as shown in Figure~\ref{fig:fluct}.
-Instead of visualisation the joint distribution with a fluct, we could focus on the conditional distribution of marital status, given age, or vice versa.
-Figure~\ref{fig:marital} shows two ways to do this.
-The left plot shows $\sim marital + age$ with a vspine nested in a hspine, and the right plot $\sim age + martial$ with a hbar nested in a vspine.
-These displays show the same data but are arranged in such a way to make different comparisons easier.
-
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=0.5\linewidth]{part-marital-1}%
-    \includegraphics[width=0.5\linewidth]{part-marital-2}
-  \caption{(Left) The joint distribution of marital status and age (in decades) partitioned by a vspine and hspine.
-(Right) The joint distribution of age and marital status partitioned by a vspine and hbar.}
-  \label{fig:marital}
-\end{figure}
-
-Conditioning is also an important tool, because it allows one to remove relationships that are well-known or uninteresting.
-Figure~\ref{fig:part-cond} uses a fluct and a vspine to explore the relationship between happiness, health and financial status.
-The left plot displays raw proportions, showing that most people are in good health and average financial standing.
-However, it is difficult to see how happiness varies within these conditions because the perceptual task is comparing areas.
-Conditioning on financial status and health produces the plot on the right (known as an equal bin size plot) and makes it easier to see the conditional distribution of happiness, because the perceptual task is simpler: comparing length on a common non-aligned scales.
-Here we can see people tend to be happier the healthier and richer they are.
-
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=0.5\linewidth]{part-fluctuation}%
-    \includegraphics[width=0.5\linewidth]{part-equal-area}
-  \caption{Two displays of $\sim happy + health + finrela$.
-(Left) Unconditional values partitioned with a vspine and fluct.
-(Right) Same partitions, but health and relative financial status conditioned to uniform distribution.
-We can no longer see the joint distribution of health and financial status, but it is much easier to see the conditional distribution of happiness: healthier and richer people are happier.
-(Maybe money does buy happiness?).
-(\key{not-too-happy} Not too happy, \key{pretty-happy} pretty happy, \key{very-happy} very happy)}
-  \label{fig:part-cond}
-\end{figure}
-
-% http://peltiertech.com/Utility/ClusterStackUtility.html
-
 Trellis graphics \citep{becker:1996}, as known as lattice or facetted or
@@
 
-\section{Relaxing constraints}
-\label{sec:relax}
+These techniques are also closely related to dimensional stacking \citep{leblanc:1990}
 
-There are a number of cases in which it is useful to relax the constraints that we defined above.  These are described in sequence below.
+\section{Modelling}
+\label{sec:models}
+
+\section{Variations and extensions}
+\label{sec:variations}
+
+On top of this basic framework, it is useful to consider a few variations and extensions.
+Three are based on variations of the underlying data (missing data, weighted data, and continuous data), and three are based on relaxing the display constraints (area not proportional to weight, non-disjoint partitions, non-rectangular partitions).
+
+\subsection{Missing values}
+\label{sub:missing_values}
+
+\subsection{Weighting}
+\label{sub:weighting}
+
+We have assumed that the probabilities represent counts, but without loss of generality, we can use any set of non-negative, additive weights.
+For example, in the happy dataset, the {\sf wtssall} variable gives analytic survey weights.
+Figure~\ref{fig:weighting} shows the difference between the weighted and unweighted distributions of age and sex.
+The distribution is subtly different.
+\citep{unwin:2007} describes the applications of weighted plots in more detail.
+
+To express that we are displaying weights, rather than counts, we extend the formula syntax by putting the weighting variable on the left-hand side: $wtssall \sim sex + age$ vs. $ \sim sex + age$.
+
+\begin{figure}[htbp]
+  \centering
+    \includegraphics[width=0.5\linewidth]{wt-count}%
+    \includegraphics[width=0.5\linewidth]{wt-wtssall}%
+  \caption{Joint distribution of age and sex (\key{male} male, \key{female} female). (Left) counts and (right) probability weights.}
+  \label{fig:weighting}
+\end{figure}
+
+\subsection{Continuous data}
+\label{sub:continuous_data}
+
+These techniques can also be used with continuous data, if the data is binned to create a categorical variable.
+There are many different possible choices of binning algorithms, but we will focus on two:
+
+\begin{itemize}
+  \item Intervals of equal width.
+  \item Intervals containing equal numbers of points.
+\end{itemize}
+
+This gives rise to the histogram and spinogram, continuous analogues of the bar chart and spine chart.
+A long standing tradition is that no gaps are displayed between adjacent rectangles when used for continuous data.
+Two examples are shown in Figure \ref{fig:cont-examples}.
+
+\begin{figure}[htbp]
+  \centering
+  \caption{Area plots of originally continuous data}
+  \label{fig:cont-examples}
+\end{figure}
+
+Example exploring joint and conditional distributions of age and year.
+
+% An alternative approach is use density estimates (or conditional density estimates) and plot with continuous area plots.
+As in: Hofmann, H., Theus, M.
+(2005), Interactive graphics for visualizing conditional distributions, Unpublished Manuscript.
 
@@
 By controlling this cap it is possible to zoom in on the smallest values.
+- called censored zooming by Antony and Martin.
 
@@
 For example, we could display the square root of the weights, to stabilise the variance of the areas.
+Tukey applied this technique to histograms to create rootograms.
+
 If values span multiple orders of magnitude it may be useful to map log value to area.
@@
 
-\section{Variations}
-\label{sec:variations}
-
-On top of this basic framework, there are a few variations designed for dealing with special types of data: missing data, weighted data, and continuous data.  These variations are described below.
-
-\subsection{Missing values}
-\label{sub:missing_values}
-
-\subsection{Weighting}
-\label{sub:weighting}
-
-We have assumed that the probabilities represent counts, but without loss of generality, we can use any set of non-negative, additive weights.
-For example, in the happy dataset, the {\sf wtssall} variable gives analytic survey weights.
-Figure~\ref{fig:weighting} shows the difference between the weighted and unweighted distributions of age and sex.
-The distribution is subtly different.
-\citep{unwin:2007} describes the applications of weighted plots in more detail.
-
-To express that we are displaying weights, rather than counts, we extend the formula syntax by putting the weighting variable on the left-hand side: $wtssall \sim sex + age$ vs. $ \sim sex + age$.
-
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=0.5\linewidth]{wt-count}%
-    \includegraphics[width=0.5\linewidth]{wt-wtssall}%
-  \caption{Joint distribution of age and sex (\key{male} male, \key{female} female). (Left) counts and (right) probability weights.}
-  \label{fig:weighting}
-\end{figure}
-
-\subsection{Continuous data}
-\label{sub:continuous_data}
-
-These techniques can also be used with continuous data, if the data is binned to create a categorical variable.
-There are many different possible choices of binning algorithms, but we will focus on two:
-
-\begin{itemize}
-  \item Intervals of equal width.
-  \item Intervals containing equal numbers of points.
-\end{itemize}
-
-This gives rise to the histogram and spinogram, analogues of the bar chart and spine chart respectively.
-A long standing tradition is that no gaps are displayed between adjacent rectangles when used for originally continuous data.
-Two examples are shown in Figure \ref{fig:cont-examples}.
-
-\begin{figure}[htbp]
-  \begin{center}
-  \end{center}
-  \caption{Area plots of originally continuous data}
-  \label{fig:cont-examples}
-\end{figure}
-
-Give a few examples of non-traditional continuous-categorical plots.  Fluctuation diagram for exploring joint 2d distribution.  Mosaic for conditional.  Age and year distribution.
-
-Cat + continuous.
-
-An alternative approach is use density estimates (or conditional density estimates) and plot with continuous area plots.
-As in: Hofmann, H., Theus, M.
-(2005), Interactive graphics for visualizing conditional distributions, Unpublished Manuscript.
-
-% \subsection{Nested data}
-% \label{sub:nested_data}
-% 
-% Use space-filling!
-% 
-% We distinguish nested data from crossed data by the number of interactions with missing values.  Completely crossed data contains every combination of all variables.  Nested data contains 
-% 
-% 
-% Aka sparse vs. dense.
-% 
-% It is not always possible to determine nested data from inspection of the data alone, but may need knowledge of the data collection.
-For example, if a survey of teachers within schools was performed, and each teacher was identified by a identifier unique within a school, it might look like the data was only missing a few combinations, when in fact it is meaningless.
-% 
-% 
-% 
-% Or when you have a hierarchy.
-% 
-% Example dataset - Doug Bates teachers/schools?
-% 
-% Possibilities: create new variable which is interaction of nested vars, or drop missings/zeros.  Treemaps deal almost exclusively with nested data.
-
-\section{Labelling}
-\label{sec:legends}
-
-Due to the hierarchical nature of these plots, it's difficult to label them in an informative way.
-We take a pragmatic approach, and label a single horizontal and single vertical split, where possible, using colour to identify a third variable, if necessary.
-
-Understanding the composition of these plots is most easy when:
-
-\begin{itemize}
-
-  \item Exploration: You have interactive control over the creation of the plot, and can interactively query regions of interest
-
-  \item Communication: The plot is built up step-by-step, and shows the minimum number of splits.  It may be easier to understand a single complex plot when factored into two or more simpler plots.
-
-\end{itemize}
-
-To create the labels in used in this paper, we inspect each level of the factorisation looking for the first level in which there are rows or columns.
-We do this rather inspecting the partitions that make up the plot, because there are many possible combinations that create columns: hsplines, hbars, vsplines, vbars for uniform data, flucts, ...
-and then can occur at any level in the hierarchy depending on what's come before.
-
-Apart from textual labels on the x-axis, other work has labelled individual cells with attributes of the data:
-
-\begin{itemize}
-  \item colour (map of the market)
-  \item texture, such as in sieve plots \citep{friendly:2000}
-  \item photographs \citep{bederson:2001}
-  \item text (tables)
-  \item embedded plots (time series in lab escape)
-\end{itemize}
-
-Or attributes of the hierarchy:
-
-\begin{itemize}
-  \item Spacing / borders
-  \item Shading
-  \item Cascading
-  \item Labelling
-\end{itemize}
-
-\section{Conclusions} % (fold)
-\label{sec:conclusions}
-
-% section conclusions (end)
+\section{Conclusions}
 

### 2010-04-25 plots-partition.r

@@ -9,7 +9,7 @@ ggsave("plots/part-hbar.pdf", width = 4, height = 4)
 prodplot(happy, ~ happy, "hspine", na.rm = TRUE) + fill_happy
 ggsave("plots/part-hspine.pdf", width = 4, height = 4)
 
-prodplot(happy, ~ happy, "treemap", na.rm = TRUE) + fill_happy
+prodplot(happy, ~ happy, "tile", na.rm = TRUE) + fill_happy
 ggsave("plots/part-treemap.pdf", width = 4, height = 4)
 
 # 2d partitions --------------------------------------------------------------

### 2010-04-26 product-graphics.tex
Slowly getting it together

@@
 
-Area plots, like bar charts and mosaic plots, are the graphical equivalent of contingency tables, displaying how a total is broken down.
-
-Display of discrete distribution, pmf, contingency table.
-
-This paper presents a unifying framework for area plots, showing how many existing plots are interrelated, and by drawing an
+Area plots, like bar charts, mosaic plots and treemaps, are the graphical equivalent of contingency tables, visualising a discrete distribution by mapping proportion to area.
+This paper presents a unifying framework for area plots, showing how many existing plots are interrelated, and developing a set of tools for creating new plots.
+In particular, we create a graphical factorisation of area plots by analogy to the mathematical factorisation of joint distributions into products of marginals and conditionals.
 
 The paper proceeds as follows.
-Section~\ref{sec:primitives} defines the simple one- and two-d primitives used for all more complicated graphics.
-Section~\ref{sec:combination} provides a mathematical framework that shows how these simple primitives can be combined to display higher-d tables.
-Section~\ref{sec:variations} shows how these plots can be adapted to deal with missing values, weights and continuous data.
+Section~\ref{sec:primitives} described the simple one- and two-d graphical primitives that define all area plots.
+Section~\ref{sec:combination} provides a mathematical framework that shows how these simple primitives can be combined to display higher-d distributions.
+Section~\ref{sec:variations} extends these techniques for other types of data, and shows variations based on different low-d graphical primitives.
 
-To illustrate these ideas, the same data for all examples. It's a small sample
-of variables related to happiness from the general social survey ({\sc gss}).
-The {\sc gss} is a yearly cross-sectional survey of Americans. Combining
-surveys from 1972 to 2006 gives 51\,020 observations. Of the over 5\,000
-variables, we select eight related to happiness, as described in
-Table~\ref{tbl:happy}. The data is included in the supplemental materials.
+To illustrate these ideas, we use the same data for all examples.
+It's a small sample of variables related to happiness from the general social survey ({\sc gss}).
+The {\sc gss} is a yearly cross-sectional survey of Americans, run from 1976.
+We combine data for 25 years to to give gives 51\,020 observations, and of the over 5\,000 variables, we select eight related to happiness, as described in Table~\ref{tbl:happy}.
+The data is included in the supplemental materials.
 
-\begin{table}
+\begin{table}[htb]
   \begin{center}
-  \begin{tabular}{llp{8cm}}
+  \begin{tabular}{llp{9cm}}
     \toprule
-    Variable name & Description & Values \\
+    Variable & Description & Values \\
     \midrule
@@
     \bottomrule
-
   \end{tabular}
@@
 
-Tilings of the plane.
-In general, we could also think of partitions of higher-dimensional spaces (e.g.
-3d or 4d), but we will need to project these down to two for viewing anyway, so there is little disadvantage to working directly in 2d.
+All area plots correspond to tilings of the plane.
+We could also consider partitions of higher-d spaces (e.g.
+3d or 4d), but given that we have to project these down to 2d for viewing, there is little disadvantage to working directly in 2d.
+
 To derive the graphical primitives we start with three constraints:
@@
 \begin{itemize}
-  \item Area must be {\bf proportional} to value.
-Because the total area for a graphic is usually constrained, an area plot serves to partition up this total, displaying proportions rather than counts.
+  \item Area must be {\bf proportional} to value. Because the total area for a graphic is usually constrained, an area plot partitions this total, displaying proportions rather than counts.
 
@@
 It is easier to compare the areas of simple shape (i.e.
-rectangle vs polygon), and even better when one the length of one border is fixed, so we only need compare the length of the other \citep{cleveland:1984}.
+rectangle vs polygon), and even easier when the length of one border is fixed, so we only need compare the length of the other \citep{cleveland:1984}.
 Rectangular partitions are also recursive in the sense that we can always tile a rectangle with smaller rectangles.
@@
   \item Partitions must be {\bf disjoint}.
-To be able to see the complete area of each rectangle they must be non-overlapping, and each split must be nested within the shapes of the previous level.
-Note that we do not impose the constraint that the tiling be space-filling.
-Space filling methods use will ensure that areas are larger, but make it harder to compare magnitudes because scales no longer align.
+To be able to see the complete area of each rectangle they must be non-overlapping.
+This does not imply that the tiling be space-filling: generally, space filling methods use space more efficiently, but make areas harder to compare.
 
@@
   \item {\bf bar}s: height is proportional to value, width equally divides space.
-Bars allow comparison between absolute numbers, but are not space filling.
-Bars are not space filling, occupying $\sum(p_i / \max(p_i))$ of the total area.
+Bars are not space filling, occupying $\mbox{mean}(x - \max(x))$ of the total area.
 Bars can be arranged horizontally (``hbar'') or vertically (``vbar'').
 
-  \item {\bf spine}s: width is proportional to value, height is fixed.
+  \item {\bf spine}s: width is proportional to value, height occupies full range.
 Splines are space filling.
-Splines can be arranged horizontally (``hspine''), vertically (``vspine''), or can automatically pick their orientation (``spine'').
+Splines can be arranged horizontally (``hspine''), vertically (``vspine''), or can automatically pick their orientation (``spine'') by splitting the largest dimension.
+The appearance of bars and splines is identical for uniformly distributed data.
 
-  \item {\bf tile}s: tile the plane with rectangles, trying to keep the aspect ratio of each rectangle close to 1.
-There are no restrictions on height or width, just the overall area.
+  \item {\bf tile}s: no restrictions on height or width, just tile the plane with rectangles, trying to keep the aspect ratio of each rectangle close to 1.
 This is partitioning used by the squarified treemap \citep{bruls:1999}.
@@
 The perceptual task associated with bar charts is the simplest, but they do not occupy all the area.
-Spines and tiles are harder to read, but occupy the complete space and work better recursively \citep{heer:2010}.
+Spines and tiles are harder to read \citep{heer:2010}, but occupy the complete space and work better recursively.
 
@@
 
-The appearance of bars and splines is identical for uniformly distributed data.
-
 \subsection{2d primitives}
@@
 
-We are currently aware of one 2d primitive, the {\bf fluct}, which has both height and width proportional to the square root of the proportion.
-Flucts allow comparisons of relative magnitudes in two directions.
-Each rectangle is arranged on a regular grid formed by the levels of the two variables, and so is not space filling.
-
-One special case of the fluct is the equal bin size plot which occurs when the two variables are jointly uniformly distributed, usually as a result of conditioning.
-The equal bin size plot is useful to visualise missing combinations, or when supplemented with a third variable, to visualise the distribution conditional on two other variables.
+We are currently aware of one 2d primitive, the {\bf fluct}, which has both height and width proportional to the square root of the proportion \citep{hofmann:2000}.
+Each rectangle is arranged on a regular grid formed by the levels of the two variables, allowing comparisons in two directions.
+A special case of the fluct is the equal bin size plot which occurs when the two variables are jointly uniformly distributed, usually as a result of the conditioning described in the next section.
+The equal bin size plot is useful to visualise missings.
 
@@
     \includegraphics[width=0.5\linewidth]{part-fluct-cond}
-  \caption{(Left) Distribution of age (in decades) and marital status. (Right) Equal bin-size plot showing health status and survey year. Health status was not recorded for three years.}
+  \caption{(Left) A fluct showing distribution of age (in decades) and marital status.
+(Right) Equal bin-size plot showing health status and survey year.
+Health status was not recorded for three years.}
   \label{fig:fluct}
@@
 
-To construct plots for discrete pdfs with more than two dimensions, we need a mechanism to factorise them into low dimensional products. This factorisation has three roles:
+To construct plots for higher-dimensional tables, we need a way to decompose them into 1d and 2d components.
+To do we use some very basic properties of probability.
+For example, let $f(x, y, z)$ be a 3d pmf.
+We can write this pmf as product of marginals and conditionals:
 
 \begin{itemize}
-  \item as products of marginal and conditional distributions
-  \item as areas in plots
-  \item as sums of parameters in log-linear models
+  \item $f(x, y, z) = f(z) f(x, y | z)$
+  \item $f(x, y, z) = f(y, z) f(x | y, z) $
+  \item $f(x, y, z) = f(z) f(y | z) f(x | y, z)$
 \end{itemize}
 
-Let $f(x, y, z)$ be the 3d pmf function. We can can write this pmf as product of marginals and conditionals in many different ways.  We can decompose product graphics in a similar way.
+\noindent Each of these decompositions is made up of products of one or two-d pmfs that can visualised used one of the graphical primitives described previously.
+Note that while $f(x, y, z)$ and $f(z, y, x)$ represent the same pmf, the factorisation will be different because you will get different marginal tables.
 
-\begin{itemize}
-  \item $f(x, y, z) = f(x, y | z) f(z)$
-  \item $f(x, y, z) = f(x | y, z) f(y, z)$
-  \item $f(x, y, z) = f(x | y, z) f(y | z) f(z)$
-\end{itemize}
-
-% It can be useful to display partial decompositions, i.e. just $f(x, y | z)$, which is still a 3 variable pmf, and is defined as
-%
-% \[ f(x, y | z) = \frac{f(x, y, z)}{f(z)} \]
-%
-% so that each category of z sums to one, and for any fixed value of z, $f(x, y | z)$ is a valid pmf.
-%
-% \[ \forall b \sum_a f(a | b) = 1 \]
-%
-% Because of their hierarchical nature, an area plot of $f(x, y, z)$ also shows $f(y, z)$ and $f(z)$.
-
-Based on the modelling specification of \citet{wilkinson:1973} we define an algebra for describing these tables.
+Writing the full pmf is tedious, so we define an algebra based on the modelling specification of \citet{wilkinson:1973}:
 
@@
   \item \verb!~ a | b! represents $f(a | b)$
-  \item \verb!~ . | a! is a special case of $f(a) = \frac{1}{\#A}$.
+  \item \verb!~ . | a! is a special case, $f(a) = \frac{1}{\#A}$.
 \end{itemize}
 
-By convention we will write the top most level of the product plot on the left, and lower levels to the right.
-While $f(x, y, z)$ and $f(z, y, x)$ display the same counts, their interpretation is different because of which marginal tables you see.
-
-Figure~\ref{fig:fact-simple} uses the distribution of sex and happiness to illustrates these ideas.
+Figure~\ref{fig:fact-simple} uses the distribution of sex and happiness to illustrate these ideas.
 From left to right, we have plots of \verb|~ happy|, \verb!~ happy | sex!
-and their product \verb|~ happy + sex|.
+and the product \verb|~ happy + sex|.
 Observe that the heights and widths of the rectangles are multiplied in the same way as the components of the pmf.
@@
 That is, to display $f(x, y, z)$ we first create a partition based on $f(z)$ then partition each of those pieces based on $f(y | z)$ and again by $f(x | y, z)$.
+
+\subsection{Examples}
+\label{sub:examples}
 
@@
     \includegraphics[width=0.5\linewidth]{part-equal-area}
-  \caption{Two displays of $\sim happy + health + finrela$.
-(Left) Unconditional values partitioned with a vspine and fluct.
-(Right) Same partitions, but health and relative financial status conditioned to uniform distribution.
-We can no longer see the joint distribution of health and financial status, but it is much easier to see the conditional distribution of happiness: healthier and richer people are happier.
-(Maybe money does buy happiness?).
+  \caption{(Left) $\sim happy + health + finrela$, partitioned with a vspine and fluct.
+(Right) $\sim happy \,|\, health + finrela$.
+We can no longer see the joint distribution of health and financial status, but it is much easier to see the conditional distribution of happiness.
+Healthier and richer people are happier: maybe money does buy happiness?
 (\key{not-too-happy} Not too happy, \key{pretty-happy} pretty happy, \key{very-happy} very happy)}
@@
 
-% http://peltiertech.com/Utility/ClusterStackUtility.html
+\subsection{Existing plot types}
 
-\subsection{Existing plot types}
+You may have noticed that the previous plots resemble area plots that have previously be named. The one-d primitives already have their own names:
 
@@
 %
-% Generalised treemaps: \citep{vliegen:2006}
-% Cocktailmaps: \citep{ahlberg:1996}
-% Extension to mosaic: \citep{friendly:1999,hofmann:2000}
-% Connection to models: \citep{theus:1999,hofmann:2001}
-
-The sequence of partitioning choices defines different named area plots:
+%
 
 \begin{itemize}
-  \item {\bf Barchart} (1d). 1 hbar.
+  \item {\bf Bar chart} (1d). 1 hbar.
   \item {\bf Column chart} (1d). 1 vbar.
   \item {\bf Spineplot} (1d). 1 vspine.
+  \item {\bf Fluctuation} diagram (2d): fluct.
+\end{itemize}
 
-  \item {\bf Stacked} barchart (2d). 1 hbar and 1 vspine.
-  \item {\bf Nested} barchart (2d).  2 hbars. \citep{peltier:2009}
-  \item {\bf Fluctuation} diagram (2d): fluct.
+\noindent And many more plots correspond to higher order combinations:
 
-  \item {\bf Equal bin size} plot (3d): fluct and vspine, provided the first two variables are uniform.
+\begin{itemize}
+  \item {\bf Stacked} bar chart (2d). 1 hbar and 1 vspine.
+  \item {\bf Nested} bar chart (2d).  2 hbars. \citep{peltier:2009}
+
+  \item {\bf Equal bin size} plot (3d): fluct and vspine, conditioned on the first two variables \citep{hofmann:2000}.
 
@@
   \item {\bf Squarified treemap} (nd): n tiles. \citep{bruls:1999}
+  \item {\bf Generalised treemaps} (nd): any plot ending with a tile \citep{vliegen:2006}.
 
@@
 
-These techniques are also closely related to dimensional stacking \citep{leblanc:1990}
-
 \section{Modelling}
 \label{sec:models}
+
+This framework leads to a combinatorial explosion of possibilities. A 4d pmf, f(a, b, c, d) can be factorised in five different ways:
+
+\begin{itemize}
+  \item $f(a, b, c, d) = f(a | b, c, d) f(b | c, d) f(c, d)$
+  \item $f(a, b, c, d) = f(a | b, c, d) f(b | c, d) f(c | d) f(d)$
+  \item $f(a, b, c, d) = f(a | b, c, d) f(b, c | d) f(d)$
+  \item $f(a, b, c, d) = f(a, b | c, d) f(c, d)$
+  \item $f(a, b, c, d) = f(a, b | c, d) f(c | d) f(d)$
+\end{itemize}
+
+\noindent There are 24 possible ways of ordering the variables in the pdf, and each 1d pmf can be displayed in five different ways.
+For a 4d problem this gives $5 * 24 * (5^3 + 5^4 + 5^2 + 1 + 5^2) = 96120$ plots!
+How can we use statistical models to guide us towards an appropriate choice?
+
+% Connection to models: \citep{theus:1999,hofmann:2001}
 
@@
 
-\subsection{Missing values}
+\subsection{Missing values and zeros}
 \label{sub:missing_values}
+
+Crossed vs. nested
 
@@
 Example exploring joint and conditional distributions of age and year.
+
+If an equal bin sized plot is used, and value is mapped to colour, then we get  dimensional stacking \citep{leblanc:1990}
 

### 2010-05-28 product-graphics.tex
Proof reading

@@
 \label{sec:primitives}
+
+% See "Hierarchical Pixel Bar Charts" for constraints
 
@@
 
-To construct plots for higher-dimensional tables, we need a way to decompose them into 1d and 2d components.
-To do we use some very basic properties of probability.
+To construct plots of higher-dimensional tables, we need a way to decompose them into 1d and 2d components.
+To do so we use some very basic properties of probability.
 For example, let $f(x, y, z)$ be a 3d pmf.
-We can write this pmf as product of marginals and conditionals:
+We can write this pmf as product of 1d and 2d marginals and conditionals:
 
@@
 
-\noindent Each of these decompositions is made up of products of one or two-d pmfs that can visualised used one of the graphical primitives described previously.
-Note that while $f(x, y, z)$ and $f(z, y, x)$ represent the same pmf, the factorisation will be different because you will get different marginal tables.
+\noindent Since each component of the product is only 1 or 2d, we can use one of the primitives described above.
+For example, Figure \ref{fig:recursive} shows two ways to display the joint distribution of sex and happiness as broken down into the marginal distribution of happiness, and the condition distribution of sex, given happiness.
+
+\begin{figure}[htbp]
+  \centering
+    \includegraphics[width=0.33\linewidth]{part-comb-1}%
+    \includegraphics[width=0.33\linewidth]{part-comb-2}%
+    \includegraphics[width=0.33\linewidth]{part-comb-3}
+  \caption{Conditional on marital status, are men or women happier?
+This figure shows the construction of $\sim happy + sex + marital$ from (left to right) vspines by marital status, vspines by marital status and sex, and hspines by marital status, sex and happiness.
+For all levels of marital status, men are slightly less happy.}
+  \label{fig:recursive}
+\end{figure}
+
+% Note that while $f(x, y, z)$ and $f(z, y, x)$ represent the same pmf, the factorisation will be different because you will get different marginal tables.
 
@@
 
-\subsection{Examples}
-\label{sub:examples}
-
-Figure \ref{fig:recursive} illustrates this process with a graphic that explores whether, conditional on marital status, there is a difference in happiness between men and women.
-
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=0.33\linewidth]{part-comb-1}%
-    \includegraphics[width=0.33\linewidth]{part-comb-2}%
-    \includegraphics[width=0.33\linewidth]{part-comb-3}
-  \caption{Conditional on marital status, are men or women happier?
-This figure shows the construction of $\sim happy + sex + marital$ from (left to right) vspines by marital status, vspines by marital status and sex, and hspines by marital status, sex and happiness.
-For all levels of marital status, men are slightly less happy.}
-  \label{fig:recursive}
-\end{figure}
-
 Different combinations of partitions reveal different features of the data.
@@
 
-Conditioning is also an important tool, because it allows one to remove relationships that are well-known or uninteresting.
+Conditioning is also an important tool by itself, because it allows one to remove relationships that are well-known or uninteresting.
 Figure~\ref{fig:part-cond} uses a fluct and a vspine to explore the relationship between happiness, health and financial status.
@@
 
-\subsection{Existing plot types}
+\section{Existing plot types}
 
 You may have noticed that the previous plots resemble area plots that have previously be named. The one-d primitives already have their own names:
-
-% Other papers to cite:
-%
-%
 
@@
   \item {\bf Column chart} (1d). 1 vbar.
-  \item {\bf Spineplot} (1d). 1 vspine.
-  \item {\bf Fluctuation} diagram (2d): fluct.
+  \item {\bf Spineplot} (1d). 1 spine.
+  \item {\bf Fluctuation} diagram (2d): 1 fluct.
 \end{itemize}
@@
   \item {\bf Stacked} bar chart (2d). 1 hbar and 1 vspine.
+
   \item {\bf Nested} bar chart (2d).  2 hbars. \citep{peltier:2009}
@@
 
-  \item {\bf Mosaic} plot (nd).  Alternating hspines and vspines.  \citep{hartigan:1984,hartigan:1981,friendly:1994,hofmann:2003}
+  \item {\bf Mosaic} plot (nd).  Alternating hspines and vspines.  \citep{hartigan:1981,friendly:1994,hofmann:2003}
+
   \item {\bf Double-decker} plot (nd).  $n-1$ hspines and 1 vspine. \citep{hofmann:2001}
+
   \item {\bf Treemap} (nd): n spines. \citep{shneiderman:1992}
+
   \item {\bf Squarified treemap} (nd): n tiles. \citep{bruls:1999}
+
   \item {\bf Generalised treemaps} (nd): any plot ending with a tile \citep{vliegen:2006}.
@@
 
-Trellis graphics \citep{becker:1996}, as known as lattice or facetted or
-conditioned graphics are another related display. They uses categorical
-variables to generate multiple panels, each containing a plot of the subset of
-the data. Trellised area plots also fall into our framework and can be
-created by conditioning on the trellising variables.
+Trellis graphics \citep{becker:1996}, also known as lattice, facetted and conditioned graphics, are another related display.
+They use categorical variables to generate multiple panels, each containing a plot of the subset of the data.
+Trellised area plots also fall into our framework and can be created by conditioning on the trellising variables.
 

### 2011-03-07 prodplots-infovis.tex
Paper in infovis format

--- 2010-05-28 product-graphics.tex
@@
-\documentclass[letterpaper,oneside]{scrartcl}
-\usepackage{fullpage}
+% zip -r graphical-inference.zip *.pdf *.tex images references.bib *.cls
+
+% 8 pages, including reference 
+% 
+% Frame as design paper:
+% 
+%   * explain the target problem
+%   * enough background that the reader can pass judgement about whether your
+%     solution is good
+%   * you should present results that back up the claim that your approach is
+%     better than others
+%
+% Supplemental material to create plots
+\documentclass[journal]{vgtc}
+\usepackage{mathptmx}
+\usepackage{graphicx}
+\graphicspath{{plots/}}
+\usepackage{times}
+\usepackage{booktabs}
 \usepackage[utf8]{inputenc}
-\usepackage[pdftex]{graphicx}
+\usepackage[square,sort&compress]{natbib}
+
 \usepackage{tikz}
-\DeclareGraphicsExtensions{.png,.pdf}
-\graphicspath{{plots/}}
-\usepackage{hyperref}
-\usepackage{booktabs}
-\usepackage{url}
-\usepackage[round,sectionbib]{natbib}
-\bibliographystyle{abbrvnat}
-\usepackage[small]{caption}
-\usepackage[small]{titlesec}
-\renewcommand\familydefault{bch}
-
 \definecolor{female}{HTML}{EF8A62}
@@
 
-\title{Product graphics}
-\author{Hadley Wickham and Heike Hofmann}
+\onlineid{0}
+\vgtccategory{Research}
+\vgtcinsertpkg
+
+\title{Product plots}
+\author{Hadley Wickham, and Heike Hofmann}
+\authorfooter{
+  \item Hadley Wickham is an Assistant Professor of Statistics at Rice University, Email: hadley@rice.edu.
+  \item Heike Hofmann is an Associate Professor of Statistics at Iowa State University.
+}
+\shortauthortitle{Wickham \MakeLowercase{\textit{et al.}}: Product plots}
+
+\abstract{
+
+}
+\keywords{Statistics, joint distribution, conditional distribution, treemap, bar chart, mosaic plot}
+\CCScatlist{ % not used in journal version
+  \category{H.5.2}{Information Interfaces and Presentation}{ User Interfaces --- Graphical user interfaces (GUI), Interaction styles, Screen design, Evaluation/methodology}
+  \CCScat{I.6.8}{Computing Methodologies}%
+  {Simulation and Modeling}{Visual Simulation};
+}
+
+% \teaser{
+% \centering
+% \includegraphics[width=16cm]{tx-cancer}
+% \caption{One of these plots doesn't belong.
+These six plots show choropleth maps of cancer deaths in Texas, where darker colors = more deaths.
+Can you spot which of the six plots is made from a real dataset and not simulated under the null hypothesis of spatial independence?
+If so, you've provided formal statistical evidence that deaths from cancer have spatial dependence.
+See Section~\ref{sec:solutions} for the answer.}
+% \label{fig:teaser}
+% }
+
+\renewcommand{\manuscriptnotetxt}{\itshape Manuscript received 31 March 2011; accepted XX XXXX XXXX; posted online XX XXXX XXXX; mailed on XX XXXX XXXX.
+
+For information on obtaining reprints of this article, please send email to: tvcg@computer.org.}
+
 \begin{document}
+\firstsection{Introduction}
 \maketitle
-
-\section{Introduction}
-
-% JASA methods
-% Infovis?
-
-% Aims of this paper:
-%
-%  * emphasise the connection between area graphics and factorising
-%    probabilities
-%
-%  * pull all area plots on a common footing - provide a grammar for
-%    categorical graphics
-
 Area plots, like bar charts, mosaic plots and treemaps, are the graphical equivalent of contingency tables, visualising a discrete distribution by mapping proportion to area.
@@
 
-\begin{table}[htb]
+\begin{table*}[htb]
   \begin{center}
@@
   \label{tbl:happy}
-\end{table}
+\end{table*}
 
@@
 
-\section{Modelling}
-\label{sec:models}
-
-This framework leads to a combinatorial explosion of possibilities. A 4d pmf, f(a, b, c, d) can be factorised in five different ways:
-
-\begin{itemize}
-  \item $f(a, b, c, d) = f(a | b, c, d) f(b | c, d) f(c, d)$
-  \item $f(a, b, c, d) = f(a | b, c, d) f(b | c, d) f(c | d) f(d)$
-  \item $f(a, b, c, d) = f(a | b, c, d) f(b, c | d) f(d)$
-  \item $f(a, b, c, d) = f(a, b | c, d) f(c, d)$
-  \item $f(a, b, c, d) = f(a, b | c, d) f(c | d) f(d)$
-\end{itemize}
-
-\noindent There are 24 possible ways of ordering the variables in the pdf, and each 1d pmf can be displayed in five different ways.
-For a 4d problem this gives $5 * 24 * (5^3 + 5^4 + 5^2 + 1 + 5^2) = 96120$ plots!
-How can we use statistical models to guide us towards an appropriate choice?
-
-% Connection to models: \citep{theus:1999,hofmann:2001}
-
 \section{Variations and extensions}
@@
 
+\section{Interaction}
+\label{sec:interaction}
+
 \section{Conclusions}
 
-% bibtool -x product-graphics -o references.bib
+This framework leads to a combinatorial explosion of possibilities. A 4d pmf, f(a, b, c, d) can be factorised in five different ways:
+
+\begin{itemize}
+  \item $f(a, b, c, d) = f(a | b, c, d) f(b | c, d) f(c, d)$
+  \item $f(a, b, c, d) = f(a | b, c, d) f(b | c, d) f(c | d) f(d)$
+  \item $f(a, b, c, d) = f(a | b, c, d) f(b, c | d) f(d)$
+  \item $f(a, b, c, d) = f(a, b | c, d) f(c, d)$
+  \item $f(a, b, c, d) = f(a, b | c, d) f(c | d) f(d)$
+\end{itemize}
+
+\noindent There are 24 possible ways of ordering the variables in the pdf, and each 1d pmf can be displayed in five different ways.
+For a 4d problem this gives $5 * 24 * (5^3 + 5^4 + 5^2 + 1 + 5^2) = 96120$ plots!
+In future work, we hope to connect product plots to statistical models to help guide the choice of appropriate graphic.
+
+% \acknowledgments{ The authors wish to thank Robert Kosara for helpful discussion.
+This work was partly supported by the National Science Foundation grant DMS0706949.
+Graphics produced with ggplot2 \citep{me:ggplot2}.}
+
+% bibtool -x graphical-inference -x coordination -o references.bib
+\bibliographystyle{ieeetr}
 \bibliography{references}
+
 \end{document}

### 2011-03-11 prodplots-infovis.tex
Working on paper

@@
 
-\abstract{
+\abstract{We propose a new framework for visualising counts or proportions, including existing graphics such as bar charts, mosaic plots, treemaps, equal area plots and fluctuation diagrams.
+We call our framework product plots, alluding to the computation of area as a product of height and width, and the statistical concept of generating a joint distribution from the product of conditional and marginal distributions.}
 
-}
 \keywords{Statistics, joint distribution, conditional distribution, treemap, bar chart, mosaic plot}
@@
 \maketitle
-Area plots, like bar charts, mosaic plots and treemaps, are the graphical equivalent of contingency tables, visualising a discrete distribution by mapping proportion to area.
-This paper presents a unifying framework for area plots, showing how many existing plots are interrelated, and developing a set of tools for creating new plots.
-In particular, we create a graphical factorisation of area plots by analogy to the mathematical factorisation of joint distributions into products of marginals and conditionals.
 
-The paper proceeds as follows.
-Section~\ref{sec:primitives} described the simple one- and two-d graphical primitives that define all area plots.
-Section~\ref{sec:combination} provides a mathematical framework that shows how these simple primitives can be combined to display higher-d distributions.
-Section~\ref{sec:variations} extends these techniques for other types of data, and shows variations based on different low-d graphical primitives.
+Tables of counts or proportions are an extremely common form of data.
+Correspondingly, there are many visualisations to display them.
+In this paper, we develop a framework that encompasses many existing visualisations, from bar charts to treemaps to pie charts, showing how graphics that previously seemed unrelated in fact share a deep underlying connection, and providing a systematic foundation for the developing new types of count visualisations \citep{cox:1978}.
 
-To illustrate these ideas, we use the same data for all examples.
-It's a small sample of variables related to happiness from the general social survey ({\sc gss}).
+Our framework focusses on area charts, where the area of a graphical element is proportional to the underlying count.
+We call our framework product plots in allusion to two products: the product of width and height to compute area, and the product of conditional and marginal distributions to produce joint distributions.
+A key development of our framework is the inverse operation: the factorisation of area plots.
+This allows easy extension from low-dimensional graphical primitives to high-dimensional data sets.
+
+Section~\ref{sec:primitives} we motivate the three constraints that give rise to the fundamental graphical component of productplots: the rectangle.
+The rectangle gives rises to three fundamental 1d atoms (bar, spine, tile) and one fundamental 2d atom (the fluct).
+These constraints will later be relaxed in Section~\ref{sec:variations} to extend our framework to include even more visualisations including histograms, pie charts, cascaded treemaps and weighted plots.
+
+Section~\ref{sec:combination} provides the mathematical framework that extends the simple 1d and 2d atoms to display data of any dimensionality, by showing just as any high dimensional joint distribution can be represented as a product of low-dimensional conditional and marginal distributions, any area plot can be decomposed into products of simple low-d primitives.
+Using this framework, Section~\ref{sec:existing} shows how 13 existing named graphics are just special cases of this general pattern.
+
+To illustrate these ideas, we will use the same data for all examples.
+The data is a small sample of variables related to happiness from the general social survey ({\sc gss}) \citep{davis:2008}.
 The {\sc gss} is a yearly cross-sectional survey of Americans, run from 1976.
-We combine data for 25 years to to give gives 51\,020 observations, and of the over 5\,000 variables, we select eight related to happiness, as described in Table~\ref{tbl:happy}.
-The data is included in the supplemental materials.
+We combine data for 25 years to gives 51\,020 observations, and of the over 5\,000 variables, we select eight related to happiness, as described in Table~\ref{tbl:happy}.
 
@@
 
-% See "Hierarchical Pixel Bar Charts" for constraints
+It is useful to consider the constraints on our visualisation to determine the best solution to the problem \citep{keim:2002}.
+All area plots correspond to tilings of the plane.
+We could consider partitions of higher-d spaces (e.g.
+3d or 4d), but given that we have to project these down to 2d for viewing on paper or screen, there is little disadvantage to working directly in 2d.
 
-All area plots correspond to tilings of the plane.
-We could also consider partitions of higher-d spaces (e.g.
-3d or 4d), but given that we have to project these down to 2d for viewing, there is little disadvantage to working directly in 2d.
-
-To derive the graphical primitives we start with three constraints:
+To derive our graphical primitives we start with three constraints:
 
 \begin{itemize}
-  \item Area must be {\bf proportional} to value. Because the total area for a graphic is usually constrained, an area plot partitions this total, displaying proportions rather than counts.
 
-  \item Partitions must be {\bf rectangular}.
-It is easier to compare the areas of simple shape (i.e.
-rectangle vs polygon), and even easier when the length of one border is fixed, so we only need compare the length of the other \citep{cleveland:1984}.
-Rectangular partitions are also recursive in the sense that we can always tile a rectangle with smaller rectangles.
+  \item Area must be {\bf proportional} to value. Because the total area for a graphic is usually constrained, an area plot partitions the total, displaying proportions rather than counts.
+
+  \item Partitions must be {\bf rectangular}, because rectangles are easier to compare than arbitrary polygons \citep{cleveland:1984,cleveland:1986,heer:2010}, and are recursive in the sense that we can always tile a rectangle with smaller rectangles.
 This is not true for most shapes.
 
-  \item Partitions must be {\bf disjoint}.
-To be able to see the complete area of each rectangle they must be non-overlapping.
-This does not imply that the tiling be space-filling: generally, space filling methods use space more efficiently, but make areas harder to compare.
+  \item Partitions must be {\bf disjoint}. To be able to see the complete area, each rectangle must be non-overlapping. This does not imply that the tiling must be space-filling.
 
@@
 
-\noindent Each of these three constraints can be profitably relaxed, as detailed in Section~\ref{sec:variations}.
+\noindent Each of three constraints can be profitably relaxed, as detailed in Section~\ref{sec:variations}.
 
-The following sections describe the partitions generated by these constraints.
-Section~\ref{sub:part-1d} discusses 1d partitions, bars, spines and tiles, and  Section~\ref{sub:part-2d} discusses the 2d partition, the fluct.
+These constraints give rise to four graphical primitives.
+Section~\ref{sub:part-1d} describes the partitions of 1d data, bars, spines and tiles, and Section~\ref{sub:part-2d} describes the one partition of 2d data, the fluct.
 
@@
 
-There are three 1d primitives, as shown in Figure~\ref{fig:part-1d}:
+1d primitives display one dimensional proportion data, i.e.\ proportions broken down by a single variable. There are three 1d primitives, as shown in Figure~\ref{fig:part-1d}.
 
@@
   \item {\bf spine}s: width is proportional to value, height occupies full range.
-Splines are space filling.
-Splines can be arranged horizontally (``hspine''), vertically (``vspine''), or can automatically pick their orientation (``spine'') by splitting the largest dimension.
-The appearance of bars and splines is identical for uniformly distributed data.
+Spines are space filling and can be arranged horizontally (``hspine''), vertically (``vspine''), or can automatically pick their orientation (``spine'') by splitting the largest dimension.
+The name spine is evocative of books sitting on a library shelf \citep{hummel:1996}.
+Bars and spines are identical when visualising uniformly distributed data.
 
@@
 The perceptual task associated with bar charts is the simplest, but they do not occupy all the area.
-Spines and tiles are harder to read \citep{heer:2010}, but occupy the complete space and work better recursively.
+Spines and tiles are harder to read \citep{heer:2010}, but occupy the complete space and so work better recursively.
 
@@
 
+We can create two primitives by combining these one primitives, as shown in Section~\ref{sec:combination}, but there is one 2d primitive that does not arise in this way.
+
 \subsection{2d primitives}
@@
 
-We are currently aware of one 2d primitive, the {\bf fluct}, which has both height and width proportional to the square root of the proportion \citep{hofmann:2000}.
+2d primitives display two dimensional proportion data, i.e.\ proportions broken down by two variables.
+We are currently aware of one primitive for 2d data, the {\bf fluct}, derived from the fluctuation diagram \citep{hofmann:2000}.
+The fluct has height and width proportional to the square root of the proportion.
 Each rectangle is arranged on a regular grid formed by the levels of the two variables, allowing comparisons in two directions.
-A special case of the fluct is the equal bin size plot which occurs when the two variables are jointly uniformly distributed, usually as a result of the conditioning described in the next section.
-The equal bin size plot is useful to visualise missings.
+
+A special case of the fluct is the equal bin size plot \citep{hofmann:2000} which occurs when the two variables are jointly uniformly distributed, usually as a result of the conditioning described in the following section.
+The equal bin size plot is particularly useful as a way visualise missing combinations.
 
@@
 
-To construct plots of higher-dimensional tables, we need a way to decompose them into 1d and 2d components.
-To do so we use some very basic properties of probability.
-For example, let $f(x, y, z)$ be a 3d pmf.
-We can write this pmf as product of 1d and 2d marginals and conditionals:
+To construct plots of higher-dimensional data sets, we need a way to decompose them into 1d and 2d components.
+Some statistical vocabulary is useful.
+For example, let $f(x, y, z)$ be a 3d probability mass function.
+A probability mass function is just a way of describing the table of counts or proportions as a function: the three parameters are three indices, saying where to look up the number.
+We can write this pmf as product of 1d and 2d marginals and conditionals.
+
+\begin{itemize}
+
+  \item Joint distribution:
+
+  \item Marginal distribution is just a name for a lower dimensional projection of a distribution - in a different context, these could be joint distributions.
+For example, in the three d case, there are three 2d marginal distributions: $f(x, y)$, $f(x, z)$ and $f(y, z)$, and three 1d marginal distributions, $f(x)$, $f(y)$ and $f(z)$.
+
+  \item A conditional distribution is a slice of a higher dimension distribution, fixing one or more variables at a particular value.
+
+\end{itemize}
+
+\begin{figure*}[ht]
+\centering
+\begin{tabular}{rrr}
+  \toprule
+  & male & female \\ 
+  \midrule
+  not too happy & 0.05 & 0.07 \\ 
+  pretty happy & 0.25 & 0.31 \\ 
+  very happy & 0.14 & 0.18 \\ 
+  \bottomrule
+\end{tabular}\hspace{2em}%
+\begin{tabular}{rrrr}
+  \toprule
+  & male & female & {\bf margin} \\ 
+  \midrule
+  not too happy & 0.43 & 0.57 & \hspace{2em} 0.12 \\ 
+  pretty happy & 0.45 & 0.55 & 0.56 \\ 
+  very happy & 0.43 & 0.57 & 0.32 \\ 
+  \bottomrule
+\end{tabular}\hspace{1em}%
+\begin{tabular}{rrr}
+  \toprule
+  & male & female \\ 
+  \midrule
+  not too happy & 0.12 & 0.12 \\ 
+  pretty happy & 0.57 & 0.55 \\ 
+  very happy & 0.31 & 0.32 \\[1em] 
+  {\bf margin} & 0.44 & 0.56 \\ 
+  \bottomrule
+\end{tabular}
+  \caption{The distribution of happiness and sex, displayed in three equivalent ways.
+(Left) Joint distribution.
+Overall table sums to one.
+(Middle) Conditional distribution of sex given happiness and marginal distribution of happiness.
+(Right) Conditional distribution of happiness given sex and marginal distribution of sex.}
+\label{fig:fluct}
+
+\end{figure*}
+
+This is easy to see in two dimensions.  Figure XXX shows three representations of the
+
+An important property of the definition of conditional probability is that it means you can write a joint distribution as a product of conditional and marginals:
 
@@
 \end{itemize}
+
+This means that if we can translate this multiplication into some kind of graphical product, we can display any nD data set using low-D primitives.
 
@@
 \section{Existing plot types}
+\label{sec:existing}
 
@@
   \item {\bf Column chart} (1d). 1 vbar.
-  \item {\bf Spineplot} (1d). 1 spine.
+  \item {\bf Spineplot} (1d). 1 spine \citep{hummel:1996}.
   \item {\bf Fluctuation} diagram (2d): 1 fluct.
@@
 
-This gives rise to the histogram and spinogram, continuous analogues of the bar chart and spine chart.
+This gives rise to the histogram and spinogram \citep{hummel:1996}, continuous analogues of the bar chart and spine chart.
 A long standing tradition is that no gaps are displayed between adjacent rectangles when used for continuous data.
@@
 
+Approach is also taken by \citep{keim:2002}, where a minimum bar width of 10 pixels is used.
+
 Other non-linear transformations may also be useful.
@@
 
-A pie chart is a popular method of displaying proportions, but is not a rectangular partition.
-However, it turns out that there is a simple relationship between product plots and pie charts: an pie chart is an hspine drawn in polar coordinates with the x coordinate mapped to angle and the y coordinate to radius.
-Many other displays that use circular arcs turn out to be special cases of product plots, drawn in polar coordinates:
+A pie chart is a popular method of displaying proportions, but it is not a rectangular partition and so does not seem to fall in the framework of this paper.
+However, there is a simple relationship between product plots and pie charts: an pie chart is an hspine drawn in polar coordinates with the x coordinate mapped to angle and the y coordinate to radius.
+
+Many other circular displays turn out to be special cases of product plots drawn in polar coordinates.
+Generally, the y axis (mapped to radius) must be square-root transformed to ensure that that counts stay proportional to areas.
+Stasko's radial displays \citep{stasko:2000} and fan-lenses \citep{lou:2007} deliberately do not do this in order to emphasise the outer levels.
+
+We have identified the following radial plots as polar transformations of product plots:
 
 \begin{itemize}
-  \item Fourfold displays \citep{friendly:1995}, wind rose, sector graphic: hbar.
+  \item The wind rose \citep{lalanne:1843}, fourfold displays \citep{friendly:1995}, and sector graphic use the hbar.
+Nightingales's coxcomb \citep{nightingale:1857} is very similar, but the slices overlap and so violate the constraint of disjoint area.
 
-  \item Bullseye chart: hspine.
+  \item The bullseye chart uses a hspine.
 
-  \item Doughnut plot: vspine + hspine.
+  \item The doughnut plot, as popularised by Excel, uses a vspine and a hspine.
 
-  \item Circular bar chart or racetrack plot: vbar.
+  \item The circular bar chart or racetrack plot, uses a vbar.
 
-  \item Infoslices \citep{andrews:1998}, nested vbar. But they only use half of the disk, and are specialised for highly nested data.
+  \item Infoslices \citep{andrews:1998} use a nested vbar. But they only use half of the polar plane, and are specialised for highly nested data.
 
-  \item Nightingales's coxcomb, is very similar to vspine / hbar in polar coordinates, except that the pie slices are overlapping and so violates the constraint of disjoint area.
 \end{itemize}
 
-\noindent Eight variants are displayed in Figure~\ref{fig:polar}.
-
-To make this work more generally, the y-variable (mapped to radius) should be square-root transformed to ensure that that counts stay proportional to areas.
-The radial displays of \citet{stasko:2000} and fanlens of \citet{lou:2007} deliberately choose not to do this to highlight the outer levels.
+\noindent Eight polar variants are displayed in Figure~\ref{fig:polar}.
+Many of these are familiar and already have names.
+They are all of dubious utility because of the increased difficulty when comparing non-rectangular areas.
 
@@
 
-A number of non-rectangular treemaps have been proposed, such as the circular treemaps of \citet{wetzel:2008}, the space-filling curves of \citet{wattenberg:2005} and the voronoi treemaps of \citet{balzer:2005}.
-However, because it is so much more difficult to compare the areas of arbitrary polygons, these approaches tend to be attractive rather than useful.
+A number of non-rectangular treemaps have been proposed, such as circular treemaps \citep{wetzel:2008}, space-filling curves \citep{wattenberg:2005} and voronoi treemaps \citep{balzer:2005}.
+The perceptual task associated with these visualisations is the comparison of the area of arbitrary polygons, which is much more difficult than comparing the area of rectangles \citep{cleveland:1986}.
+This makes these approaches attractive rather than useful.
+
+\section{Aesthetics}
+\label{sec:aesthetics}
+
+\subsection{Labelling} 
+\label{sub:labelling}
+
+\subsection{Offsets}
+\label{sub:offsets}
 
@@
 
-% bibtool -x graphical-inference -x coordination -o references.bib
+% bibtool -x prodplots-infovis -o references.bib
 \bibliographystyle{ieeetr}

### 2011-03-11 tables.r

@@ -1,0 +1,15 @@
+l(productplots)
+library(xtable)
+
+joint <- with(happy, prop.table(table(happy, sex)))
+xtable(joint)
+
+cond_happy <- with(happy, prop.table(table(happy, sex), 1))
+marg_happy <- t(t(with(happy, prop.table(table(happy)))))
+happy <- cbind(cond_happy, marg_happy)
+xtable(happy)
+
+cond_sex <- with(happy, prop.table(table(happy, sex), 2))
+marg_sex <- t(with(happy, prop.table(table(sex))))
+sex <- rbind(cond_sex, marg_sex)
+xtable(sex)

### 2011-03-13 prodplots-infovis.tex
Improving paper further

@@
     {\sf degree} & highest education & lt high school, high school, junior college, bachelor, graduate \\
-    {\sf finrela} & relative financial status & far above average, above average, average, below average, far below average \\
+    {\sf finrela} & relative financial status & far above, above average, average, below average, far below \\
     {\sf happy} & happiness & \key{very-happy} very happy, \key{pretty-happy} pretty happy, \key{not-too-happy} not too happy \\
@@
 
-It is useful to consider the constraints on our visualisation to determine the best solution to the problem \citep{keim:2002}.
 All area plots correspond to tilings of the plane.
@@
 3d or 4d), but given that we have to project these down to 2d for viewing on paper or screen, there is little disadvantage to working directly in 2d.
-
-To derive our graphical primitives we start with three constraints:
+There are many possible ways to tile the plane.
+To weed these down to a manageable number, we identify three constraints \citep{keim:2002} important for visualising counts:
 
@@
 
-  \item Area must be {\bf proportional} to value. Because the total area for a graphic is usually constrained, an area plot partitions the total, displaying proportions rather than counts.
+  \item Area must be {\bf proportional} to count.
+Because the total area for a graphic is usually constrained, and area plot partitions the total, area plots display proportions, not counts.
+From here on, without loss of generality, we will refer to visualisation of proportions, rather than counts.
 
@@
 
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.33\linewidth]{part-hbar}%
+  \includegraphics[width=0.33\linewidth]{part-hspine}%
+  \includegraphics[width=0.33\linewidth]{part-treemap}
+
+  \caption{1d partitions showing the distribution of happiness.  From left to right: bars, spines and tiles.}
+  \label{fig:part-1d}
+\end{figure}
+
 \begin{itemize}
+   \setlength{\itemsep}{0em}
 
@@
 
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=0.33\linewidth]{part-hbar}%
-  \includegraphics[width=0.33\linewidth]{part-hspine}%
-  \includegraphics[width=0.33\linewidth]{part-treemap}
-
-  \caption{1d partitions showing the distribution of happiness.  From left to right: bars, spines and tiles.}
-  \label{fig:part-1d}
-\end{figure}
-
-We can create two primitives by combining these one primitives, as shown in Section~\ref{sec:combination}, but there is one 2d primitive that does not arise in this way.
+We can create 2d primitives by combining these 1d primitives, as shown in Section~\ref{sec:combination}, but there is one 2d primitive that does not arise in this way.
 
@@
 
-A special case of the fluct is the equal bin size plot \citep{hofmann:2000} which occurs when the two variables are jointly uniformly distributed, usually as a result of the conditioning described in the following section.
-The equal bin size plot is particularly useful as a way visualise missing combinations.
-
 \begin{figure}[htbp]
   \centering
-    \includegraphics[width=0.5\linewidth]{part-fluct}%
-    \includegraphics[width=0.5\linewidth]{part-fluct-cond}
-  \caption{(Left) A fluct showing distribution of age (in decades) and marital status.
+    \includegraphics[width=0.9 \linewidth]{part-fluct}
+    \includegraphics[width=0.9 \linewidth]{part-fluct-cond}
+  \caption{(Left) A fluctuation diagram showing distribution of age (in decades) and marital status.
 (Right) Equal bin-size plot showing health status and survey year.
@@
 
-\section{Factorising area plots}
+A special case of the fluct is the equal bin size plot \citep{hofmann:2000} which occurs when the two variables are jointly uniformly distributed, usually as a result of the conditioning described in the following section.
+The equal bin size plot is particularly useful as a way of visualising missing combinations.
+Figure~\ref{fig:fluct} illustrates these two types of graphics.
+
+\section{Probability and plot products}
 \label{sec:combination}
@@
 Some statistical vocabulary is useful.
-For example, let $f(x, y, z)$ be a 3d probability mass function.
-A probability mass function is just a way of describing the table of counts or proportions as a function: the three parameters are three indices, saying where to look up the number.
-We can write this pmf as product of 1d and 2d marginals and conditionals.
+One way of describing the input data is as a {\bf probability mass function}, or {\sc pmf}.
+A {\sc pmf} is a function with n inputs, each indexing one dimension with an integer, and gives the probability of each combination of inputs.
+A {\sc pmf} has two restrictions: every value must be greater than or equal to zero, and all value must sum to one.
 
-\begin{itemize}
+\subsection{Products of probabilities}
 
-  \item Joint distribution:
+Figure~\ref{fig:2d-table} shows three ways to represent the the 2d table of proportions of sex and happiness.
+The top table displays the {\bf joint distribution}, and allows us to answer questions of the form ``what proportion of all people are male and very happy?'' The middle two tables displays two {\bf conditional distributions}: the distribution of sex given happiness, and happiness given sex.
+These correspond to restriction the row or column sums to equal one, and support questions such as ``what proportion of pretty happy people are female?'' or ``what proportion of males are not too happy?''.
+Finally, the last table display the two {\bf marginal distributions} of sex and happiness.
+These allow us to answer questions like ``what proportion of respondents were male'' or ``what proportion of respondents were very happy?''.
 
-  \item Marginal distribution is just a name for a lower dimensional projection of a distribution - in a different context, these could be joint distributions.
-For example, in the three d case, there are three 2d marginal distributions: $f(x, y)$, $f(x, z)$ and $f(y, z)$, and three 1d marginal distributions, $f(x)$, $f(y)$ and $f(z)$.
-
-  \item A conditional distribution is a slice of a higher dimension distribution, fixing one or more variables at a particular value.
-
-\end{itemize}
-
-\begin{figure*}[ht]
-\centering
+\begin{figure}[ht]
   \begin{tabular}{rrr}
@@
   \bottomrule
-\end{tabular}\hspace{2em}%
-\begin{tabular}{rrrr}
+  \end{tabular}
+  \\[1em]
+
+  \begin{tabular}{rrr}
   \toprule
-  & male & female & {\bf margin} \\ 
+  & male & female \\
   \midrule
-  not too happy & 0.43 & 0.57 & \hspace{2em} 0.12 \\ 
-  pretty happy & 0.45 & 0.55 & 0.56 \\ 
-  very happy & 0.43 & 0.57 & 0.32 \\ 
+  not too happy & 0.43 & 0.57 \\
+  pretty happy & 0.45 & 0.55 \\ 
+  very happy & 0.43 & 0.57 \\
   \bottomrule
@@
   pretty happy & 0.57 & 0.55 \\ 
-  very happy & 0.31 & 0.32 \\[1em] 
-  {\bf margin} & 0.44 & 0.56 \\ 
+  very happy & 0.31 & 0.32 \\
   \bottomrule
   \end{tabular}
+  \\[1em]
+
+  \begin{tabular}{rrrr}
+  \toprule
+  & male & female & \\ 
+  \midrule
+  not too happy & & & 0.12 \\ 
+  pretty happy & & & 0.56 \\ 
+  very happy & & & 0.30 \\
+  & 0.44 & 0.56 \\ 
+  \bottomrule
+  \end{tabular}
+
   \caption{The distribution of happiness and sex, displayed in three equivalent ways.
@@
 (Right) Conditional distribution of happiness given sex and marginal distribution of sex.}
-\label{fig:fluct}
+  \label{fig:2d-table}
+\end{figure}
 
-\end{figure*}
+Formally, the conditional distribution is written as $f(x | y)$ and is equal to $f(x, y) / f(y)$.
+This definition illustrates that given a conditional distribution and a marginal distribution, we can always find the joint distribution: $f(x, y) = f(x | y) f(y)$ \footnotemark.
+This definition extends in a straightforward way to higher dimensions.
+For example, a 3d pmf can be written as the product of 2d and 1d conditional and marginal distributions as follows:
 
-This is easy to see in two dimensions.  Figure XXX shows three representations of the
-
-An important property of the definition of conditional probability is that it means you can write a joint distribution as a product of conditional and marginals:
+\footnotetext{Because the parameters of a pmf are identifying, statisticians often fail to distinguish the different functions.
+To be precise, the above statement should be written: $f_{X, Y}(x, y) = f_{(X|Y=y)}(x | y) f_Y(y)$}
 
 \begin{itemize}
+  \setlength{\itemsep}{0em}
   \item $f(x, y, z) = f(z) f(x, y | z)$
@@
 
-This means that if we can translate this multiplication into some kind of graphical product, we can display any nD data set using low-D primitives.
+This means that we can turn any high-dimensional {\sc pmf} into a product of low-dimensional conditional and marginal {\sc pmf}s products.
+It remains to find an graphical analog of this multiplication.
 
-\noindent Since each component of the product is only 1 or 2d, we can use one of the primitives described above.
-For example, Figure \ref{fig:recursive} shows two ways to display the joint distribution of sex and happiness as broken down into the marginal distribution of happiness, and the condition distribution of sex, given happiness.
+\subsection{Area is a product of height and width}
 
-\begin{figure}[htbp]
-  \centering
-    \includegraphics[width=0.33\linewidth]{part-comb-1}%
-    \includegraphics[width=0.33\linewidth]{part-comb-2}%
-    \includegraphics[width=0.33\linewidth]{part-comb-3}
-  \caption{Conditional on marital status, are men or women happier?
-This figure shows the construction of $\sim happy + sex + marital$ from (left to right) vspines by marital status, vspines by marital status and sex, and hspines by marital status, sex and happiness.
-For all levels of marital status, men are slightly less happy.}
-  \label{fig:recursive}
-\end{figure}
-
-% Note that while $f(x, y, z)$ and $f(z, y, x)$ represent the same pmf, the factorisation will be different because you will get different marginal tables.
-
-Writing the full pmf is tedious, so we define an algebra based on the modelling specification of \citet{wilkinson:1973}:
-
-\begin{itemize}
-  \item \verb|~ a| represents $f(a)$,
-  \item \verb|~ a + b| represents $f(a, b)$
-  \item \verb!~ a | b! represents $f(a | b)$
-  \item \verb!~ . | a! is a special case, $f(a) = \frac{1}{\#A}$.
-\end{itemize}
-
-Figure~\ref{fig:fact-simple} uses the distribution of sex and happiness to illustrate these ideas.
-From left to right, we have plots of \verb|~ happy|, \verb!~ happy | sex!
-and the product \verb|~ happy + sex|.
-Observe that the heights and widths of the rectangles are multiplied in the same way as the components of the pmf.
+We connect probability products to our rectangular primitives by noting that areas are also products, products of height and width.
+It's easiest to show this with a picture: Figure~\ref{fig:fact-simple} shows how our simple one primitives combine to get two familiar plots: the mosaic plot and the stacked bar chart.
+From left to right, we have plots of $f(happy)$, $f(happy | sex)$ and the product $f(happy, sex)$.
+The heights and widths of the rectangles are multiplied in the same way as the components of the {\sc pmf}.
 
@@
 
+Figure \ref{fig:recursive} shows the steps along by which a 3d dataset is displayed.
+We first display $f(martial status)$, then $f(marital sex, sex) = f(sex | marital status) f(marital status)$, and finally $f(marital sex, sex, happiness) = f(happiness | sex, marital status) f(sex | marital status) f(marital status)$.
+
+\begin{figure}[htbp]
+  \centering
+    \includegraphics[width=0.33\linewidth]{part-comb-1}%
+    \includegraphics[width=0.33\linewidth]{part-comb-2}%
+    \includegraphics[width=0.33\linewidth]{part-comb-3}
+  \caption{Conditional on marital status, are men or women happier?
+This figure shows the construction of $f(happy, sex, marital)$ from (left to right) vspines by marital status, vspines by marital status and sex, and hspines by marital status, sex and happiness.
+For all levels of marital status, men are slightly less happy.}
+  \label{fig:recursive}
+\end{figure}
+
+% Note that while $f(x, y, z)$ and $f(z, y, x)$ represent the same pmf, the factorisation will be different because you will get different marginal tables.
+% 
+% Writing the full pmf is tedious, so we define an algebra based on the Wilkinson's modelling specification \citep{wilkinson:1973}:
+% 
+% \begin{itemize}
+%   \setlength{\itemsep}{0em}
+%   \item \verb|~ a| represents $f(a)$,
+%   \item \verb|~ a + b| represents $f(a, b)$
+%   \item \verb!~ a | b! represents $f(a | b)$
+%   \item \verb!~ . | a! is a special case, $f(a) = \frac{1}{\#A}$.
+% \end{itemize}
+%
+
 % To visualise higher-dimensional pmfs, we recursively partition using the previously described primitives.
@@
 Figure~\ref{fig:marital} shows two ways to do this.
-The left plot shows $\sim marital + age$ with a vspine nested in a hspine, and the right plot $\sim age + martial$ with a hbar nested in a vspine.
+The left plot shows $f(marital, age)$ with a vspine nested in a hspine, and the right plot $f(age, marital)$ with a hbar nested in a vspine.
 These displays show the same data but are arranged in such a way to make different comparisons easier.
+The second plot, with bars nested inside spines, also illustrates important feature of non-space filling tilings: the relationship between proportion and area is only constant within a level.
+The bars are scaled to be as high as possible, without overflowing the bounding area.
 
@@
 However, it is difficult to see how happiness varies within these conditions because the perceptual task is comparing areas.
-Conditioning on financial status and health produces the plot on the right (an equal bin size plot) and makes it easier to see the conditional distribution of happiness, because the perceptual task is simpler: comparing length on a common non-aligned scales.
-Here we can see people tend to be happier the healthier and richer they are.
+Conditioning on financial status and health produces the plot on the right (an equal bin size plot) and makes it easier to see the conditional distribution of happiness given sex and health, because the perceptual task is simpler.
+Depending on the comparison we are most interested in, we can make it easier to compare across wealth for fixed health, or health for fixed wealth.
 
-\begin{figure}[htbp]
+For this data, for a fixed income level, happiness always increases as health increases.
+The same is not true for fixed health: those with poor or fair health are actually less happy if they are far above average, compared to just above average.
+
+\begin{figure*}[htbp]
   \centering
-    \includegraphics[width=0.5\linewidth]{part-fluctuation}%
-    \includegraphics[width=0.5\linewidth]{part-equal-area}
-  \caption{(Left) $\sim happy + health + finrela$, partitioned with a vspine and fluct.
-(Right) $\sim happy \,|\, health + finrela$.
+    \includegraphics[width=0.33\linewidth]{part-fluctuation}%
+    \includegraphics[width=0.33\linewidth]{part-equal-area}%
+    \includegraphics[width=0.33\linewidth]{part-equal-area-2}
+  \caption{(Left) $f(happy, health, finrela) = f(happy | health, finrela) \times f(health, finrela)  $, partitioned with a vspine and fluct.
+(Middle) $f(happy | health, finrela)$.
 We can no longer see the joint distribution of health and financial status, but it is much easier to see the conditional distribution of happiness.
 Healthier and richer people are happier: maybe money does buy happiness?
+(Right) $f(happy | health, finrela)$ partitioned with a fluct and {\bf h}spine.
 (\key{not-too-happy} Not too happy, \key{pretty-happy} pretty happy, \key{very-happy} very happy)}
   \label{fig:part-cond}
-\end{figure}
+\end{figure*}
 
@@
 
-You may have noticed that the previous plots resemble area plots that have previously be named. The one-d primitives already have their own names:
+Many existing plots fall into this framework. The primitives already have their own names:
 
 \begin{itemize}
+  \setlength{\itemsep}{0em}
   \item {\bf Bar chart} (1d). 1 hbar.
   \item {\bf Column chart} (1d). 1 vbar.
-  \item {\bf Spineplot} (1d). 1 spine \citep{hummel:1996}.
+  \item {\bf Spineplot} (1d). 1 spine.
   \item {\bf Fluctuation} diagram (2d): 1 fluct.
@@
 \begin{itemize}
+  \setlength{\itemsep}{0em}
+
   \item {\bf Stacked} bar chart (2d). 1 hbar and 1 vspine.
 
-  \item {\bf Nested} bar chart (2d).  2 hbars. \citep{peltier:2009}
+  \item {\bf Nested} bar chart \citep{peltier:2009} (2d).  2 hbars.
 
-  \item {\bf Equal bin size} plot (3d): fluct and vspine, conditioned on the first two variables \citep{hofmann:2000}.
+  \item {\bf Equal bin size} \citep{hofmann:2000} plot (3d): fluct and vspine, conditioned on the first two variables.
 
-  \item {\bf Mosaic} plot (nd).  Alternating hspines and vspines.  \citep{hartigan:1981,friendly:1994,hofmann:2003}
+  \item {\bf Mosaic} plot \citep{hartigan:1981,friendly:1994,hofmann:2003} (nd). Alternating hspines and vspines.
 
-  \item {\bf Double-decker} plot (nd).  $n-1$ hspines and 1 vspine. \citep{hofmann:2001}
+  \item {\bf Double-decker} plot \citep{hofmann:2001} (nd). $n-1$ hspines and 1 vspine.
 
-  \item {\bf Treemap} (nd): n spines. \citep{shneiderman:1992}
+  \item {\bf Treemap} \citep{shneiderman:1992} (nd): n spines.
 
-  \item {\bf Squarified treemap} (nd): n tiles. \citep{bruls:1999}
+  \item {\bf Squarified treemap} \citep{bruls:1999} (nd): n tiles.
 
-  \item {\bf Generalised treemaps} (nd): any plot ending with a tile \citep{vliegen:2006}.
+  \item {\bf Generalised treemaps} \citep{vliegen:2006} (nd): any plot ending with a tile.
 
@@
 On top of this basic framework, it is useful to consider a few variations and extensions.
-Three are based on variations of the underlying data (missing data, weighted data, and continuous data), and three are based on relaxing the display constraints (area not proportional to weight, non-disjoint partitions, non-rectangular partitions).
-
-\subsection{Missing values and zeros}
-\label{sub:missing_values}
-
-Crossed vs. nested
+Instead of count of discrete levels, we can plot weighted data or continuous data, or we can relax the display constraints to allow displays where area is not proportional to weight, partitions are non-disjoint or non-rectangular.
 
@@
 
-We have assumed that the probabilities represent counts, but without loss of generality, we can use any set of non-negative, additive weights.
+We have assumed that the proportions represent counts, but without loss of generality, we can use any set of non-negative, additive weights.
 For example, in the happy dataset, the {\sf wtssall} variable gives analytic survey weights.
+These are use to account for oversampling of black respondents in certain years, and to reduce the effect of non-response by other demographics.
 Figure~\ref{fig:weighting} shows the difference between the weighted and unweighted distributions of age and sex.
-The distribution is subtly different.
-\citep{unwin:2007} describes the applications of weighted plots in more detail.
-
-To express that we are displaying weights, rather than counts, we extend the formula syntax by putting the weighting variable on the left-hand side: $wtssall \sim sex + age$ vs. $ \sim sex + age$.
+The distribution is subtly different, and suggests that we don't need to worry about weights for this plot.
+In other datasets, weights can be useful to move from numbers of counties to numbers of people, or to areas, or to other relevant quantities.
+Some examples of weighted data graphics can be found in \citep{unwin:1998,unwin:2003aa,unwin:2006}.
 
@@
 
-These techniques can also be used with continuous data, if the data is binned to create a categorical variable.
-There are many different possible choices of binning algorithms, but we will focus on two:
-
-\begin{itemize}
-  \item Intervals of equal width.
-  \item Intervals containing equal numbers of points.
-\end{itemize}
-
-This gives rise to the histogram and spinogram \citep{hummel:1996}, continuous analogues of the bar chart and spine chart.
+The framework can be trivially extended to work with continuous data: just first bin continuous variables.
+There are many different to ways to create intervals for continuous data, but two are most important \citep{denby:2009}: intervals of equal width, and intervals containing equal numbers of points.
+This extension allows the product plot framework to also describe {\bf histograms} and {\bf spinograms} \citep{hummel:1996}, continuous analogues of the bar and charts.
 A long standing tradition is that no gaps are displayed between adjacent rectangles when used for continuous data.
@@
   \centering
-  \caption{Area plots of originally continuous data}
+  \includegraphics[width=0.5\linewidth]{cont-hbar}%
+  \includegraphics[width=0.5\linewidth]{cont-hspine}
+
+  \caption{The distribution of happyness with age. (Left) hbar + vspine: an extension of the histogram. (Right) hspine + vspine: an extesion of the spinogram.}
   \label{fig:cont-examples}
@@
 
-Example exploring joint and conditional distributions of age and year.
-
-If an equal bin sized plot is used, and value is mapped to colour, then we get  dimensional stacking \citep{leblanc:1990}
-
-% An alternative approach is use density estimates (or conditional density estimates) and plot with continuous area plots.
-As in: Hofmann, H., Theus, M.
-(2005), Interactive graphics for visualizing conditional distributions, Unpublished Manuscript.
+With one more extension, displaying the final probability with colour, not area, this allows us to describe {\bf dimensional stacking} \citep{leblanc:1990} as a plot of conditioned, binned continuous data, with proportion shown with colour.
 
@@
 It can be useful to violate the constraint that area should be proportional to weight to distinguish between zeros and very small values.
-A zero weight should have zero area, but giving it positive area can be useful, so you can actually see it!
+A zero weight should have zero area, but giving it positive area can be useful that so you can actually see it!
 In general, it's useful to constrain all areas to be above a certain minimal perceptible size.
 Areas which are constrained in such a way should have a visual flag (such as a difference colour) so ensure that the reader knows that the relationship between area and value has been violated.
+This type of non-linear mapping has been implemented in {\sc manet} \citep{unwin:1996} and hierarchical pixel bar charts \citep{keim:2002}.
 
-On the other end of the spectrum, it can be useful to constrain put a cap on the largest possible values \citetext{Antony Unwin, priv.
+On the other end of the spectrum, it can be useful to constrain put a cap on the size of largest values to get censored zooming \citetext{Antony Unwin, priv.
 comm.}.
-By controlling this cap it is possible to zoom in on the smallest values.
-- called censored zooming by Antony and Martin.
-
-Approach is also taken by \citep{keim:2002}, where a minimum bar width of 10 pixels is used.
+Controlling this cap makes it possible to focus on the smallest values.
 
@@
 For example, we could display the square root of the weights, to stabilise the variance of the areas.
-Tukey applied this technique to histograms to create rootograms.
-
-If values span multiple orders of magnitude it may be useful to map log value to area.
+Tukey applied this technique to histograms to create rootograms \citep{wainer:1974,tukey:1977}: histograms where the y-axis has a square root scale.
 
@@
 The cascaded treemaps of \citet{lu:2008} is an idea that illustrates how the violation of containment can be productive.
-In the cascaded treemap, each level is slightly offset from the one above to create a pseudo-3d perspective.
+In the {\bf cascaded treemap}, each level is slightly offset from the one above to create a pseudo-3d perspective.
 This makes it easier to see all the levels of the hierarchy, not just the lowest level.
@@
 
-Many other circular displays turn out to be special cases of product plots drawn in polar coordinates.
+Many other circular displays turn out to be special cases of product plots drawn in polar coordinates \citep{draper:2009}.
 Generally, the y axis (mapped to radius) must be square-root transformed to ensure that that counts stay proportional to areas.
@@
 \begin{itemize}
-  \item The wind rose \citep{lalanne:1843}, fourfold displays \citep{friendly:1995}, and sector graphic use the hbar.
+  \setlength{\itemsep}{0em}
+
+  \item {\bf Wind rose} (aka sector graphic) \citep{lalanne:1843} and {\bf fourfold displays} \citep{friendly:1995} (2d): 1 hbar, and 1 vspine.
 Nightingales's coxcomb \citep{nightingale:1857} is very similar, but the slices overlap and so violate the constraint of disjoint area.
 
-  \item The bullseye chart uses a hspine.
+  \item {\bf Concentric pie chart} (aka bullseye chart) (1d): 1 hspine.
 
-  \item The doughnut plot, as popularised by Excel, uses a vspine and a hspine.
+  \item {\bf Doughnut plot} (2d): 1 hspine, and 1 vspine.
 
-  \item The circular bar chart or racetrack plot, uses a vbar.
+  \item {\bf Racetrack plot} (aka circular bar chart) (1d): 1 vbar.
 
-  \item Infoslices \citep{andrews:1998} use a nested vbar. But they only use half of the polar plane, and are specialised for highly nested data.
+  \item {\bf Infoslices} \citep{andrews:1998} (nd): n vbars. But they only use half of the polar plane, and are specialised for highly nested data.
 
@@
 
-\section{Aesthetics}
-\label{sec:aesthetics}
-
-\subsection{Labelling} 
-\label{sub:labelling}
-
-\subsection{Offsets}
-\label{sub:offsets}
-
-\section{Interaction}
-\label{sec:interaction}
+% \section{Aesthetics}
+% \label{sec:aesthetics}
+% 
+% \subsection{Labelling} 
+% \label{sub:labelling}
+% 
+% \subsection{Offsets}
+% \label{sub:offsets}
+% 
+% 
+% \section{Interaction}
+% \label{sec:interaction}
 
@@
 \noindent There are 24 possible ways of ordering the variables in the pdf, and each 1d pmf can be displayed in five different ways.
-For a 4d problem this gives $5 * 24 * (5^3 + 5^4 + 5^2 + 1 + 5^2) = 96120$ plots!
+For a 4d problem this gives $5 * 24 * (5^3 + 5^4 + 5^2 + 1 + 5^2) = 96,120$ plots!
 In future work, we hope to connect product plots to statistical models to help guide the choice of appropriate graphic.
+
+We are also interested in exploring the automated construction of a graphic to answer specific questions, and for given a plot, suggesting which type of questions it is most suitable to answer.
 

### 2011-03-13 plots-partition.r

@@ -26,11 +26,11 @@ ggsave("plots/part-treemap.pdf", width = 4, height = 4)
 # age & marital status
 happy$decade <- cut(happy$age, seq(10, 90, by = 10))
 prodplot(happy, ~  decade + marital, "fluct", na.rm = T) + fill_marital
-ggsave("plots/part-fluct.pdf", width = 6, height = 6)
+ggsave("plots/part-fluct.pdf", width = 12, height = 6)
 
 # year & health
 prodplot(happy, ~ . | year + health, "fluct", na.rm = TRUE) + fill_health
-ggsave("plots/part-fluct-cond.pdf", width = 6, height = 6)
+ggsave("plots/part-fluct-cond.pdf", width = 12, height = 6)
 
 
 # nd partitions --------------------------------------------------------------
@@ -64,10 +64,16 @@ prodplot(happy, ~ happy + marital, c("vspine", "hspine")) +
   aes(fill = marital)
 
 # fluctuation & equal bin size
-prodplot(happy, ~ happy + finrela + health, na.rm = TRUE,
-  set_offset(c("vspine", "fluct"), c(0, 0.05))) + fill_happy
+prodplot(happy, ~ happy + finrela + health, na.rm = TRUE, 
+  set_offset(c("vspine", "fluct"), c(0, 0.05)), subset = .(level == 2)) +
+  fill_happy
 ggsave("plots/part-fluctuation.pdf", width = 6, height = 6)
 
 prodplot(happy, ~ happy | finrela + health, na.rm = TRUE,
   set_offset(c("vspine", "fluct"), c(0, 0.05))) + fill_happy
 ggsave("plots/part-equal-area.pdf", width = 6, height = 6)
+
+prodplot(happy, ~ happy | finrela + health, na.rm = TRUE,
+  set_offset(c("hspine", "fluct"), c(0, 0.05)), subset = .(level == 2)) +
+  fill_happy
+ggsave("plots/part-equal-area-2.pdf", width = 6, height = 6)

### 2011-03-13 plots-variations.r

@@ -1,31 +1,38 @@
 library(productplots)
 source("colour.r")
 
+happy$age_i <- cut_interval(happy$age, length = 10)
+happy$year_i <- cut_interval(happy$year, length = 5)
+
 # Weighting ------------------------------------------------------------------
 
 happy2 <- transform(happy, .wt = wtssall)
+
 
 # prodplot(happy, ~ sex, "hbar") + fill_sex
 # prodplot(happy2, ~ sex, "hbar") + fill_sex
 # prodplot(happy, ~ finrela + health, "fluct")
 # prodplot(happy2, ~ finrela + health, "fluct")
 
-prodplot(happy, ~ sex  + age, set_offset(c("vspine", "hbar")), na.rm = TRUE) +
-  fill_sex
+prodplot(happy, ~ sex + age_i, set_offset(c("vspine", "hbar")), subset = .(level == 2), na.rm = TRUE) + fill_sex
 ggsave("plots/wt-count.pdf", width = 6, height = 4)
 
-prodplot(happy2, ~ sex + age, set_offset(c("vspine", "hbar")), na.rm = TRUE) +
- fill_sex
+prodplot(happy2, ~ sex + age_i, set_offset(c("vspine", "hbar")), subset = .(level == 2), na.rm = TRUE) + fill_sex
 ggsave("plots/wt-wtssall.pdf", width = 6, height = 4)
 
 
 # Continuous data ------------------------------------------------------------
 
-happy$age_n <- cut_number(happy$age, 10)
-happy$age_i <- cut_interval(happy$age, 10)
-
-prodplot(happy, ~ age_n, set_offset("hbar"))
 prodplot(happy, ~ age_i, set_offset("hbar"))
 
-prodplot(happy, ~ age_n, set_offset("hspine"))
-prodplot(happy, ~ age_i, set_offset("hspine"))
+prodplot(happy, ~ happy + age_i, c(vspine, set_offset("hspine")), 
+  subset = .(level == 2)) + fill_happy
+ggsave("plots/cont-hspine.pdf", width = 6, height = 4)
+
+prodplot(happy, ~ happy + age_i, c(vspine, set_offset("hbar")), 
+  subset = .(level == 2)) + fill_happy
+ggsave("plots/cont-hbar.pdf", width = 6, height = 4)
+
+prodplot(happy, ~ age_i + year_i, "fluct")
+prodplot(happy, ~ age_i | year_i, c("hspine", "vspine"), 
+  subset = .(level == 2))

### 2011-03-15 plots-polar.r
Improve polar plots

@@ -1,44 +1,44 @@
 # Polar coordinates
+l(productplots)
 source("colour.r")
 
 shade <- c(fill_sex, list(
   opts(legend.position = "none", 
     axis.text.x = theme_blank(),
-    axis.text.y = theme_blank()
+    axis.text.y = theme_blank(),
+    axis.ticks = theme_blank()
   ), 
   xlab(NULL),
-  ylab(NULL),
-  scale_y_sqrt()
+  ylab(NULL)
 ))
+
+polar_x <- list(coord_polar(), aes(xmin = sqrt(l), xmax = sqrt(r)))
+polar_y <- list(coord_polar("y"), aes(ymin = sqrt(b), ymax = sqrt(t)))
 
 prodplot(happy, ~ sex + happy, set_offset(c("vspine", "hspine")), na.rm = T) + shade
 ggsave("plots/hs-vs-cartesian.pdf", width = 4, height = 4)
-last_plot() + coord_polar()
+last_plot() + polar_x
 ggsave("plots/hs-vs-polar.pdf", width = 4, height = 4)
-last_plot() + coord_polar("y")
+last_plot() + polar_y
 ggsave("plots/hs-vs-polar-2.pdf", width = 4, height = 4)
-
-prodplot(happy, ~ sex + happy, set_offset(c("hspine", "vspine")), na.rm = T) + shade
-last_plot() + coord_polar()
-
 
 prodplot(happy, ~ sex + happy, set_offset(c("hbar", "hbar")), na.rm = T) + shade
 ggsave("plots/hb-vb-cartesian.pdf", width = 4, height = 4)
-last_plot() + coord_polar()
+last_plot() + polar_x
 ggsave("plots/hb-vb-polar.pdf", width = 4, height = 4)
-last_plot() + coord_polar("y")
+last_plot() + polar_y
 ggsave("plots/hb-vb-polar-2.pdf", width = 4, height = 4)
 
 prodplot(happy, ~ sex + happy, divider = set_offset(c("hspine", "hspine")), na.rm = T) + shade
 ggsave("plots/hs-hs-cartesian.pdf", width = 4, height = 4)
-last_plot() + coord_polar()
+last_plot() + polar_x
 ggsave("plots/hs-hs-polar.pdf", width = 4, height = 4)
-last_plot() + coord_polar("y")
+last_plot() + polar_y
 ggsave("plots/hs-hs-polar-2.pdf", width = 4, height = 4)
 
[…]

### 2011-03-15 prodplots-infovis.tex
Improve polar plots
More tweaks to paper

@@
 \definecolor{very-happy}{HTML}{91CF60}
+\definecolor{not-too-happy}{HTML}{FC8D59}
+\definecolor{pretty-happy}{HTML}{FFFFBF}
+\definecolor{very-happy}{HTML}{91CF60}
+\definecolor{married}{HTML}{F1EEF6} 
+\definecolor{never-married}{HTML}{BDC9E1} 
+\definecolor{divorced}{HTML}{74A9CF} 
+\definecolor{widowed}{HTML}{2B8CBE} 
+\definecolor{separated}{HTML}{045A8D} 
 \definecolor{grey30}{HTML}{4D4D4D}
@@
 \begin{table*}[htb]
-  \begin{center}
+  \centering
   \begin{tabular}{llp{9cm}}
@@
     {\sf health} & health & excellent, good, fair, poor \\
+    {\sf marital} & marital status & \key{married} married, \key{never-married} never married, \key{divorced} divorced, \key{widowed} widowed, \key{separated} separated \\
     {\sf sex} & sex & \key{female} female, \key{male} male\\
@@
   \end{tabular}
-  \end{center}
   \caption{Description of sample data. Common plot colours are shown next to respective levels.}
@@
 
-  \caption{The distribution of happyness with age. (Left) hbar + vspine: an extension of the histogram. (Right) hspine + vspine: an extesion of the spinogram.}
+  \caption{The distribution of happiness with age. (Left) hbar + vspine: an extension of the histogram. (Right) hspine + vspine: an extesion of the spinogram.}
   \label{fig:cont-examples}
@@
 
-With one more extension, displaying the final probability with colour, not area, this allows us to describe {\bf dimensional stacking} \citep{leblanc:1990} as a plot of conditioned, binned continuous data, with proportion shown with colour.
+With one more extension, displaying the innermost proportion with colour (instead of area), this allows us to describe {\bf dimensional stacking} \citep{leblanc:1990} as a plot of binned, conditioned continuous data, with proportion shown by colour.
 
@@
 
-It can be useful to violate the constraint that area should be proportional to weight to distinguish between zeros and very small values.
-A zero weight should have zero area, but giving it positive area can be useful that so you can actually see it!
+It can be useful to violate the constraint that area is proportional to value to distinguish between zeros and very small values.
+A zero weight should have zero area, but giving it positive area can be useful that so you can see it!
 In general, it's useful to constrain all areas to be above a certain minimal perceptible size.
-Areas which are constrained in such a way should have a visual flag (such as a difference colour) so ensure that the reader knows that the relationship between area and value has been violated.
-This type of non-linear mapping has been implemented in {\sc manet} \citep{unwin:1996} and hierarchical pixel bar charts \citep{keim:2002}.
+Areas which are constrained in such a way should have a visual flag (such as a different colour) to ensure that the reader knows that the relationship between area and value has been violated.
+This type of non-linear mapping has been implemented in {\sc manet} \citep{unwin:1996} and in hierarchical pixel bar charts \citep{keim:2002}.
 
-On the other end of the spectrum, it can be useful to constrain put a cap on the size of largest values to get censored zooming \citetext{Antony Unwin, priv.
+On the other end of the spectrum, it can be useful to constrain the size of largest values to get {\bf censored zooming} \citetext{Antony Unwin, priv.
 comm.}.
-Controlling this cap makes it possible to focus on the smallest values.
+Controlling this cap makes it possible to focus on small values.
 
 Other non-linear transformations may also be useful.
-For example, we could display the square root of the weights, to stabilise the variance of the areas.
-Tukey applied this technique to histograms to create rootograms \citep{wainer:1974,tukey:1977}: histograms where the y-axis has a square root scale.
+For example, we could take square roots to stabilise the variance of the areas.
+Tukey applied this technique to histograms to create rootograms \citep{wainer:1974,tukey:1971}: histograms where the y-axis has a square root scale.
 
@@
 
-The cascaded treemaps of \citet{lu:2008} is an idea that illustrates how the violation of containment can be productive.
+Cascaded treemaps \citep{lu:2008} illustrate how the violation of containment can be productive.
 In the {\bf cascaded treemap}, each level is slightly offset from the one above to create a pseudo-3d perspective.
@@
 Figure~\ref{fig:cascade} shows an example of how cascading can help illuminate the structure of a complex mosaic plot.
+This technique is probably more effective when implemented interactively.
 
@@
 However, there is a simple relationship between product plots and pie charts: an pie chart is an hspine drawn in polar coordinates with the x coordinate mapped to angle and the y coordinate to radius.
-
 Many other circular displays turn out to be special cases of product plots drawn in polar coordinates \citep{draper:2009}.
 Generally, the y axis (mapped to radius) must be square-root transformed to ensure that that counts stay proportional to areas.
-Stasko's radial displays \citep{stasko:2000} and fan-lenses \citep{lou:2007} deliberately do not do this in order to emphasise the outer levels.
+Fan-lenses \citep{lou:2007} and Stasko's radial displays \citep{stasko:2000} deliberately do not do this in order to emphasise the outer levels.
 
@@
 
-This framework leads to a combinatorial explosion of possibilities. A 4d pmf, f(a, b, c, d) can be factorised in five different ways:
+This framework leads to a combinatorial explosion of possibilities. A 4d {\sc pmf}, $f(a, b, c, d)$, can be factorised in five different ways:
 
 \begin{itemize}
+  \setlength{\itemsep}{0em}
+
+  \item $f(a, b, c, d) = f(a | b, c, d) f(b | c, d) f(c | d) f(d)$
   \item $f(a, b, c, d) = f(a | b, c, d) f(b | c, d) f(c, d)$
-  \item $f(a, b, c, d) = f(a | b, c, d) f(b | c, d) f(c | d) f(d)$
   \item $f(a, b, c, d) = f(a | b, c, d) f(b, c | d) f(d)$
+  \item $f(a, b, c, d) = f(a, b | c, d) f(c | d) f(d)$
   \item $f(a, b, c, d) = f(a, b | c, d) f(c, d)$
-  \item $f(a, b, c, d) = f(a, b | c, d) f(c | d) f(d)$
 \end{itemize}
 
-\noindent There are 24 possible ways of ordering the variables in the pdf, and each 1d pmf can be displayed in five different ways.
-For a 4d problem this gives $5 * 24 * (5^3 + 5^4 + 5^2 + 1 + 5^2) = 96,120$ plots!
-In future work, we hope to connect product plots to statistical models to help guide the choice of appropriate graphic.
+\noindent There are 24 possible ways of ordering the variables in the {\sc pmf}, 5 ways of displaying a 1d {\sc pmf}, and 1 way of displaying a 2d {\sc pmf}, leading to a possible $24 * (5^4 + 5^2 + 5^2 + 5^2 + 1) = 16,824$ plots, before we even consider conditioning!
+Well thought out interaction can make it easy to navigate this space, but it In future work, we hope to connect product plots to statistical models to help guide the choice of appropriate graphic.
 
@@
 
-% bibtool -x prodplots-infovis -o references.bib
+% bibtool -x prodplots-infovis > references.bib
 \bibliographystyle{ieeetr}

### 2011-03-15 plots-banner.r
Add a teaser banner

@@ -1,0 +1,22 @@
+l(productplots)
+source("colour.r")
+
+no_axes <- list(scale_x_continuous(breaks = NA), scale_y_continuous(breaks = NA))
+
+prodplot(happy, ~ happy + marital, c("vspine", "hspine"), na.rm = TRUE) + fill_marital + no_axes
+ggsave("plots/banner-1.pdf", width = 4, height = 4)
+
+prodplot(happy, ~ marital + happy, c("tile", "tile"), na.rm = TRUE) + fill_marital + no_axes
+ggsave("plots/banner-2.pdf", width = 4, height = 4)
+
+prodplot(happy, ~ marital + happy, c("hbar", "vbar"), na.rm = TRUE) + fill_marital + no_axes
+ggsave("plots/banner-3.pdf", width = 4, height = 4)
+
+prodplot(happy, ~ marital + happy, c("hbar", "hbar"), na.rm = TRUE) + fill_marital + no_axes
+ggsave("plots/banner-4.pdf", width = 4, height = 4)
+
+prodplot(happy, ~ marital + happy, c("hspine", "vbar"), na.rm = TRUE) + fill_marital + no_axes
+ggsave("plots/banner-5.pdf", width = 4, height = 4)
+
+prodplot(happy, ~ marital + happy, c("fluct"), na.rm = TRUE) + fill_marital + no_axes
+ggsave("plots/banner-6.pdf", width = 4, height = 4)

### 2011-03-15 plots-cont.r

@@ -1,4 +1,4 @@
-source("../load.r")
+l(productplots)
 
 qplot(age, year, data = h)
 ggsave("plots/age-year.pdf", width = 4, height = 4)

### 2011-03-15 plots-banner.r
Tweaking banner

@@ -9,13 +9,13 @@ ggsave("plots/banner-1.pdf", width = 4, height = 4)
 prodplot(happy, ~ marital + happy, c("tile", "tile"), na.rm = TRUE) + fill_marital + no_axes
 ggsave("plots/banner-2.pdf", width = 4, height = 4)
 
-prodplot(happy, ~ marital + happy, c("hbar", "vbar"), na.rm = TRUE) + fill_marital + no_axes
+prodplot(happy, ~ marital + happy, c("hbar", "vbar"), na.rm = TRUE, levels = NA) + fill_marital + no_axes
 ggsave("plots/banner-3.pdf", width = 4, height = 4)
 
-prodplot(happy, ~ marital + happy, c("hbar", "hbar"), na.rm = TRUE) + fill_marital + no_axes
+prodplot(happy, ~ marital + happy, c("hbar", "hspine"), na.rm = TRUE, levels = NA) + fill_marital + no_axes
 ggsave("plots/banner-4.pdf", width = 4, height = 4)
 
-prodplot(happy, ~ marital + happy, c("hspine", "vbar"), na.rm = TRUE) + fill_marital + no_axes
+prodplot(happy, ~ marital + happy, c("hspine", "vbar"), na.rm = TRUE, levels = NA) + fill_marital + no_axes
 ggsave("plots/banner-5.pdf", width = 4, height = 4)
 
 prodplot(happy, ~ marital + happy, c("fluct"), na.rm = TRUE) + fill_marital + no_axes

### 2011-03-30 prodplots-infovis.tex
Add a teaser banner
Tweaking banner
Lots more work

@@
+% TO DO:
+% * look up Heike's kronecker product notation
+% * complete display considerations
+% * complete conclusions and plans for future work
+
 % zip -r graphical-inference.zip *.pdf *.tex images references.bib *.cls
@@
 
+\newcommand{\includegraphicsinline}[2]{\begin{minipage}{#1}\includegraphics[width=\linewidth]{#2}\end{minipage}}
+
 \onlineid{0}
@@
 
-\abstract{We propose a new framework for visualising counts or proportions, including existing graphics such as bar charts, mosaic plots, treemaps, equal area plots and fluctuation diagrams.
-We call our framework product plots, alluding to the computation of area as a product of height and width, and the statistical concept of generating a joint distribution from the product of conditional and marginal distributions.}
+\abstract{
+
+We propose a new framework for visualising tables counts, proportions or probabilities.
+We call our framework product plots, alluding to the computation of area as a product of height and width, and the statistical concept of generating a joint distribution from the product of conditional and marginal distributions.
+The framework, with extensions, is sufficient to encompass over 20 visualisations previously described in fields of statistical graphics and infovis, including bar charts, mosaic plots, treemaps, equal area plots and fluctuation diagrams.
+
+}
 
@@
 
-% \teaser{
-% \centering
-% \includegraphics[width=16cm]{tx-cancer}
-% \caption{One of these plots doesn't belong.
-These six plots show choropleth maps of cancer deaths in Texas, where darker colors = more deaths.
-Can you spot which of the six plots is made from a real dataset and not simulated under the null hypothesis of spatial independence?
-If so, you've provided formal statistical evidence that deaths from cancer have spatial dependence.
-See Section~\ref{sec:solutions} for the answer.}
-% \label{fig:teaser}
-% }
+\teaser{
+\centering
+\includegraphics[width=0.16\linewidth]{banner-1}%
+\includegraphics[width=0.16\linewidth]{banner-2}%
+\includegraphics[width=0.16\linewidth]{banner-5}%
+\includegraphics[width=0.16\linewidth]{banner-6}%
+\includegraphics[width=0.16\linewidth]{banner-3}%
+\includegraphics[width=0.16\linewidth]{banner-4}
+\caption{A selection of plots encompassed by the product plots framework.
+All display the same data (the distribution of happiness and marital status), but each supports a different question.
+From left to right: mosaic plot, treemap, stacked  bar chart, fluctuation diagram, and two new plots that don't have names.
+In all plots, area is proportional to probability.
+}
+\label{fig:teaser}
+}
 
@@
 
-Tables of counts or proportions are an extremely common form of data.
-Correspondingly, there are many visualisations to display them.
-In this paper, we develop a framework that encompasses many existing visualisations, from bar charts to treemaps to pie charts, showing how graphics that previously seemed unrelated in fact share a deep underlying connection, and providing a systematic foundation for the developing new types of count visualisations \citep{cox:1978}.
+Tables of counts, proportions and probabilities are an extremely common form of data, and many researchers have developed visualisations to display them.
+In this paper, we develop a framework that encompasses many existing visualisations, from bar charts to treemaps to pie charts, showing how graphics that previously seemed unrelated in fact share a deep underlying connection \citep{cox:1978}.
+Our framework makes it easier to describe and create visualisations of categorical data, shows how existing methods are related and where new methods can be developed, and will make it easier to match questions about categorical data to the visualisations that will provide revealing answers.
 
-Our framework focusses on area charts, where the area of a graphical element is proportional to the underlying count.
-We call our framework product plots in allusion to two products: the product of width and height to compute area, and the product of conditional and marginal distributions to produce joint distributions.
-A key development of our framework is the inverse operation: the factorisation of area plots.
-This allows easy extension from low-dimensional graphical primitives to high-dimensional data sets.
+Our framework focusses on area charts, where the area of a graphical element is proportional to the underlying count, proportion, or probability.
+We call our framework \emph{product plots} in allusion to two products: the product of width and height to generate area, and the product of conditional and marginal distributions to produce joint distributions.
+A key development of the products plot framework is the inverse operation: the factorisation of high-dimensional data to products of low-dimensional plots.
+This allows us to  combine simple, low-dimensional graphical primitives to display complex, high-dimensional data.
 
-Section~\ref{sec:primitives} we motivate the three constraints that give rise to the fundamental graphical component of productplots: the rectangle.
-The rectangle gives rises to three fundamental 1d atoms (bar, spine, tile) and one fundamental 2d atom (the fluct).
-These constraints will later be relaxed in Section~\ref{sec:variations} to extend our framework to include even more visualisations including histograms, pie charts, cascaded treemaps and weighted plots.
+We begin in Section~\ref{sec:primitives} by motivating the three constraints that give rise to the basic graphical component of product plots, the rectangle.
+We constrain the rectangle further to produce three 1d atoms (bar, spine, tile) and one 2d atom (the fluct).
+These constraints will later be relaxed in Section~\ref{sec:variations} to extend our framework to include more visualisations such as histograms, pie charts, cascaded treemaps and weighted plots.
 
-Section~\ref{sec:combination} provides the mathematical framework that extends the simple 1d and 2d atoms to display data of any dimensionality, by showing just as any high dimensional joint distribution can be represented as a product of low-dimensional conditional and marginal distributions, any area plot can be decomposed into products of simple low-d primitives.
-Using this framework, Section~\ref{sec:existing} shows how 13 existing named graphics are just special cases of this general pattern.
+Section~\ref{sec:combination} provides the mathematical framework that allows us to combine the 1d and 2d atoms to display data of any dimensionality.
+This framework is based on the fundamental statistical idea that high any high dimensional distribution can be factorised into a product of low-dimensional conditional and marginal distributions.
+Section~\ref{sec:existing} shows how many existing named graphics are special cases of this general pattern.
 
-To illustrate these ideas, we will use the same data for all examples.
+Section~\ref{sec:display} discusses some considerations for the display of product plots, and Section~\ref{sec:package} introduces the R package {\tt prodplot}, our reference implementation of the product plots framework.
+Finally, Section~\ref{sec:conclusion} discusses our plans for the future.
+
+To illustrate these ideas, we will use the same data in all examples.
 The data is a small sample of variables related to happiness from the general social survey ({\sc gss}) \citep{davis:2008}.
 The {\sc gss} is a yearly cross-sectional survey of Americans, run from 1976.
-We combine data for 25 years to gives 51\,020 observations, and of the over 5\,000 variables, we select eight related to happiness, as described in Table~\ref{tbl:happy}.
+We combine data for 25 years to yield 51\,020 observations, and of the over 5\,000 variables, we select nine related to happiness, as described in Table~\ref{tbl:happy}.
 
@@
   \end{tabular}
-  \caption{Description of sample data. Common plot colours are shown next to respective levels.}
+  \caption{Description of sample data. Common plot colours are shown next to respective levels.  Colours from ColorBrewer palettes \citep{brewer:1994}.}
   \label{tbl:happy}
@@
 
-All area plots correspond to tilings of the plane.
+All area plots correspond to tilings of the (2d) plane.
 We could consider partitions of higher-d spaces (e.g.
 3d or 4d), but given that we have to project these down to 2d for viewing on paper or screen, there is little disadvantage to working directly in 2d.
-There are many possible ways to tile the plane.
-To weed these down to a manageable number, we identify three constraints \citep{keim:2002} important for visualising counts:
+
+There are many possible ways to tile the plane. To weed these down to a manageable number, we identify three constraints \citep{keim:2002} important for visualising counts:
 
@@
   \item Area must be {\bf proportional} to count.
-Because the total area for a graphic is usually constrained, and area plot partitions the total, area plots display proportions, not counts.
-From here on, without loss of generality, we will refer to visualisation of proportions, rather than counts.
+This is the key constraint underlying all area plots.
+The total area for a graphic is usually constrained, which means that area plots display typically proportions or probabilities, rather than counts.
+We will use count, probability and proportion interchangeably throughout the paper.
 
-  \item Partitions must be {\bf rectangular}, because rectangles are easier to compare than arbitrary polygons \citep{cleveland:1984,cleveland:1986,heer:2010}, and are recursive in the sense that we can always tile a rectangle with smaller rectangles.
+  \item Partitions must be {\bf rectangular}, because the areas of rectangles are easier to compare than areas of other shapes \citep{cleveland:1984,cleveland:1986,heer:2010}, and are recursive, in the sense that we can always tile a rectangle with smaller rectangles.
 This is not true for most shapes.
 
-  \item Partitions must be {\bf disjoint}. To be able to see the complete area, each rectangle must be non-overlapping. This does not imply that the tiling must be space-filling.
+  \item Partitions must be {\bf disjoint}.
+To be able to see the complete area, each rectangle must be non-overlapping.
+Note that this does not imply that the tiling must be space-filling, and two of the four primitives, described next, are not.
 
@@
 
-\noindent Each of three constraints can be profitably relaxed, as detailed in Section~\ref{sec:variations}.
+\noindent Each of the three constraints can be profitably relaxed, as detailed in Section~\ref{sec:variations}.
 
 These constraints give rise to four graphical primitives.
-Section~\ref{sub:part-1d} describes the partitions of 1d data, bars, spines and tiles, and Section~\ref{sub:part-2d} describes the one partition of 2d data, the fluct.
-
-% Let $a(i)$ give the area assigned to category $i$, which has proportion $f(i)$.
-Since we are using rectangles, $a(i) = k \dot a(i) = k \dot w(i) \times h(i)$ where $w$ and $h$ represent width and height respectively.
-Certain partitions also satisfy the constraint that $h(i) = \sqrt{k} f(i)^x$ and $w(i) = \sqrt{k} f(i)^y$ where $x + y = 1$.
-This is useful because it is only necessary to inspect one border to compare magnitudes.
+Section~\ref{sub:part-1d} describes the three partitions of 1d data, bars, spines and tiles, and Section~\ref{sub:part-2d} describes the one partition of 2d data, the fluct.
 
@@
 
-1d primitives display one dimensional proportion data, i.e.\ proportions broken down by a single variable. There are three 1d primitives, as shown in Figure~\ref{fig:part-1d}.
+1d primitives display 1d \emph{data}, i.e.\ counts broken down by a single variable. There are three 1d primitives, as shown in Figure~\ref{fig:part-1d} and described below.
 
@@
 The name spine is evocative of books sitting on a library shelf \citep{hummel:1996}.
-Bars and spines are identical when visualising uniformly distributed data.
+Bars and spines are indistinguishable when the underlying data is uniformly distributed.
 
   \item {\bf tile}s: no restrictions on height or width, just tile the plane with rectangles, trying to keep the aspect ratio of each rectangle close to 1.
-This is partitioning used by the squarified treemap \citep{bruls:1999}.
+This is partitioning defined by the squarified treemap \citep{bruls:1999}.
 
@@
 Each of these three displays has different strengths and weaknesses.
-The perceptual task associated with bar charts is the simplest, but they do not occupy all the area.
-Spines and tiles are harder to read \citep{heer:2010}, but occupy the complete space and so work better recursively.
+It is easiest to compare the value associated with bars because the perceptual task is the easiest: comparing position on a common scale.
+Reading spines and tiles are harder because we must compare lengths or areas, but they occupy the complete space and so work better recursively.
 
@@
 
-2d primitives display two dimensional proportion data, i.e.\ proportions broken down by two variables.
-We are currently aware of one primitive for 2d data, the {\bf fluct}, derived from the fluctuation diagram \citep{hofmann:2000}.
-The fluct has height and width proportional to the square root of the proportion.
-Each rectangle is arranged on a regular grid formed by the levels of the two variables, allowing comparisons in two directions.
+2d primitives display 2d \emph{data}, i.e.\ a count broken down by two variables.
+We are currently aware of only one primitive for 2d data, the {\bf fluct}, derived from the fluctuation diagram \citep{hofmann:2000}.
+The fluct has height and width proportional to the square root of the count.
+Each rectangle is arranged on a regular grid formed by the levels of the two variables, allowing comparisons both vertically and horizontally.
+
+A special case of the fluct is the equal bin size plot \citep{hofmann:2000} which occurs when the two variables are jointly uniformly distributed, usually as a result of the conditioning described in the following section.
+The equal bin size plot is particularly useful as a way of visualising missing combinations.
+Figure~\ref{fig:fluct} illustrates these two types of graphics.
 
@@
 (Right) Equal bin-size plot showing health status and survey year.
-Health status was not recorded for three years.}
+Thee three empty columns show that health status was not recorded for three years.}
   \label{fig:fluct}
 \end{figure}
-
-A special case of the fluct is the equal bin size plot \citep{hofmann:2000} which occurs when the two variables are jointly uniformly distributed, usually as a result of the conditioning described in the following section.
-The equal bin size plot is particularly useful as a way of visualising missing combinations.
-Figure~\ref{fig:fluct} illustrates these two types of graphics.
 
@@
 To construct plots of higher-dimensional data sets, we need a way to decompose them into 1d and 2d components.
-Some statistical vocabulary is useful.
-One way of describing the input data is as a {\bf probability mass function}, or {\sc pmf}.
+Some statistical vocabulary is useful: one way of describing the input data is as a {\bf probability mass function}, or {\sc pmf}.
 A {\sc pmf} is a function with n inputs, each indexing one dimension with an integer, and gives the probability of each combination of inputs.
-A {\sc pmf} has two restrictions: every value must be greater than or equal to zero, and all value must sum to one.
+A {\sc pmf} has two restrictions: every value must be greater than or equal to zero, and all values must sum to one.
 
-\subsection{Products of probabilities}
+\subsection{Joint distributions are the product of marginals and conditionals}
 
@@
 The top table displays the {\bf joint distribution}, and allows us to answer questions of the form ``what proportion of all people are male and very happy?'' The middle two tables displays two {\bf conditional distributions}: the distribution of sex given happiness, and happiness given sex.
-These correspond to restriction the row or column sums to equal one, and support questions such as ``what proportion of pretty happy people are female?'' or ``what proportion of males are not too happy?''.
-Finally, the last table display the two {\bf marginal distributions} of sex and happiness.
+These correspond to restricting the row or column sums to one, and support questions such as ``what proportion of pretty happy people are female?'' or ``what proportion of males are not too happy?''.
+Finally, the last table displays the two {\bf marginal distributions} of sex and happiness.
 These allow us to answer questions like ``what proportion of respondents were male'' or ``what proportion of respondents were very happy?''.
@@
   \caption{The distribution of happiness and sex, displayed in three equivalent ways.
-(Left) Joint distribution.
+(Top) Joint distribution.
 Overall table sums to one.
 (Middle) Conditional distribution of sex given happiness and marginal distribution of happiness.
-(Right) Conditional distribution of happiness given sex and marginal distribution of sex.}
+(Bottom) Conditional distribution of happiness given sex and marginal distribution of sex.}
   \label{fig:2d-table}
@@
 
-Formally, the conditional distribution is written as $f(x | y)$ and is equal to $f(x, y) / f(y)$.
-This definition illustrates that given a conditional distribution and a marginal distribution, we can always find the joint distribution: $f(x, y) = f(x | y) f(y)$ \footnotemark.
+Formally, a conditional distribution function is written $f(x | y)$ and is equal to $f(x, y) / f(y)$.
+This definition illustrates that an important statistical fact: given a conditional distribution and a marginal distribution, we can always find the joint distribution: $f(x, y) = f(x | y)\,f(y)$ \footnotemark.
 This definition extends in a straightforward way to higher dimensions.
-For example, a 3d pmf can be written as the product of 2d and 1d conditional and marginal distributions as follows:
-
-\footnotetext{Because the parameters of a pmf are identifying, statisticians often fail to distinguish the different functions.
-To be precise, the above statement should be written: $f_{X, Y}(x, y) = f_{(X|Y=y)}(x | y) f_Y(y)$}
+For example, a 3d joint distribution can be written as the product of 2d and 1d conditional and marginal distributions in the following three ways:
 
@@
   \setlength{\itemsep}{0em}
-  \item $f(x, y, z) = f(z) f(x, y | z)$
-  \item $f(x, y, z) = f(y, z) f(x | y, z) $
-  \item $f(x, y, z) = f(z) f(y | z) f(x | y, z)$
+  \item $f(x, y, z) = f(z)\,f(x, y | z)$
+  \item $f(x, y, z) = f(y, z)\,f(x | y, z) $
+  \item $f(x, y, z) = f(z)\,f(y | z)\,f(x | y, z)$
 \end{itemize}
 
-This means that we can turn any high-dimensional {\sc pmf} into a product of low-dimensional conditional and marginal {\sc pmf}s products.
+\footnotetext{Because the parameters of a {\sc pmf} identify it, statisticians often fail to explicitly label the different functions.
+To be precise, the above statement should be written: $f_{X, Y}(x, y) = f_{X|Y=y}(x | y)\,f_Y(y)$}
+
+This means that we can turn any high-dimensional {\sc pmf} into a product of low-dimensional conditional and marginal {\sc pmf}s, a computationally trivial operation.
 It remains to find an graphical analog of this multiplication.
 
-\subsection{Area is a product of height and width}
+\subsection{Area is the product of height and width}
 
 We connect probability products to our rectangular primitives by noting that areas are also products, products of height and width.
-It's easiest to show this with a picture: Figure~\ref{fig:fact-simple} shows how our simple one primitives combine to get two familiar plots: the mosaic plot and the stacked bar chart.
+It's easiest to show this with a picture: Figure~\ref{fig:fact-simple} shows how our simple 1d primitives combine to get two familiar plots: the mosaic plot and the stacked bar chart.
 From left to right, we have plots of $f(happy)$, $f(happy | sex)$ and the product $f(happy, sex)$.
@@
   \centering
-  \includegraphics[width=0.3\linewidth]{fact-happy} $\times$ %
-  \includegraphics[width=0.3\linewidth]{fact-happy|sex} = %
-  \includegraphics[width=0.3\linewidth]{fact-happy-sex}
+  \includegraphicsinline{0.3\linewidth}{fact-happy} = %
+  \includegraphicsinline{0.3\linewidth}{fact-happy|sex} = %
+  \includegraphicsinline{0.3\linewidth}{fact-happy-sex}
 
-  \includegraphics[width=0.3\linewidth]{fact-happy-2} $\times$ %
-  \includegraphics[width=0.3\linewidth]{fact-happy|sex-2} = %
-  \includegraphics[width=0.3\linewidth]{fact-happy-sex-2}
+  \includegraphicsinline{0.3\linewidth}{fact-happy-2} $\times$ %
+  \includegraphicsinline{0.3\linewidth}{fact-happy|sex-2} = %
+  \includegraphicsinline{0.3\linewidth}{fact-happy-sex-2}
 
@@
 
-Figure \ref{fig:recursive} shows the steps along by which a 3d dataset is displayed.
-We first display $f(martial status)$, then $f(marital sex, sex) = f(sex | marital status) f(marital status)$, and finally $f(marital sex, sex, happiness) = f(happiness | sex, marital status) f(sex | marital status) f(marital status)$.
+Figure \ref{fig:recursive} shows a more complicated example: visualising a 3d distribution as a product of three low-dimensional distributions.
+We first display $f(marital)$, then $f(marital, sex) = f(sex | marital) f(marital)$, and finally $f(marital, sex, happy) = f(happy | sex, marital) f(sex | marital) f(marital)$.
+This plot uses two vspines and an hspine to produce a mosaic plot.
 
@@
   \caption{Conditional on marital status, are men or women happier?
-This figure shows the construction of $f(happy, sex, marital)$ from (left to right) vspines by marital status, vspines by marital status and sex, and hspines by marital status, sex and happiness.
+This figure shows the construction of $f(happy, sex, marital)$ with (from left to right) a vspine by marital status, a vspine by marital status and sex, and a hspine by marital status, sex and happiness.
 For all levels of marital status, men are slightly less happy.}
@@
 
-% Note that while $f(x, y, z)$ and $f(z, y, x)$ represent the same pmf, the factorisation will be different because you will get different marginal tables.
-% 
-% Writing the full pmf is tedious, so we define an algebra based on the Wilkinson's modelling specification \citep{wilkinson:1973}:
-% 
-% \begin{itemize}
-%   \setlength{\itemsep}{0em}
-%   \item \verb|~ a| represents $f(a)$,
-%   \item \verb|~ a + b| represents $f(a, b)$
-%   \item \verb!~ a | b! represents $f(a | b)$
-%   \item \verb!~ . | a! is a special case, $f(a) = \frac{1}{\#A}$.
-% \end{itemize}
-%
+Formally, a product plot is constructed by the following recursive partitioning algorithm, which takes three parameters: {\sf data}, a multi-dimensional array, with dimensions ordered in the same way as the desired partitions; {\sf bounds}, a vector giving the top, left, right and bottom boundaries; and {\sf divider}, a list of the desired low-d drawing primitives.
+The following algorithm gives an rough idea of the computation:
 
-% To visualise higher-dimensional pmfs, we recursively partition using the previously described primitives.
-That is, to display $f(x, y, z)$ we first create a partition based on $f(z)$ then partition each of those pieces based on $f(y | z)$ and again by $f(x | y, z)$.
+\begin{itemize}
+  \item Calculate the one marginal and $i$ conditional distributions.
 
-Different combinations of partitions reveal different features of the data.
+  \item For each probability in the marginal distribution, divide the current bounds into $i$ sets of new bounds, one for each level of the categorical variable.
+The new bounds are found using the algorithm defined by the drawing primitive.
+
+  \item For each new bound and matching conditional distribution, call the partition function recursively, dropping one element from the list of drawing primitives.
+
+\end{itemize}
+
+Different partitions reveal different features of the data.
 Take for example, the distribution of age and marital status, as shown in Figure~\ref{fig:fluct}.
-Instead of visualisation the joint distribution with a fluct, we could focus on the conditional distribution of marital status, given age, or vice versa.
+Instead of visualising the joint distribution with a fluct, we could focus on the conditional distribution of marital status given age, or age given marital status.
 Figure~\ref{fig:marital} shows two ways to do this.
 The left plot shows $f(marital, age)$ with a vspine nested in a hspine, and the right plot $f(age, marital)$ with a hbar nested in a vspine.
-These displays show the same data but are arranged in such a way to make different comparisons easier.
-The second plot, with bars nested inside spines, also illustrates important feature of non-space filling tilings: the relationship between proportion and area is only constant within a level.
-The bars are scaled to be as high as possible, without overflowing the bounding area.
+These displays show the same data, but support different comparisons: on the left, we can see that most young people are unmarried, and on the right, that few unmarried people are over the age of 30.
+The right plot, with bars nested inside spines, also illustrates an important feature of non-space filling tilings: the relationship between proportion and area is only constant within a level.
+The bars are scaled to be as tall as possible, without overflowing any bounding region.
 
@@
 
-Conditioning is also an important tool by itself, because it allows one to remove relationships that are well-known or uninteresting.
+Conditioning is also an important tool by itself, because it allows one to remove relationships that are known or uninteresting.
 Figure~\ref{fig:part-cond} uses a fluct and a vspine to explore the relationship between happiness, health and financial status.
 The left plot displays raw proportions, showing that most people are in good health and average financial standing.
-However, it is difficult to see how happiness varies within these conditions because the perceptual task is comparing areas.
-Conditioning on financial status and health produces the plot on the right (an equal bin size plot) and makes it easier to see the conditional distribution of happiness given sex and health, because the perceptual task is simpler.
-Depending on the comparison we are most interested in, we can make it easier to compare across wealth for fixed health, or health for fixed wealth.
-
-For this data, for a fixed income level, happiness always increases as health increases.
-The same is not true for fixed health: those with poor or fair health are actually less happy if they are far above average, compared to just above average.
+However, it is difficult to see how happiness varies within these conditions because we must compare areas, not positions.
+Conditioning on financial status and health produces the plot on the right (an equal bin size plot) and makes it easier to see the conditional distribution of happiness given sex and health, because comparing positions along a common scale is an easier perceptual task.
+Depending on the comparison we are most interested in, we can make it easier to compare across wealth given health, or health given wealth.
+Here we see that for a fixed income level, better health is correlated to increased happiness.
+The same is not true for a fixed level of health: rich people with poor health seem to be less happy than poorer people in poor health.
 
@@
   \caption{(Left) $f(happy, health, finrela) = f(happy | health, finrela) \times f(health, finrela)  $, partitioned with a vspine and fluct.
+Health is on the $x$-axis, financial status on the $y$-axis.
 (Middle) $f(happy | health, finrela)$.
@@
 
-Many existing plots fall into this framework. The primitives already have their own names:
+Many existing plots fall into this framework. The low-d primitives already have their own names:
 
@@
 
-  \item {\bf Stacked} bar chart (2d). 1 hbar and 1 vspine.
+  \item {\bf Stacked} bar chart (2d). $1$ hbar and $1$ vspine.
 
-  \item {\bf Nested} bar chart \citep{peltier:2009} (2d).  2 hbars.
+  \item {\bf Nested} bar chart \citep{peltier:2009} (2d). $2$ hbars.
 
-  \item {\bf Equal bin size} \citep{hofmann:2000} plot (3d): fluct and vspine, conditioned on the first two variables.
+  \item {\bf Equal bin size} plot \citep{hofmann:2000} (3d): $1$ fluct and $1$ vspine, conditioned on the first two variables.
 
@@
 
-  \item {\bf Double-decker} plot \citep{hofmann:2001} (nd). $n-1$ hspines and 1 vspine.
+  \item {\bf Double-decker} plot \citep{hofmann:2001} (nd). $n-1$ hspines and $1$ vspine.
 
-  \item {\bf Treemap} \citep{shneiderman:1992} (nd): n spines.
+  \item {\bf Treemap} \citep{shneiderman:1992} (nd): $n$ spines.
 
-  \item {\bf Squarified treemap} \citep{bruls:1999} (nd): n tiles.
+  \item {\bf Squarified treemap} \citep{bruls:1999} (nd): $n$ tiles.
 
@@
 
-Trellis graphics \citep{becker:1996}, also known as lattice, facetted and conditioned graphics, are another related display.
+Trellis graphics \citep{becker:1996}, also known as latticed, facetted and conditioned graphics, are another related display.
 They use categorical variables to generate multiple panels, each containing a plot of the subset of the data.
-Trellised area plots also fall into our framework and can be created by conditioning on the trellising variables.
+Trellised plots of area graphics also fall into our framework and can be created by conditioning on the trellising variables.
+
+\section{Display considerations}
+\label{sec:display}
+
+Labelling product plots is challenging.
+In this paper, we use a carefully selected combination of colour and axis labels, as well displaying the process by which a product plot is created.
+Axis labels are only available under certain conditions: when we have rows or columns that all share the same dimensional index.
+This can occur because of the structure of the graphical primitive (e.g.
+bars, flucts), or because the display is of a conditional distribution.
+Take Figure~\ref{fig:marital} as an example: the left plot can only have the $x$ axis labelled, but the right plot can have both axes labelled.
+There are other ways to label the cells apart from colour.
+\citep{slingsby:2009} uses text labels, sized to occupy the space of the region they label.
+In the interactive setting, dynamic labelling in the form of tool tips is extremely helpful, as is the ability to use linked brushing to connect high-d plots to low-d plots.
+
+We have observed that some aspect ratios are more aesthetically pleasing than others.
+This is particularly obvious for the fluctuation diagram where, in our experience, the plots are most appealing when the flucts are square.
+In other plots there does not seem to be an easy rule of thumb, but we wonder if aspect ratios close to the golden ratio might be more appealing.
+We are not aware of any previous work on aesthetics of aspect ratio, although they do seem to affect perception \citep{kong:2010}.
+More research in this area is needed, particularly to explore the balance of aesthetics and usability.
 
@@
 On top of this basic framework, it is useful to consider a few variations and extensions.
-Instead of count of discrete levels, we can plot weighted data or continuous data, or we can relax the display constraints to allow displays where area is not proportional to weight, partitions are non-disjoint or non-rectangular.
+Instead of counts, we can plot weighted data or continuous data, or we can relax the display constraints to allow displays where area is not proportional to weight, partitions are non-disjoint or non-rectangular.
 
@@
 Figure~\ref{fig:weighting} shows the difference between the weighted and unweighted distributions of age and sex.
-The distribution is subtly different, and suggests that we don't need to worry about weights for this plot.
-In other datasets, weights can be useful to move from numbers of counties to numbers of people, or to areas, or to other relevant quantities.
+The distribution is barely different, and suggests that we don't need to worry about weights for this plot.
+(Unfortunately for this dataset of we have been unable to find any plots where weighting makes a big difference.)  In other datasets, weights can be useful to move from numbers of counties to numbers of people, or to areas, or to other relevant quantities.
 Some examples of weighted data graphics can be found in \citep{unwin:1998,unwin:2003aa,unwin:2006}.
@@
     \includegraphics[width=0.5\linewidth]{wt-wtssall}%
-  \caption{Joint distribution of age and sex (\key{male} male, \key{female} female). (Left) counts and (right) probability weights.}
+  \caption{Joint distribution of age and sex (\key{male} male, \key{female} female). (Left) counts and (right) probability weights. In this case, there is very little difference between the plots.}
   \label{fig:weighting}
@@
 
-The framework can be trivially extended to work with continuous data: just first bin continuous variables.
+The framework can be trivially extended to work with continuous data: just  bin continuous variables to make them discrete.
 There are many different to ways to create intervals for continuous data, but two are most important \citep{denby:2009}: intervals of equal width, and intervals containing equal numbers of points.
-This extension allows the product plot framework to also describe {\bf histograms} and {\bf spinograms} \citep{hummel:1996}, continuous analogues of the bar and charts.
+This extension allows the product plot framework to also describe {\bf histograms} and {\bf spinograms} \citep{hummel:1996}, continuous analogues of the bar and charts, shown in Figure \ref{fig:cont-examples}.
 A long standing tradition is that no gaps are displayed between adjacent rectangles when used for continuous data.
-Two examples are shown in Figure \ref{fig:cont-examples}.
 
@@
 
-  \caption{The distribution of happiness with age. (Left) hbar + vspine: an extension of the histogram. (Right) hspine + vspine: an extesion of the spinogram.}
+  \caption{The distribution of happiness with age. (Left) hbar + vspine: an extension of the histogram. (Right) hspine + vspine: an extension of the spinogram.}
   \label{fig:cont-examples}
@@
 
-With one more extension, displaying the innermost proportion with colour (instead of area), this allows us to describe {\bf dimensional stacking} \citep{leblanc:1990} as a plot of binned, conditioned continuous data, with proportion shown by colour.
+With one more extension, displaying the innermost proportion with colour (instead of area), we can also describe {\bf dimensional stacking} \citep{leblanc:1990} as a plot of binned, conditioned continuous data, with proportion shown by colour.
 
@@
 
-It can be useful to violate the constraint that area is proportional to value to distinguish between zeros and very small values.
-A zero weight should have zero area, but giving it positive area can be useful that so you can see it!
-In general, it's useful to constrain all areas to be above a certain minimal perceptible size.
-Areas which are constrained in such a way should have a visual flag (such as a different colour) to ensure that the reader knows that the relationship between area and value has been violated.
+It can be useful to violate the constraint that area is proportional to value to distinguish between zeros, missing values and very small values.
+A zero weight should have zero area, but giving it positive area can be useful that it is visible.
+In general, it's useful to constrain all areas to be above a certain minimal perceptible size (e.g.\ 4 square pixels).
+Areas which are constrained in such a way need a visual flag (such as a different colour) to ensure that the reader knows that the relationship between area and value has been violated.
 This type of non-linear mapping has been implemented in {\sc manet} \citep{unwin:1996} and in hierarchical pixel bar charts \citep{keim:2002}.
-
-On the other end of the spectrum, it can be useful to constrain the size of largest values to get {\bf censored zooming} \citetext{Antony Unwin, priv.
-comm.}.
-Controlling this cap makes it possible to focus on small values.
+At the other end of the spectrum, it can be useful to constrain the size of largest values to get censored zooming \citetext{Antony Unwin, priv.
+comm.}, which makes it possible to focus on small values.
 
@@
     \includegraphics[width=0.5\linewidth]{cascade}
-  \caption{A mosaic plot of happiness by martial status and sex.
+  \caption{A mosaic plot of happiness by marital status and sex.
 (Left) Coloured by happiness (\key{not-too-happy} Not too happy, \key{pretty-happy} pretty happy, \key{very-happy} very happy).
@@
 Many other circular displays turn out to be special cases of product plots drawn in polar coordinates \citep{draper:2009}.
-Generally, the y axis (mapped to radius) must be square-root transformed to ensure that that counts stay proportional to areas.
-Fan-lenses \citep{lou:2007} and Stasko's radial displays \citep{stasko:2000} deliberately do not do this in order to emphasise the outer levels.
-
 We have identified the following radial plots as polar transformations of product plots:
@@
 Many of these are familiar and already have names.
-They are all of dubious utility because of the increased difficulty when comparing non-rectangular areas.
+They are all of dubious utility because research suggests that visualisations in polar coordinates are harder to reader accurately than visualisations in Cartesian coordinates \citep{diehl:2010}.
+
+Generally, the y axis (mapped to radius) must be square-root transformed to ensure that that counts stay proportional to areas.
+Fan-lenses \citep{lou:2007} and Stasko's radial displays \citep{stasko:2000} deliberately do not do this in order to emphasise the outer levels.
 
@@
 A number of non-rectangular treemaps have been proposed, such as circular treemaps \citep{wetzel:2008}, space-filling curves \citep{wattenberg:2005} and voronoi treemaps \citep{balzer:2005}.
-The perceptual task associated with these visualisations is the comparison of the area of arbitrary polygons, which is much more difficult than comparing the area of rectangles \citep{cleveland:1986}.
+Reading these visualisations requires the comparison of the area of arbitrary polygons, which is much more difficult than comparing the area of rectangles \citep{cleveland:1986}.
 This makes these approaches attractive rather than useful.
 
-% \section{Aesthetics}
-% \label{sec:aesthetics}
-% 
-% \subsection{Labelling} 
-% \label{sub:labelling}
-% 
-% \subsection{Offsets}
-% \label{sub:offsets}
-% 
-% 
-% \section{Interaction}
-% \label{sec:interaction}
+\section{R package}
+\label{sec:package}
 
-\section{Conclusions}
+We have provided a reference implementation of these ideas in an R \citep{R} package called {\tt prodplots}, available from {\tt http://github.com/hadley/prodplots}.
+There are two main functions: {\tt prodcalc}, which computes the coordinates of each rectangle; and {\tt prodplot}, which displays the rectangles with the ggplot2 package \citep{me:ggplot2}.
+The code is well tested, and ensures that the constraints are always satisfied.
+The {\tt prodplots} code to used to create the images in this paper is available from the authors.
 
-This framework leads to a combinatorial explosion of possibilities. A 4d {\sc pmf}, $f(a, b, c, d)$, can be factorised in five different ways:
+The {\tt prodplots} package has been designed to be flexible and extensive.
+For example, each graphical primitive is represented by a function: {\tt hspine()}, {\tt vspine()}, {\tt spine()}, {\tt hbar()}, {\tt vbar()}, {\tt tile()} and {\tt fluct()}.
+Adding a new graphical primitive is easy: you just write a new function, and can use the existing calculation and display algorithms.
+
+\section{Conclusion}
+\label{sec:conclusion}
+
+The product plots framework is successful at describing many existing graphics for tables of counts, proportions and probabilities.
+It lays the framework for much future work, particularly because the framework leads to a combinatorial explosion of possibilities.
+For example, a 4d {\sc pmf}, $f(a, b, c, d)$, can be factorised in five different ways:
 
@@
 \noindent There are 24 possible ways of ordering the variables in the {\sc pmf}, 5 ways of displaying a 1d {\sc pmf}, and 1 way of displaying a 2d {\sc pmf}, leading to a possible $24 * (5^4 + 5^2 + 5^2 + 5^2 + 1) = 16,824$ plots, before we even consider conditioning!
-Well thought out interaction can make it easy to navigate this space, but it In future work, we hope to connect product plots to statistical models to help guide the choice of appropriate graphic.
+Well thought out interaction can make it easy to navigate this space.
+We have a prototype interaction model developed as part of the cranvas suite of interactive graphics in R, {\tt http://github.com/ggobi/cranvas}, but much work and user testing remains to be done.
 
-We are also interested in exploring the automated construction of a graphic to answer specific questions, and for given a plot, suggesting which type of questions it is most suitable to answer.
+We also are hopeful that developing the connection between product plots and log-linear (aka Poisson) models will also help better understand this space of plots.
+Some special cases have already been worked out.
+For example, \citep{hofmann:2003} shows how looking for straight lines in a mosaic plot corresponds to a formal test of independence between two variables.
+Generally, what visual comparisons are equivalent to what formal statistical tests?
+And conversely, how can inspecting the significant coefficients in a model help us choose a set of useful plots for a given dataset?
+Generally, given a question, can we suggest appropriate plots?
+Or given a plot, can we suggest questions that it might answer?
 
-% \acknowledgments{ The authors wish to thank Robert Kosara for helpful discussion.
-This work was partly supported by the National Science Foundation grant DMS0706949.
-Graphics produced with ggplot2 \citep{me:ggplot2}.}
+Product plots need to be extended with systematic ways of displaying uncertainty, to help users identify whether differences in the plot represent real differences in the underlying population or are just the result of random variation.
+This is particularly important because area plots typically have fixed area, meaning that the total number of observations in a plot is not directly displayed.
+This is crucial for determining whether a difference is truly significant or not.
+
+Finally, continuous data is currently incorporated in a simplistic manner.
+Can we extend the product plots framework to include plots designed to show continuous distributions like the average shifted histogram \citep{scott:1985} or the kernel density estimate \citep{scott:1992}?
+
+\acknowledgments{Computations performed in R \citep{R} and graphics produced with ggplot2 \citep{me:ggplot2}.}
 

### 2011-03-30 colour.r

@@ -1,3 +1,5 @@
+theme_set(theme_grey(16))
+
 fill_happy <- list(
   aes(fill = happy),
   opts(legend.position = "none"),

### 2011-03-30 plot-constraints.r

@@ -2,15 +2,20 @@ l(productplots)
 source("colour.r")
 
 prodplot(happy, ~ happy + marital | sex, na.rm = T, 
-  set_offset(c("vspine", "hspine", "hspine"), c(0, 0.01, 0.02))) + fill_happy
+  set_offset(c("vspine", "hspine", "hspine"), c(0, 0.01, 0.02))) + 
+  fill_happy  + 
+  scale_x_continuous(breaks = NA) + 
+  scale_y_continuous(breaks = NA)
 ggsave("plots/sex-marital-happy.pdf", width = 6, height = 4)
 
 df <- prodcalc(happy, ~ happy + marital | sex, na.rm = T, set_offset(c("vspine", "hspine", "hspine"), c(0, 0.01, 0.02)), cascade = 0.1)
 
-ggplot(df, aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, order = level)) + 
+ggplot(df, aes(xmin = l, xmax = r, ymin = b, ymax = t, order = level)) + 
   geom_rect(aes(colour = level, size = -level), fill = alpha("white", 0.9)) +
   xlab(NULL) + ylab(NULL) + 
   scale_colour_gradient("Level", low = "black", high = "grey60", breaks = 1:3) + 
   scale_size(to = c(1, 2)) +
-  opts(legend.position = "none")
+  opts(legend.position = "none") + 
+  scale_x_continuous(breaks = NA) + 
+  scale_y_continuous(breaks = NA)
 ggsave("plots/cascade.pdf", width = 6, height = 4)

### 2011-03-30 plots-partition.r

@@ -50,10 +50,10 @@ ggsave("plots/part-comb-3.pdf", width = 6, height = 6)
 
 
 # alternative displays of marital status & age
-prodplot(happy, ~ marital + decade, set_offset(c("vspine", "hspine"), c(0, 0.02)), na.rm = T) + fill_marital
+prodplot(happy, ~ marital + decade, set_offset(c("vspine", "hspine"), c(0, 0.02)), na.rm = T, levels = NA) + fill_marital
 ggsave("plots/part-marital-1.pdf", width = 6, height = 6)
 
-prodplot(happy, ~ decade + marital, set_offset(c("hbar", "vspine"), c(0, 0.02)), na.rm = T) + fill_marital
+prodplot(happy, ~ decade + marital, set_offset(c("hbar", "vspine"), c(0, 0.02)), na.rm = T, levels = NA) + fill_marital
 ggsave("plots/part-marital-2.pdf", width = 6, height = 6)
 
 prodplot(happy, ~ happy + finrela, c("vspine", "hspine")) + 
@@ -65,15 +65,15 @@ prodplot(happy, ~ happy + marital, c("vspine", "hspine")) +
 
 # fluctuation & equal bin size
 prodplot(happy, ~ happy + finrela + health, na.rm = TRUE, 
-  set_offset(c("vspine", "fluct"), c(0, 0.05)), subset = .(level == 2)) +
+  set_offset(c("vspine", "fluct"), c(0, 0.05)), levels = NA) +
   fill_happy
 ggsave("plots/part-fluctuation.pdf", width = 6, height = 6)
 
 prodplot(happy, ~ happy | finrela + health, na.rm = TRUE,
-  set_offset(c("vspine", "fluct"), c(0, 0.05))) + fill_happy
+  set_offset(c("vspine", "fluct"), c(0, 0.05)), levels = NA) + fill_happy
 ggsave("plots/part-equal-area.pdf", width = 6, height = 6)
 
 prodplot(happy, ~ happy | finrela + health, na.rm = TRUE,
-  set_offset(c("hspine", "fluct"), c(0, 0.05)), subset = .(level == 2)) +
+  set_offset(c("hspine", "fluct"), c(0, 0.05)), levels = NA) +
   fill_happy
 ggsave("plots/part-equal-area-2.pdf", width = 6, height = 6)

### 2011-03-30 prodplots-infovis.tex
Fix name of package\!

@@
 
-We have provided a reference implementation of these ideas in an R \citep{R} package called {\tt prodplots}, available from {\tt http://github.com/hadley/prodplots}.
+We have provided a reference implementation of these ideas in an R \citep{R} package called {\tt productplots}, available from {\tt http://github.com/hadley/productplots}.
 There are two main functions: {\tt prodcalc}, which computes the coordinates of each rectangle; and {\tt prodplot}, which displays the rectangles with the ggplot2 package \citep{me:ggplot2}.
 The code is well tested, and ensures that the constraints are always satisfied.
-The {\tt prodplots} code to used to create the images in this paper is available from the authors.
+The {\tt productplots} code to used to create the images in this paper is available from the authors.
 
-The {\tt prodplots} package has been designed to be flexible and extensive.
+The {\tt productplots} package has been designed to be flexible and extensive.
 For example, each graphical primitive is represented by a function: {\tt hspine()}, {\tt vspine()}, {\tt spine()}, {\tt hbar()}, {\tt vbar()}, {\tt tile()} and {\tt fluct()}.

### 2011-06-20 reviews/response.txt
Add reviews

# Summary review

The reviewers judge Product Plots to be a useful and parsimonious model
for constructing a large variety of plots and potentially also an example
that might spur further research in formal models of visualization
construction. All reviewers agree that this work makes a substantive
contribution to InfoVis research and argue for acceptance.

That said, reviewers also raise a number of issues that the authors
should address prior to publication. Most notable among these is the
treatment of related work. R1 and R4 in particular list elements of prior
work that should be considered and discussed. I recommend that the
authors introduce a dedicated related work section in their next
revision.

In addition, the reviewers make a number of helpful suggestions that
might inform future drafts. I strongly encourage the authors to revise
the paper accordingly.

# Major changes

* Related work section & missing references
* Improve justification of rectangles using references provided.

# Review 2

* First, related work is distributed throughout the paper. A consolidated
related work section addressing similar work and clearly delineating the
contributions of the present work is needed. Other visualization models
of interest include Slingsby et al's HiVE, Wilkinson's Grammar of
Graphics, Mackinlay's APT, and Stolte et al's Polaris. A detailed
comparison with such prior work is a pre-requisite to publication. How do
such models compare in terms of expressiveness and notational efficiency?

* I'm curious how much effort is necessary to use product plots in
practice. Some examples of statements in R and the corresponding plots
should suffice (and potentially examples of an interactive UI for product
plot construction).

* Is anyone using the system to perform analyses? While not strictly
necessary, example analysis case studies supported by product plots might
help convey the value of the model.

* "the areas of rectangles are easier to compare than areas of other shapes
[5-7]" -> This is somewhat glibly stated. A quick review of the cited
papers does not support the claim: [5] does not assess area empirically;
[6] assesses circular and arbitrary polygonal areas, but not rectangles;
[7] compares circles and rectangles, but the difference does not appear
to be statistically significant. Later in the paper, there is a similar
critique of non-rectangular polygons using the results of [6]. I don't
think this is appropriate -- [6] uses irregular polygons, whereas
techniques like voronoi treemaps generate convex polygons with average
bounding box aspect ratios near unity. Thus the argument here is also
unsubstantiated. It may well be that rectangles are better, but this has
not been convincingly demonstrated. I recommend the authors consider
dropping these perceptual arguments.

# Review 4

* A surprising omission is the work of Wilkinson (1999, 2005) on The
 Grammar of Graphics. Wilkinson introduced the idea of combining graphical
 elements through products of sets (called facets) and embedding these in
 different coordinate systems. In fact, Figure 12 echoes several of the
 figures in the Coordinates chapter of that book. Wilkinson was also the
 first to draw a link between the recursive partitioning in treemaps and
 mosaics and a set product operator. As he explains in Chapter 16 of the
 second edition, the algebra for specifying visualizations of product
 models maps one-to-one to the most common factorial experimental design
 algebra for log-linear models or ANOVA. In any case, the framework in
 this paper appears closer to that of Wilkinson's than to any of the other
 papers cited.

 Bottom of page:
 Using the word "all" is a bit strong here. Better to delete the word and
 confine the discussion to "area" plots. There are exceptions, as the
 authors point out later, so avoid the inclusive term.

 The claim for rectangles at the bottom of page 1 is not supported by
 research. The authors, like many statisticians and visualization
 researchers, tend to cite statisticians and computer scientists for
 supportive psychological principles; however, the perceptual and
 cognitive research by non-psychologists tends to have serious
 methodological flaws. I've provided some relevant citations below. There
 are significant biases in area judgments that are exacerbated by extreme
 aspect ratios (something even Kong, Heer & Agrawala noticed in Treemaps).
 The perceptual combining rule for rectangles appears to be more Attneave
 (absolute metric) than product (multiplicative), although see Algom et
 al. for support for the multiplicative rule. The authors need to rephrase
 this statement, eliminating the word "must."

 Also, Simken and Hastie, in one of the few perception articles published
 by psychologists in a statistical journal, demonstrated conclusively that
 pie charts outperform rectangular area charts for proportion of whole
 data. There are some exceptions to their findings, but statisticians need
 to understand that they are largely wrong in their scorn for pie charts.

 We can tile triangles with triangles. And these are "simpler" than
 rectangles. One can also tile the sphere with triangles, which cannot be
 done with rectangles. You need to rethink this assertion.

 References:
 Algom, D., Wolf, Y., & Bergman, B. (1985). Integration of stimulus
 dimensions in perception and memory: Composition rules and psychophysical
 relations. Journal of Experimental Psychology: General, 114, 451-471.

 Anderson, N.H. and Weiss, D.J. (1971). Test of a multiplying model for
 estimated are of rectangles. American Journal of Psychology, 84, 4,
 543-548.

 Borg, I. and Leutner, D. (1983). Dimensional models for the perception of
 rectangles. Perception & Psychophysics, 34, 257-267.

 Kidman, G. (1997). Area integration rules. In F. Biddulph & K. Carr
 (Eds.), People in mathematics (Proceedings of the 20th annual conference
 of the Mathematics Education Research Group of Australasia (Vol. 1, pp.
 271 - 277).

 Mates, J., Di Maio, V., and Lansky, P. (1992). A model of the perception
 of area. Spatial Vision, 6, 101-116.

 Morgan M.J. (2005). The visual computation of 2-D area by human
 observers. Vision Research, 45(19), 2564-70.

# Review 1
 Finally, I wonder if the authors considered any other graphical
 composition method than nesting areas? While nesting areas allowed them
 to encompass a number of existing plot types, other composition methods
 might be more useful for extension to higher dimensions. For example,
 what about simply showing a separate plot for each component?

 -label all appropriate plots in the paper with the corresponding
 distribution decomposition (directly label, not just mention in the text
 or in the caption)
 -flesh out section 3.1. Assuming that infovis readers will not know what
 conditional and marginal distributions are, the authors should
 demonstrate how they are computed from the joint distribution. Providing
 the graphical analogs to the tables in Figure 4 would be extremely
 helpful. (I realize that a marginal and a conditional distribution from
 Figure 4 are represented in Figure 5, but the connection is never made
 explicitly in the text. Some hand holding would be very helpful here.)
 Hand hold better the analytic questions that each distribution helps
 answer. E.g. "What proportion of males are very happy? The bottom left
 entries tells us it's 31%."

 Minor comments----

 Why squarified treemaps? Heer work showing non-square treemaps are easier
 to read.

 Section 3.1

 Its not clear from the wording in the first paragraph that the marginal
 and conditional distributions contain less information than the joint
 distribution. If you're going to assume that your readers don't know what
 the marginal and conditional distributions are (plausible for the InfoVis
 crowd), then it would make sense to briefly describe how they are
 computed from the joint distribution.

 Fix the formatting of the tables in Figure 4.

 Clarify in the last paragraph that in addition to the 3 factorings,
 permutations of the variables leads to more variations.

 "It remains to find an [sic] graphical analog of this multiplication."
 seems unnecessarily dramatic. We already know you found one.

 Section 3.2

 Label Figure 5 with the distribution forms, f(happy), f(happy|sex),
 f(happy, sex), rather than just mention them in the text.

 Figure 7 shows, f(marital | age) f(age) and f(age | marital) f(marital),
 right? Say this in the text and label figures.

 "the relationship between proportion and area is only constant within a
 level" -- is there a better way to say this? I think you mean level as in
 nesting level, not level as is a level of a factor, right? In Figure 7
 (right), the nested bar charts are all sized proportionally with respect
 to each other even though they occur in different levels of the marital
 status factor.

 Figure 8 middle would be better with a bar chart, right? Why show both
 the middle and right plots; they're basically the same.

 Section 6.2

 Why limit yourself to a finite number of bins? Push the number of bins to
 infinity and you can display line graphs (density plots) with hbar,
 stacked line graphs (hbar+vspine), relative area (hspine+vspine),
 overlaid line charts (hbar+hbar), or (hspine+hbar) (don't know the name
 of this chart, basically overlaid line charts, but where the sum of
 values at each vertical slice is 1).

# Review 3

 I had some problems, though, in classifying this paper - it falls
 somewhere between theory/model (describes a theoretical framework that
 encompasses a large number of existing visualisations) and technique
 (code is provided to implement the framework). I'm not sure if it'd be
 better to simplify the classification by omitting mention of the code
 except as a note near the end - while it's interesting, it's not core to
 the paper.

 Doing this would leave more space to address some other minor concerns
 with the figures: in figures 2, 3, 5, 6 and 7 it's quite clear that the
 labels on the plot have spacing issues that could be resolved with larger
 figures. Additionally, figure 4 spills over the gutter into the text in
 the right column.

 In terms of text, I noticed some typos (first page, "This framework is
 based on the fundamental statistical idea that high any high dimensional
 distribution" has too many "high"s), one slight miss in phrasing ('weed
 down' sounds like a mixed metaphor between 'weed out' and 'cut down') and
 perhaps the last column of the conclusion could benefit from editing
 (repetition of 'generally' stands out).

# Minor changes

* There is a line break after "Abstract-". The abstract text should begin
on the same line.

* Figure 4 overruns its column.

* Page 2. "Uniformly distributed" suggest to me a uniform random distribution. Is there another way to say this?

* Figure 3. Left and Right are Top and Bottom. LaTeX refused to obey your
orders, as usual ;o)

* Please italicize n in the fourth line of Section 3. Check that other
parameters are properly italicized in the text.

* Page 1. "This framework is based on the fundamental statistical idea that
 high any high dimensional" Delete the first "high."

 "51 020 and 5 000" use comma delimiters.

*  Figure 12. The bottom row is not produced by transforming after rotation.
 It is produced by switching the assignment of x and y to r and theta.
 This is a pivoting operation.

### 2011-06-21 prodplots-infovis.tex
Check in a few final tweaks

@@
 
-We propose a new framework for visualising tables counts, proportions or probabilities.
+We propose a new framework for visualising tables of counts, proportions and probabilities.
 We call our framework product plots, alluding to the computation of area as a product of height and width, and the statistical concept of generating a joint distribution from the product of conditional and marginal distributions.
@@
 
-We begin in Section~\ref{sec:primitives} by motivating the three constraints that give rise to the basic graphical component of product plots, the rectangle.
+We begin in Section~\ref{sec:primitives} by motivating the three constraints that give rise to the basic graphical component of product plots: the rectangle.
 We constrain the rectangle further to produce three 1d atoms (bar, spine, tile) and one 2d atom (the fluct).
@@
 Section~\ref{sec:combination} provides the mathematical framework that allows us to combine the 1d and 2d atoms to display data of any dimensionality.
-This framework is based on the fundamental statistical idea that high any high dimensional distribution can be factorised into a product of low-dimensional conditional and marginal distributions.
+This framework is based on the fundamental statistical idea that any high dimensional distribution can be factorised into a product of low-dimensional conditional and marginal distributions.
 Section~\ref{sec:existing} shows how many existing named graphics are special cases of this general pattern.
@@
 The {\sc gss} is a yearly cross-sectional survey of Americans, run from 1976.
-We combine data for 25 years to yield 51\,020 observations, and of the over 5\,000 variables, we select nine related to happiness, as described in Table~\ref{tbl:happy}.
+We combine data for 25 years to yield 51,020 observations, and of the over 5,000 variables, we select nine related to happiness, as described in Table~\ref{tbl:happy}.
 
@@
 The total area for a graphic is usually constrained, which means that area plots display typically proportions or probabilities, rather than counts.
-We will use count, probability and proportion interchangeably throughout the paper.
+With this proviso, we will use count, probability and proportion interchangeably throughout the paper.
 
@@
 This means that we can turn any high-dimensional {\sc pmf} into a product of low-dimensional conditional and marginal {\sc pmf}s, a computationally trivial operation.
-It remains to find an graphical analog of this multiplication.
+It remains to find a graphical analog of this multiplication.
 
@@
 
-We connect probability products to our rectangular primitives by noting that areas are also products, products of height and width.
+We connect probability products to our rectangular primitives by noting that areas are also products: products of height and width.
 It's easiest to show this with a picture: Figure~\ref{fig:fact-simple} shows how our simple 1d primitives combine to get two familiar plots: the mosaic plot and the stacked bar chart.
-From left to right, we have plots of $f(happy)$, $f(happy | sex)$ and the product $f(happy, sex)$.
+From left to right, we have plots of $f(happy)$, $f(sex | happy)$ and the product $f(happy, sex)$.
 The heights and widths of the rectangles are multiplied in the same way as the components of the {\sc pmf}.
@@
 
-Formally, a product plot is constructed by the following recursive partitioning algorithm, which takes three parameters: {\sf data}, a multi-dimensional array, with dimensions ordered in the same way as the desired partitions; {\sf bounds}, a vector giving the top, left, right and bottom boundaries; and {\sf divider}, a list of the desired low-d drawing primitives.
+More precisely, a product plot is constructed by the following recursive partitioning algorithm, which takes three parameters: {\sf data}, a multi-dimensional array, with dimensions ordered in the same way as the desired partitions; {\sf bounds}, a vector giving the top, left, right and bottom boundaries; and {\sf divider}, a list of the desired low-d drawing primitives.
 The following algorithm gives an rough idea of the computation:
@@
 Figure~\ref{fig:marital} shows two ways to do this.
-The left plot shows $f(marital, age)$ with a vspine nested in a hspine, and the right plot $f(age, marital)$ with a hbar nested in a vspine.
+The left plot shows $f(marital, age)$ with a vspine nested in a hspine, and the right plots $f(age, marital)$ with a hbar nested in a vspine.
 These displays show the same data, but support different comparisons: on the left, we can see that most young people are unmarried, and on the right, that few unmarried people are over the age of 30.
@@
 However, it is difficult to see how happiness varies within these conditions because we must compare areas, not positions.
-Conditioning on financial status and health produces the plot on the right (an equal bin size plot) and makes it easier to see the conditional distribution of happiness given sex and health, because comparing positions along a common scale is an easier perceptual task.
+Conditioning on financial status and health produces the plots on the right (equal bin size plots) and makes it easier to see the conditional distribution of happiness given sex and health, because comparing positions along a common scale is an easier perceptual task.
 Depending on the comparison we are most interested in, we can make it easier to compare across wealth given health, or health given wealth.
@@
 bars, flucts), or because the display is of a conditional distribution.
-Take Figure~\ref{fig:marital} as an example: the left plot can only have the $x$ axis labelled, but the right plot can have both axes labelled.
+Take Figure~\ref{fig:marital} for example: the left plot can only have the $x$ axis labelled, but the right plot can have both axes labelled.
 There are other ways to label the cells apart from colour.
@@
 For example, in the happy dataset, the {\sf wtssall} variable gives analytic survey weights.
-These are use to account for oversampling of black respondents in certain years, and to reduce the effect of non-response by other demographics.
+These are used to account for oversampling of black respondents in certain years, and to reduce the effect of non-response in other demographics.
 Figure~\ref{fig:weighting} shows the difference between the weighted and unweighted distributions of age and sex.
 The distribution is barely different, and suggests that we don't need to worry about weights for this plot.
-(Unfortunately for this dataset of we have been unable to find any plots where weighting makes a big difference.)  In other datasets, weights can be useful to move from numbers of counties to numbers of people, or to areas, or to other relevant quantities.
+(Unfortunately for this dataset of we have been unable to find any plots where weighting does makes a difference.)  In other datasets, weights can be useful to move from numbers of counties to numbers of people, or to areas, or to other relevant quantities.
 Some examples of weighted data graphics can be found in \citep{unwin:1998,unwin:2003aa,unwin:2006}.
@@
 The framework can be trivially extended to work with continuous data: just  bin continuous variables to make them discrete.
-There are many different to ways to create intervals for continuous data, but two are most important \citep{denby:2009}: intervals of equal width, and intervals containing equal numbers of points.
-This extension allows the product plot framework to also describe {\bf histograms} and {\bf spinograms} \citep{hummel:1996}, continuous analogues of the bar and charts, shown in Figure \ref{fig:cont-examples}.
-A long standing tradition is that no gaps are displayed between adjacent rectangles when used for continuous data.
+There are many different to ways to create bins for continuous data, but two are most important \citep{denby:2009}: bins of equal width, and bins containing equal numbers of points.
+This extension allows the product plot framework to also describe {\bf histograms} and {\bf spinograms} \citep{hummel:1996}, continuous analogues of bar and spine charts, shown in Figure \ref{fig:cont-examples}.
+A long standing tradition is that no gaps are displayed between adjacent rectangles when displaying continuous data.
+
+A more theoretical approach in dealing with continuous variables  lies in increasing the number of bins infinitesimally -- this leads from a probability mass function to a density function, making the 1d primitives of hbar and vbar to density plots, and changing the spine-plot representation to a conditional density plot (citation needed - probably Martin's and Heike's paper).
+While the inclusion of one continuous variable is completely feasible that way, two variables need the use of an additional aesthetic, such a s colour,  to visualize a 2d joint density.
+Three or more continuous variables break the scope of the product plots framework.
 
@@
 
-With one more extension, displaying the innermost proportion with colour (instead of area), we can also describe {\bf dimensional stacking} \citep{leblanc:1990} as a plot of binned, conditioned continuous data, with proportion shown by colour.
+With one more extension, displaying the innermost proportion with colour (instead of area), we can also describe {\bf dimensional stacking} \citep{leblanc:1990}.
 
@@
 It can be useful to violate the constraint that area is proportional to value to distinguish between zeros, missing values and very small values.
-A zero weight should have zero area, but giving it positive area can be useful that it is visible.
+A zero weight should have zero area, but giving it positive area can be useful so that it is visible.
 In general, it's useful to constrain all areas to be above a certain minimal perceptible size (e.g.\ 4 square pixels).
@@
 Figure~\ref{fig:cascade} shows an example of how cascading can help illuminate the structure of a complex mosaic plot.
-This technique is probably more effective when implemented interactively.
+This technique is probably most effective when implemented interactively.
 
@@
 A pie chart is a popular method of displaying proportions, but it is not a rectangular partition and so does not seem to fall in the framework of this paper.
-However, there is a simple relationship between product plots and pie charts: an pie chart is an hspine drawn in polar coordinates with the x coordinate mapped to angle and the y coordinate to radius.
+However, there is a simple relationship between product plots and pie charts: a pie chart is an hspine drawn in polar coordinates with the $x$ coordinate mapped to angle and the $y$ coordinate to radius.
 Many other circular displays turn out to be special cases of product plots drawn in polar coordinates \citep{draper:2009}.
@@
 
+\section{Related Work}
+\label{sec:related}
+
+The product plots framework is heavily influenced by the work of WIlkinson \citep{wilkinson:1999, wilkinson:2005}, who introduced the set product operator to map hierarchical data to a two-dimensional framework.
+Wilkinson restricts (???
+not quite sure - need to lookit up) the use to standard mosaicplots as defined by \citet{hartigan:1981}, which could easily be extended, however the product plots framework additionally introduces the concept of conditioning on variables.
+
+ \citep{slingsby:2009}: the HiVE system provides a notation allowing to describe states of hierarchical layouts and operators for reconfiguration.
+HiVE allows both rectangular layouts and polygon layouts -- out of the rectangular layouts, all but the spatially ordered algorithm are covered by the  framework of product plots.
+
+The framework of  Polaris  \cite{stolte:2002} relies heavily on a tabular or cubic format in rendering hierarchical graphics, which makes these charts essentially a three dimensional extension of the trellis framework \citep{cleveland:1994, cleveland:1996}:  in the product plots framework, this approach is captured by conditioning on the variables used in rows and columns, and displaying the other variables unconditioned in each cell of the resulting tabular layout.
+
+ \cite{mackinlay:1986, mackinlay2007}: VisQL, a visual query language is designed to capture the flow of a graphical data exploration.
+
 \section{R package}
@@
 
-The product plots framework is successful at describing many existing graphics for tables of counts, proportions and probabilities.
+The product plots framework is successful at describing many existing graphics that display tables of counts, proportions and probabilities.
 It lays the framework for much future work, particularly because the framework leads to a combinatorial explosion of possibilities.
@@
 \noindent There are 24 possible ways of ordering the variables in the {\sc pmf}, 5 ways of displaying a 1d {\sc pmf}, and 1 way of displaying a 2d {\sc pmf}, leading to a possible $24 * (5^4 + 5^2 + 5^2 + 5^2 + 1) = 16,824$ plots, before we even consider conditioning!
-Well thought out interaction can make it easy to navigate this space.
-We have a prototype interaction model developed as part of the cranvas suite of interactive graphics in R, {\tt http://github.com/ggobi/cranvas}, but much work and user testing remains to be done.
+The product plots framework defines a large space of potential plots.
 
-We also are hopeful that developing the connection between product plots and log-linear (aka Poisson) models will also help better understand this space of plots.
-Some special cases have already been worked out.
+Well thought out interaction will make it easier to navigate this space, and we have begun to develop a prototype model in R, as part of the cranvas suite of interactive graphics, {\tt http://github.com/ggobi/cranvas}.
+However, much work and user testing remains to be done before we can be confident that we have developed a useful navigation model.
+
+We are also exploring the connection between product plots and log-linear (aka Poisson) models, the statistical models most commonly used for count data.
+Some special cases of the general connection have already been worked out.
 For example, \citep{hofmann:2003} shows how looking for straight lines in a mosaic plot corresponds to a formal test of independence between two variables.
-Generally, what visual comparisons are equivalent to what formal statistical tests?
-And conversely, how can inspecting the significant coefficients in a model help us choose a set of useful plots for a given dataset?
+Generally, we are interested in knowing what visual comparisons are equivalent to what formal statistical tests, and conversely, how significant coefficients in a model can help us choose a set of useful plots.
 Generally, given a question, can we suggest appropriate plots?
@@
 
-Product plots need to be extended with systematic ways of displaying uncertainty, to help users identify whether differences in the plot represent real differences in the underlying population or are just the result of random variation.
-This is particularly important because area plots typically have fixed area, meaning that the total number of observations in a plot is not directly displayed.
-This is crucial for determining whether a difference is truly significant or not.
+Product plots also need to be extended with systematic ways of displaying uncertainty, to help users identify whether differences in the plot represent real differences in the underlying population or are just the result of random variation.
+This is particularly important because area plots typically have fixed area, meaning that the total number of observations in a plot is not directly displayed, even though this is crucial for determining whether a difference is statistically significant or not.
 

### 2011-06-22 prodplots-infovis.tex
extended discussion on continuous variables, started related work section, fixed some typos

--- 2011-03-30 prodplots-infovis.tex
@@
 \authorfooter{
-  \item Hadley Wickham is an Assistant Professor of Statistics at Rice University, Email: hadley@rice.edu.
-  \item Heike Hofmann is an Associate Professor of Statistics at Iowa State University.
+  \item Hadley Wickham is Assistant Professor of Statistics at Rice University, Email: hadley@rice.edu.
+  \item Heike Hofmann is Associate Professor of Statistics at Iowa State University.
 }
@@
 
+A more theoretical approach in dealing with continuous variables  lies in increasing the number of bins infinitesimally -- this leads from a probability mass function to a density function, making the 1d primitives of hbar and vbar to density plots, and changing the spine-plot representation to a conditional density plot (citation needed - probably Martin's and Heike's paper).
+While the inclusion of one continuous variable is completely feasible that way, two variables need the use of an additional aesthetic, such a s colour,  to visualize a 2d joint density.
+Three or more continuous variables break the scope of the product plots framework.
+
 \begin{figure}[htbp]
@@
 \noindent There are 24 possible ways of ordering the variables in the {\sc pmf}, 5 ways of displaying a 1d {\sc pmf}, and 1 way of displaying a 2d {\sc pmf}, leading to a possible $24 * (5^4 + 5^2 + 5^2 + 5^2 + 1) = 16,824$ plots, before we even consider conditioning!
-Well thought out interaction can make it easy to navigate this space.
+Well thought out interaction can make it easier to navigate this space.
 We have a prototype interaction model developed as part of the cranvas suite of interactive graphics in R, {\tt http://github.com/ggobi/cranvas}, but much work and user testing remains to be done.
@@
 And conversely, how can inspecting the significant coefficients in a model help us choose a set of useful plots for a given dataset?
-Generally, given a question, can we suggest appropriate plots?
+Can we, given a question,  suggest appropriate plots?
 Or given a plot, can we suggest questions that it might answer?

### 2011-06-23 prodplots-infovis.tex
Weaken claims that rectangles are best

--- 2011-06-21 prodplots-infovis.tex
@@
 
-All area plots correspond to tilings of the (2d) plane.
+Area plots correspond to tilings of the (2d) plane.
 We could consider partitions of higher-d spaces (e.g.
 3d or 4d), but given that we have to project these down to 2d for viewing on paper or screen, there is little disadvantage to working directly in 2d.
+There are many possible ways to tile the plane, so to weed these down to a manageable number, we identify  constraints, in the style of \citep{keim:2002}, that are important for visualising counts.
 
-There are many possible ways to tile the plane. To weed these down to a manageable number, we identify three constraints \citep{keim:2002} important for visualising counts:
-
-\begin{itemize}
-
-  \item Area must be {\bf proportional} to count.
+Firstly, area must be {\bf proportional} to count.
 This is the key constraint underlying all area plots.
@@
 
-  \item Partitions must be {\bf rectangular}, because the areas of rectangles are easier to compare than areas of other shapes \citep{cleveland:1984,cleveland:1986,heer:2010}, and are recursive, in the sense that we can always tile a rectangle with smaller rectangles.
-This is not true for most shapes.
-
-  \item Partitions must be {\bf disjoint}.
+Secondly, partitions must be {\bf disjoint}.
 To be able to see the complete area, each rectangle must be non-overlapping.
@@
 
-\end{itemize}
-
-\noindent Each of the three constraints can be profitably relaxed, as detailed in Section~\ref{sec:variations}.
+Finally, we require partitions to be {\bf rectangular}.
+If partitions are rectangular, many interesting perceptual tasks only require comparing lengths, or positions along a common scale, tasks which are generally easier than comparing areas \citep{cleveland:1984,cleveland:1986}.
+There is little evidence to suggest that rectangles are ``best'' shape for comparison \citep{hollands:1998,simkin:1987}, but they are computationally simple, recursive, in the sense that we can always tile a rectangle with smaller rectangles, and form the basis for many existing graphics.
 
@@
 Section~\ref{sub:part-1d} describes the three partitions of 1d data, bars, spines and tiles, and Section~\ref{sub:part-2d} describes the one partition of 2d data, the fluct.
+Each of the three constraints can also be relaxed, yielding the additional types of partitions described in Section~\ref{sec:variations}.
 
@@
 
-A number of non-rectangular treemaps have been proposed, such as circular treemaps \citep{wetzel:2008}, space-filling curves \citep{wattenberg:2005} and voronoi treemaps \citep{balzer:2005}.
-Reading these visualisations requires the comparison of the area of arbitrary polygons, which is much more difficult than comparing the area of rectangles \citep{cleveland:1986}.
-This makes these approaches attractive rather than useful.
+Other non-rectangular area graphics include non-rectangular treemaps, such as circular treemaps \citep{wetzel:2008}, space-filling curves \citep{wattenberg:2005} and voronoi treemaps \citep{balzer:2005}.
 

### 2011-06-23 prodplots-infovis.tex
Fix references

@@
 Wilkinson restricts (???
-not quite sure - need to lookit up) the use to standard mosaicplots as defined by \citet{hartigan:1981}, which could easily be extended, however the product plots framework additionally introduces the concept of conditioning on variables.
+not quite sure - need to lookit up) the use to standard mosaicplots as defined by Hartigan \citep{hartigan:1981}, which could easily be extended, however the product plots framework additionally introduces the concept of conditioning on variables.
 
@@
 
-The framework of  Polaris  \cite{stolte:2002} relies heavily on a tabular or cubic format in rendering hierarchical graphics, which makes these charts essentially a three dimensional extension of the trellis framework \citep{cleveland:1994, cleveland:1996}:  in the product plots framework, this approach is captured by conditioning on the variables used in rows and columns, and displaying the other variables unconditioned in each cell of the resulting tabular layout.
+The framework of  Polaris  \cite{stolte:2002} relies heavily on a tabular or cubic format in rendering hierarchical graphics, which makes these charts essentially a three dimensional extension of the trellis framework \citep{cleveland:1994,becker:1996}:  in the product plots framework, this approach is captured by conditioning on the variables used in rows and columns, and displaying the other variables unconditioned in each cell of the resulting tabular layout.
 
- \cite{mackinlay:1986, mackinlay2007}: VisQL, a visual query language is designed to capture the flow of a graphical data exploration.
+ \cite{mackinlay:1986,mackinlay:2007}: VisQL, a visual query language is designed to capture the flow of a graphical data exploration.
 

### 2011-06-23 prodplots-infovis.tex
More work on related work

--- 2011-06-23 prodplots-infovis.tex
@@
 \end{table*}
+
+\section{Related Work}
+\label{sec:related}
+
+The product plots framework is heavily influenced by the work of Wilkinson \citep{wilkinson:1999, wilkinson:2005}, who suggested that both mosaic plots and treemaps can be described as plots with coordinate systems based on recursive partitioning.
+Our work builds on this by adding conditioning, and defining a ``grammar of area plots'', a domain specific language for describing, in detail, this small family of plots.
+Other similar efforts to build domain specific languages for visualisation are APT \citep{mackinlay:1986} and VisQL \citep{mackinlay:2007}: our framework is driven by similar forces, but carves out a much smaller niche.
+
+The HiVE system \citep{slingsby:2009} introduces a notation for describing the states of hierarchical layouts and operators for reconfiguration.
+Our work extends HiVE by adding conditioning, and including non-space filling visualisation.
+HiVE supports both rectangular layouts and polygon layouts.
+Of the rectangular layouts, all but the spatially ordered algorithm are included within the product plots framework.
+
+Polaris \cite{stolte:2002} relies heavily on a tabular or cubic format in rendering hierarchical graphics, which makes these charts essentially a three dimensional extension of the trellis framework \citep{cleveland:1994,becker:1996}.
+In the product plots framework, this approach is captured by conditioning on the variables used in rows and columns, and displaying the other variables unconditioned in each cell of the resulting tabular layout.
 
@@
 
-\section{Related Work}
-\label{sec:related}
-
-The product plots framework is heavily influenced by the work of WIlkinson \citep{wilkinson:1999, wilkinson:2005}, who introduced the set product operator to map hierarchical data to a two-dimensional framework.
-Wilkinson restricts (???
-not quite sure - need to lookit up) the use to standard mosaicplots as defined by \citet{hartigan:1981}, which could easily be extended, however the product plots framework additionally introduces the concept of conditioning on variables.
-
- \citep{slingsby:2009}: the HiVE system provides a notation allowing to describe states of hierarchical layouts and operators for reconfiguration.
-HiVE allows both rectangular layouts and polygon layouts -- out of the rectangular layouts, all but the spatially ordered algorithm are covered by the  framework of product plots.
-
-The framework of  Polaris  \cite{stolte:2002} relies heavily on a tabular or cubic format in rendering hierarchical graphics, which makes these charts essentially a three dimensional extension of the trellis framework \citep{cleveland:1994, cleveland:1996}:  in the product plots framework, this approach is captured by conditioning on the variables used in rows and columns, and displaying the other variables unconditioned in each cell of the resulting tabular layout.
-
- \cite{mackinlay:1986, mackinlay2007}: VisQL, a visual query language is designed to capture the flow of a graphical data exploration.
-
 \section{R package}

### 2011-06-23 reviews/response.txt
More on R package.  Start on response

@@
-# Summary review
+We are grateful for the helpful feedback provided by the reviewers.
+We have added a related work section, including many of the missing references we missed.
+We have also weakened our claims about the superiority of rectangles, instead providing a more pragmatic justification
 
-The reviewers judge Product Plots to be a useful and parsimonious model
-for constructing a large variety of plots and potentially also an example
-that might spur further research in formal models of visualization
-construction. All reviewers agree that this work makes a substantive
-contribution to InfoVis research and argue for acceptance.
+Other changes:
 
-That said, reviewers also raise a number of issues that the authors
-should address prior to publication. Most notable among these is the
-treatment of related work. R1 and R4 in particular list elements of prior
-work that should be considered and discussed. I recommend that the
-authors introduce a dedicated related work section in their next
-revision.
+* Rather than eliminating the section describing the R implementation altogether, we have expanded it slightly to give a taste of how it might be used.
 
-In addition, the reviewers make a number of helpful suggestions that
-might inform future drafts. I strongly encourage the authors to revise
-the paper accordingly.
-
-# Major changes
-
-* Related work section & missing references
-* Improve justification of rectangles using references provided.
-
-# Review 2
-
-* First, related work is distributed throughout the paper. A consolidated
-related work section addressing similar work and clearly delineating the
-contributions of the present work is needed. Other visualization models
-of interest include Slingsby et al's HiVE, Wilkinson's Grammar of
-Graphics, Mackinlay's APT, and Stolte et al's Polaris. A detailed
-comparison with such prior work is a pre-requisite to publication. How do
-such models compare in terms of expressiveness and notational efficiency?
-
-* I'm curious how much effort is necessary to use product plots in
-practice. Some examples of statements in R and the corresponding plots
-should suffice (and potentially examples of an interactive UI for product
-plot construction).
-
-* Is anyone using the system to perform analyses? While not strictly
-necessary, example analysis case studies supported by product plots might
-help convey the value of the model.
-
-* "the areas of rectangles are easier to compare than areas of other shapes
-[5-7]" -> This is somewhat glibly stated. A quick review of the cited
-papers does not support the claim: [5] does not assess area empirically;
-[6] assesses circular and arbitrary polygonal areas, but not rectangles;
-[7] compares circles and rectangles, but the difference does not appear
-to be statistically significant. Later in the paper, there is a similar
-critique of non-rectangular polygons using the results of [6]. I don't
-think this is appropriate -- [6] uses irregular polygons, whereas
-techniques like voronoi treemaps generate convex polygons with average
-bounding box aspect ratios near unity. Thus the argument here is also
-unsubstantiated. It may well be that rectangles are better, but this has
-not been convincingly demonstrated. I recommend the authors consider
-dropping these perceptual arguments.
-
-# Review 4
-
-* A surprising omission is the work of Wilkinson (1999, 2005) on The
- Grammar of Graphics. Wilkinson introduced the idea of combining graphical
- elements through products of sets (called facets) and embedding these in
- different coordinate systems. In fact, Figure 12 echoes several of the
- figures in the Coordinates chapter of that book. Wilkinson was also the
- first to draw a link between the recursive partitioning in treemaps and
- mosaics and a set product operator. As he explains in Chapter 16 of the
- second edition, the algebra for specifying visualizations of product
- models maps one-to-one to the most common factorial experimental design
- algebra for log-linear models or ANOVA. In any case, the framework in
- this paper appears closer to that of Wilkinson's than to any of the other
- papers cited.
-
- Bottom of page:
- Using the word "all" is a bit strong here. Better to delete the word and
- confine the discussion to "area" plots. There are exceptions, as the
- authors point out later, so avoid the inclusive term.
-
- The claim for rectangles at the bottom of page 1 is not supported by
- research. The authors, like many statisticians and visualization
- researchers, tend to cite statisticians and computer scientists for
- supportive psychological principles; however, the perceptual and
- cognitive research by non-psychologists tends to have serious
- methodological flaws. I've provided some relevant citations below. There
- are significant biases in area judgments that are exacerbated by extreme
- aspect ratios (something even Kong, Heer & Agrawala noticed in Treemaps).
- The perceptual combining rule for rectangles appears to be more Attneave
- (absolute metric) than product (multiplicative), although see Algom et
- al. for support for the multiplicative rule. The authors need to rephrase
- this statement, eliminating the word "must."
-
- Also, Simken and Hastie, in one of the few perception articles published
- by psychologists in a statistical journal, demonstrated conclusively that
- pie charts outperform rectangular area charts for proportion of whole
- data. There are some exceptions to their findings, but statisticians need
- to understand that they are largely wrong in their scorn for pie charts.
-
- We can tile triangles with triangles. And these are "simpler" than
- rectangles. One can also tile the sphere with triangles, which cannot be
- done with rectangles. You need to rethink this assertion.
-
- References:
- Algom, D., Wolf, Y., & Bergman, B. (1985). Integration of stimulus
- dimensions in perception and memory: Composition rules and psychophysical
- relations. Journal of Experimental Psychology: General, 114, 451-471.
-
- Anderson, N.H. and Weiss, D.J. (1971). Test of a multiplying model for
- estimated are of rectangles. American Journal of Psychology, 84, 4,
- 543-548.
-
- Borg, I. and Leutner, D. (1983). Dimensional models for the perception of
- rectangles. Perception & Psychophysics, 34, 257-267.
-
- Kidman, G. (1997). Area integration rules. In F. Biddulph & K. Carr
- (Eds.), People in mathematics (Proceedings of the 20th annual conference
- of the Mathematics Education Research Group of Australasia (Vol. 1, pp.
- 271 - 277).
-
- Mates, J., Di Maio, V., and Lansky, P. (1992). A model of the perception
- of area. Spatial Vision, 6, 101-116.
-
- Morgan M.J. (2005). The visual computation of 2-D area by human
- observers. Vision Research, 45(19), 2564-70.
+------
 

### 2011-06-23 prodplots-infovis.tex
More on R package.  Start on response

@@
 Other similar efforts to build domain specific languages for visualisation are APT \citep{mackinlay:1986} and VisQL \citep{mackinlay:2007}: our framework is driven by similar forces, but carves out a much smaller niche.
+This makes products plots less expressive, but more efficient, because the number of primitives required to describe the smaller domain is much smaller.
 
@@
 The code is well tested, and ensures that the constraints are always satisfied.
-The {\tt productplots} code to used to create the images in this paper is available from the authors.
+
+For example, the following code conveys the essence of the plots in Figure~\ref{fig:part-cond}.
+The function {\tt prodplot} creates plots of the {\tt happy} data, defined using the standard R formula notation (\verb|~|), with the convention that {\tt |} denotes conditioning.
+The final argument lists the graphical primitives to use for display.
+This can also be a template function like {\tt mosaic()} which produces a standard, named, graphic.
+
+\begin{verbatim}
+prodplot(happy, ~ happy + finrela + health, 
+  c("vspine", "fluct"))
+prodplot(happy, ~ happy | finrela + health, 
+  c("vspine", "fluct"))
+prodplot(happy, ~ happy | finrela + health, 
+  c("hspine", "fluct"))
+\end{verbatim}
+
+As for all presentation graphics, the actual code is somewhat more complicated, as we make a number of tweaks for optimal display.
+The complete {\tt productplots} code to used to create the images in this paper is available from the authors upon request.
 

### 2011-06-23 prodplots-infovis.tex
Label graphics with f syntax

@@
 
-  \caption{Plots of the distribution of happiness and sex (\key{male} male, \key{female} female).}
+  \caption{Plots of the distribution of happiness and sex (\key{male} male, \key{female} female) (Left) $f(happy)$, (Middle) $f(sex | happy)$, (Right) $f(happy, sex)$.}
   \label{fig:fact-simple}
@@
     \includegraphics[width=0.5\linewidth]{part-marital-2}
-  \caption{(Left) The joint distribution of marital status and age (in decades) partitioned by a vspine and hspine.
-(Right) The joint distribution of age and marital status partitioned by a vspine and hbar.}
+  \caption{(Left) The joint distribution of marital status and age, $f(marital, age)$, (in decades) partitioned by a vspine and hspine.
+(Right) The joint distribution of age and marital status, $f(age, marital)$, partitioned by a vspine and hbar.}
   \label{fig:marital}
@@
     \includegraphics[width=0.5\linewidth]{wt-wtssall}%
-  \caption{Joint distribution of age and sex (\key{male} male, \key{female} female). (Left) counts and (right) probability weights. In this case, there is very little difference between the plots.}
+  \caption{Joint distribution of age and sex (\key{male} male, \key{female} female), $f(age, sex)$.
+(Left) counts and (right) probability weights.
+In this case, there is very little difference between the plots.}
   \label{fig:weighting}
@@
 
-  \caption{The distribution of happiness with age. (Left) hbar + vspine: an extension of the histogram. (Right) hspine + vspine: an extension of the spinogram.}
+  \caption{The distribution of happiness with age, $f(age, happy)$. (Left) hbar + vspine: an extension of the histogram. (Right) hspine + vspine: an extension of the spinogram.}
   \label{fig:cont-examples}
@@
     \includegraphics[width=0.5\linewidth]{cascade}
-  \caption{A mosaic plot of happiness by marital status and sex.
+  \caption{A mosaic plot of happiness by marital status and sex, $f(happy, marital, sex)$.
 (Left) Coloured by happiness (\key{not-too-happy} Not too happy, \key{pretty-happy} pretty happy, \key{very-happy} very happy).

### 2011-06-23 reviews/response.txt
More minor fixes

@@
 * Rather than eliminating the section describing the R implementation altogether, we have expanded it slightly to give a taste of how it might be used.
+
+* Where appropriate, figures labelled with distributional decomposition.
 
@@
 
- -label all appropriate plots in the paper with the corresponding
- distribution decomposition (directly label, not just mention in the text
- or in the caption)
 -flesh out section 3.1. Assuming that infovis readers will not know what
@@
 * Figure 4 overruns its column.
-
-* Page 2. "Uniformly distributed" suggest to me a uniform random distribution. Is there another way to say this?
-
-* Figure 3. Left and Right are Top and Bottom. LaTeX refused to obey your
-orders, as usual ;o)
-
-* Please italicize n in the fourth line of Section 3. Check that other
-parameters are properly italicized in the text.
-
-* Page 1. "This framework is based on the fundamental statistical idea that
- high any high dimensional" Delete the first "high."
-
- "51 020 and 5 000" use comma delimiters.
-
-*  Figure 12. The bottom row is not produced by transforming after rotation.
- It is produced by switching the assignment of x and y to r and theta.
- This is a pivoting operation.

### 2011-06-23 prodplots-infovis.tex
More minor fixes

@@
 
-\abstract{
-
-We propose a new framework for visualising tables of counts, proportions and probabilities.
+\abstract{We propose a new framework for visualising tables of counts, proportions and probabilities.
 We call our framework product plots, alluding to the computation of area as a product of height and width, and the statistical concept of generating a joint distribution from the product of conditional and marginal distributions.
-The framework, with extensions, is sufficient to encompass over 20 visualisations previously described in fields of statistical graphics and infovis, including bar charts, mosaic plots, treemaps, equal area plots and fluctuation diagrams.
-
-}
+The framework, with extensions, is sufficient to encompass over 20 visualisations previously described in fields of statistical graphics and infovis, including bar charts, mosaic plots, treemaps, equal area plots and fluctuation diagrams.}
 
@@
 The name spine is evocative of books sitting on a library shelf \citep{hummel:1996}.
-Bars and spines are indistinguishable when the underlying data is uniformly distributed.
+Bars and spines are indistinguishable when the underlying data is evenly distributed across the categories.
 
@@
     \includegraphics[width=0.9 \linewidth]{part-fluct-cond}
-  \caption{(Left) A fluctuation diagram showing distribution of age (in decades) and marital status.
-(Right) Equal bin-size plot showing health status and survey year.
+  \caption{(Top) A fluctuation diagram showing distribution of age (in decades) and marital status.
+(Bottom) Equal bin-size plot showing health status and survey year.
 Thee three empty columns show that health status was not recorded for three years.}
@@
 Some statistical vocabulary is useful: one way of describing the input data is as a {\bf probability mass function}, or {\sc pmf}.
-A {\sc pmf} is a function with n inputs, each indexing one dimension with an integer, and gives the probability of each combination of inputs.
+A {\sc pmf} is a function with $n$ inputs, each indexing one dimension with an integer, and gives the probability of each combination of inputs.
 A {\sc pmf} has two restrictions: every value must be greater than or equal to zero, and all values must sum to one.
@@
 
-  \item {\bf Infoslices} \citep{andrews:1998} (nd): n vbars. But they only use half of the polar plane, and are specialised for highly nested data.
+  \item {\bf Infoslices} \citep{andrews:1998} (nd): $n$ vbars. But they only use half of the polar plane, and are specialised for highly nested data.
 
@@
 (Mid) Area graphics in polar coordinates.
+(Bottom) Product graphics with alternate mapping of $x$ and $y$ to $r$ and $\theta$.
 From left to right: hbar + hbar, vspline + hbar, hspine + hspine, hspine + vspine.
-(Bottom) Product graphics rotated 90 degrees before being converted to polar coordinates: vbar + vbar, hspline + vbar, vspine + vspine, vspine + hspine.
 (\key{male} male, \key{female} female)}

### 2011-06-23 reviews/response.txt
More minor changes

@@
 * Where appropriate, figures labelled with distributional decomposition.
+
+* We wish we could make the figures larger and the labels easier to read, but the page limits are extremely tight this year.
+We have done our best to ensure that all information is convey redundantly through colour or description in the captions.
 
@@
 
-Label Figure 5 with the distribution forms, f(happy), f(happy|sex),
-f(happy, sex), rather than just mention them in the text.
-
 Figure 7 shows, f(marital | age) f(age) and f(age | marital) f(marital),
@@
 
-In terms of text, I noticed some typos (first page, "This framework is
-based on the fundamental statistical idea that high any high dimensional
-distribution" has too many "high"s), one slight miss in phrasing ('weed
-down' sounds like a mixed metaphor between 'weed out' and 'cut down') and
-perhaps the last column of the conclusion could benefit from editing
-(repetition of 'generally' stands out).
-
 # Minor changes
 
-* There is a line break after "Abstract-". The abstract text should begin
-on the same line.
-
 * Figure 4 overruns its column.

### 2011-06-23 prodplots-infovis.tex
More minor changes

@@
 3d or 4d), but given that we have to project these down to 2d for viewing on paper or screen, there is little disadvantage to working directly in 2d.
-There are many possible ways to tile the plane, so to weed these down to a manageable number, we identify  constraints, in the style of \citep{keim:2002}, that are important for visualising counts.
+There are many possible ways to tile the plane, so to cut these down to a manageable number, we identify  constraints, in the style of \citep{keim:2002}, that are important for visualising counts.
 
@@
 Generally, we are interested in knowing what visual comparisons are equivalent to what formal statistical tests, and conversely, how significant coefficients in a model can help us choose a set of useful plots.
-Generally, given a question, can we suggest appropriate plots?
-Or given a plot, can we suggest questions that it might answer?
+Given a question, can we suggest appropriate plots?
+Given a plot, can we suggest questions that it might answer?
 

### 2011-06-23 prodplots-infovis.tex
added model names to figure 6
overlooked a centering
width of figure 4
additional labels on the plots
additional legends comments

@@
   \centering
+  $f(age, marital)$
     \includegraphics[width=0.9 \linewidth]{part-fluct}
+  $f(1 | age, marital)$
     \includegraphics[width=0.9 \linewidth]{part-fluct-cond}
@@
 \begin{figure}[ht]
-  \begin{tabular}{rrr}
+  \begin{tabular*}{0.5\linewidth}{rrr}
+\multicolumn{3}{l}{$f(happiness, sex)$}\\
   \toprule
@@
   \bottomrule
-  \end{tabular}
+  \end{tabular*}
   \\[1em]
 
-  \begin{tabular}{rrr}
+  \begin{tabular*}{0.48\linewidth}{rrr}
+\multicolumn{3}{l}{$f(happiness| sex)$}\\
   \toprule
@@
   \bottomrule
-  \end{tabular}\hspace{1em}%
-  \begin{tabular}{rrr}
+  \end{tabular*}
+%
+\hspace{1em}
+  \begin{tabular*}{0.48\linewidth}{rrr}
+\multicolumn{3}{l}{ $f(sex|happiness)$}\\
   \toprule
@@
   \bottomrule
-  \end{tabular}
+  \end{tabular*}
   \\[1em]
@@
   \begin{tabular}{rrrr}
+\multicolumn{3}{l}{$f(happiness), f(sex)$}\\
   \toprule
@@
 \begin{figure}[htbp]
-  \centering
-    \includegraphics[width=0.33\linewidth]{part-comb-1}%
-    \includegraphics[width=0.33\linewidth]{part-comb-2}%
-    \includegraphics[width=0.33\linewidth]{part-comb-3}
+\begin{minipage}[b]{0.3\linewidth}
+  	$f(marital)$
+      \includegraphics[width=\linewidth]{part-comb-1}%
+\end{minipage} 
+\begin{minipage}[b]{0.3\linewidth}
+$f(marital, sex)$ 
+       \includegraphics[width=\linewidth]{part-comb-2}
+\end{minipage} 
+\begin{minipage}[b]{0.3\linewidth}
+$f(happy, sex, marital) =$
+$f(happy|marital, sex) \cdot $
+$f(marital, sex)$
+      \includegraphics[width=\linewidth]{part-comb-3}
+\end{minipage}
   \caption{Conditional on marital status, are men or women happier?
@@
   \centering
-    \includegraphics[width=0.5\linewidth]{part-marital-1}%
-    \includegraphics[width=0.5\linewidth]{part-marital-2}
-  \caption{(Left) The joint distribution of marital status and age, $f(marital, age)$, (in decades) partitioned by a vspine and hspine.
+$f(marital, age) =$\\
+\begin{minipage}[t]{0.49\linewidth}
+\centering 
+$f(marital|age)f(age)$ 
+    \includegraphics[width=\linewidth]{part-marital-1}%
+ \end{minipage}
+\begin{minipage}[t]{0.49\linewidth}
+\centering
+ $f(age|marital)f(marital)$
+    \includegraphics[width=\linewidth]{part-marital-2}
+ \end{minipage}
+  \caption{(Left) The joint distribution of marital status and age, $f(marital, age)$ as $f(marital|age) \cdot f(age)$, (in decades) partitioned by a vspine and hspine.
 (Right) The joint distribution of age and marital status, $f(age, marital)$, partitioned by a vspine and hbar.}
@@
   \centering
-    \includegraphics[width=0.33\linewidth]{part-fluctuation}%
-    \includegraphics[width=0.33\linewidth]{part-equal-area}%
-    \includegraphics[width=0.33\linewidth]{part-equal-area-2}
+\begin{minipage}[b]{0.32\linewidth}
+\centering
+$f(happy, health, finrela) =$
+$f(happy | health, finrela) \times f(health, finrela)  $
+     \includegraphics[width=\linewidth]{part-fluctuation}
+ \end{minipage}
+%
+\begin{minipage}[b]{0.32\linewidth}
+\centering
+$f(happy | health, finrela)$
+    \includegraphics[width=\linewidth]{part-equal-area}
+ \end{minipage}
+%
+\begin{minipage}[b]{0.32\linewidth}
+\centering
+$f(happy | finrela, health)$
+    \includegraphics[width=\linewidth]{part-equal-area-2}
+ \end{minipage}
+%
+    %
+
   \caption{(Left) $f(happy, health, finrela) = f(happy | health, finrela) \times f(health, finrela)  $, partitioned with a vspine and fluct.
@@
 Healthier and richer people are happier: maybe money does buy happiness?
-(Right) $f(happy | health, finrela)$ partitioned with a fluct and {\bf h}spine.
+(Right) $f(happy | health, finrela)$ partitioned with a fluct and {\bf h}spine, emphasizing the relationship of happiness with finances, whereas the middle plot emphasizes the relationship with health.
 (\key{not-too-happy} Not too happy, \key{pretty-happy} pretty happy, \key{very-happy} very happy)}

### 2011-06-27 prodplots-infovis.tex
conditional and marginal density from joint

@@
   \begin{tabular*}{0.5\linewidth}{rrr}
-\multicolumn{3}{l}{$f(happiness, sex)$}\\
+\multicolumn{3}{l}{$f(happy, sex)$, see fig. \ref{fig:fact-simple} (right) }\\
   \toprule
@@
   \begin{tabular*}{0.48\linewidth}{rrr}
-\multicolumn{3}{l}{$f(happiness| sex)$}\\
+\multicolumn{3}{l}{$f(happy| sex)$}\\
   \toprule
@@
   \begin{tabular*}{0.48\linewidth}{rrr}
-\multicolumn{3}{l}{ $f(sex|happiness)$}\\
+\multicolumn{3}{l}{ $f(sex|happy)$, see fig. \ref{fig:fact-simple} (middle) }\\
   \toprule
@@
   \begin{tabular}{rrrr}
-\multicolumn{3}{l}{$f(happiness), f(sex)$}\\
+\multicolumn{3}{l}{$f(happy)$, see fig. \ref{fig:fact-simple} (left) and $f(sex)$}\\
   \toprule
@@
 Formally, a conditional distribution function is written $f(x | y)$ and is equal to $f(x, y) / f(y)$.
-This definition illustrates that an important statistical fact: given a conditional distribution and a marginal distribution, we can always find the joint distribution: $f(x, y) = f(x | y)\,f(y)$ \footnotemark.
-This definition extends in a straightforward way to higher dimensions.
+This definition illustrates that an important statistical fact: given a conditional distribution and a marginal distribution, we can always find the joint distribution: $f(x, y) = f(x | y)\,f(y)$\footnotemark.
+Conversely, we also see that we need to know both the marginal and conditional distributions in order to re-construct the joint distribution, i.e.
+only the joint distribution contains the full information about the relationship between all variables involved.
+Given the joint distribution $f(x,y)$, we can get either marginal distribution  by integrating (or summing, in the case of a categorical variable)  the other variable out: $f(x) = \int_y f(x,y) dy$.
+The marginal distribution together with the joint then lead to the conditional distribution.
+These definitions extend in a straightforward way to higher dimensions.
 For example, a 3d joint distribution can be written as the product of 2d and 1d conditional and marginal distributions in the following three ways:
@@
 
-This means that we can turn any high-dimensional {\sc pmf} into a product of low-dimensional conditional and marginal {\sc pmf}s, a computationally trivial operation.
-It remains to find a graphical analog of this multiplication.
+This means that we can build any high-dimensional {\sc pmf} as a product of low-dimensional conditional and marginal {\sc pmf}s, a computationally trivial operation.
+In the following, we are using the previously introduced low dimensional primitives  for a graphical analog of this multiplication.
 

### 2011-06-27 prodplots-infovis.tex
'hand-holding' examples

@@
 Figure~\ref{fig:2d-table} shows three ways to represent the the 2d table of proportions of sex and happiness.
-The top table displays the {\bf joint distribution}, and allows us to answer questions of the form ``what proportion of all people are male and very happy?'' The middle two tables displays two {\bf conditional distributions}: the distribution of sex given happiness, and happiness given sex.
-These correspond to restricting the row or column sums to one, and support questions such as ``what proportion of pretty happy people are female?'' or ``what proportion of males are not too happy?''.
+The top table displays the {\bf joint distribution}, and allows us to answer questions of the form ``what proportion of all people are male and very happy?'' -- the bottom left number tells us that these are 0.14.
+The middle two tables displays two {\bf conditional distributions}: the distribution of sex given happiness, and happiness given sex.
+These correspond to restricting the row or column sums to one, and support questions such as ``what proportion of pretty happy people are female?'' --  the table tells us that it's 0.55, or ``what proportion of males are not too happy?'' -- the top left number in the table tells us that that's 0.12.
 Finally, the last table displays the two {\bf marginal distributions} of sex and happiness.
-These allow us to answer questions like ``what proportion of respondents were male'' or ``what proportion of respondents were very happy?''.
+These allow us to answer questions like ``what proportion of respondents were male'' (0.44) or ``what proportion of respondents were very happy?'' (0.30).
 
@@
 
-  \begin{tabular*}{0.48\linewidth}{rrr}
-\multicolumn{3}{l}{$f(happy| sex)$}\\
+  \begin{tabular*}{0.49\linewidth}{rrr}
+\multicolumn{3}{l}{$f(sex|happy)$, see fig. \ref{fig:fact-simple} (middle) }\\
   \toprule
@@
 \hspace{1em}
-  \begin{tabular*}{0.48\linewidth}{rrr}
-\multicolumn{3}{l}{ $f(sex|happy)$, see fig. \ref{fig:fact-simple} (middle) }\\
+  \begin{tabular*}{0.47\linewidth}{rrr}
+\multicolumn{3}{l}{$f(happy| sex)$}\\
   \toprule

### 2011-06-27 reviews/response.txt
Final tweaks

@@
 We are grateful for the helpful feedback provided by the reviewers.
-We have added a related work section, including many of the missing references we missed.
+We have added a related work section, including all of the references we missed.
 We have also weakened our claims about the superiority of rectangles, instead providing a more pragmatic justification
@@
 
-* Where appropriate, figures labelled with distributional decomposition.
+* Figures are labelled with distributional decomposition.
+
+* We have expanded the section that introduces joint, marginal and conditional densities to be more helpful for newcomers to the topic.
 
@@
 
-------
-
-# Review 1
-Finally, I wonder if the authors considered any other graphical
-composition method than nesting areas? While nesting areas allowed them
-to encompass a number of existing plot types, other composition methods
-might be more useful for extension to higher dimensions. For example,
-what about simply showing a separate plot for each component?
-
--flesh out section 3.1. Assuming that infovis readers will not know what
-conditional and marginal distributions are, the authors should
-demonstrate how they are computed from the joint distribution. Providing
-the graphical analogs to the tables in Figure 4 would be extremely
-helpful. (I realize that a marginal and a conditional distribution from
-Figure 4 are represented in Figure 5, but the connection is never made
-explicitly in the text. Some hand holding would be very helpful here.)
-Hand hold better the analytic questions that each distribution helps
-answer. E.g. "What proportion of males are very happy? The bottom left
-entries tells us it's 31%."
-
-Minor comments----
-
-Why squarified treemaps? Heer work showing non-square treemaps are easier
-to read.
-
-Section 3.1
-
-Its not clear from the wording in the first paragraph that the marginal
-and conditional distributions contain less information than the joint
-distribution. If you're going to assume that your readers don't know what
-the marginal and conditional distributions are (plausible for the InfoVis
-crowd), then it would make sense to briefly describe how they are
-computed from the joint distribution.
-
-Fix the formatting of the tables in Figure 4.
-
-Clarify in the last paragraph that in addition to the 3 factorings,
-permutations of the variables leads to more variations.
-
-"It remains to find an [sic] graphical analog of this multiplication."
-seems unnecessarily dramatic. We already know you found one.
-
-Section 3.2
-
-Figure 7 shows, f(marital | age) f(age) and f(age | marital) f(marital),
-right? Say this in the text and label figures.
-
-"the relationship between proportion and area is only constant within a
-level" -- is there a better way to say this? I think you mean level as in
-nesting level, not level as is a level of a factor, right? In Figure 7
-(right), the nested bar charts are all sized proportionally with respect
-to each other even though they occur in different levels of the marital
-status factor.
-
-Figure 8 middle would be better with a bar chart, right? Why show both
-the middle and right plots; they're basically the same.
-
-Section 6.2
-
-Why limit yourself to a finite number of bins? Push the number of bins to
-infinity and you can display line graphs (density plots) with hbar,
-stacked line graphs (hbar+vspine), relative area (hspine+vspine),
-overlaid line charts (hbar+hbar), or (hspine+hbar) (don't know the name
-of this chart, basically overlaid line charts, but where the sum of
-values at each vertical slice is 1).
-
-# Review 3
-
-I had some problems, though, in classifying this paper - it falls
-somewhere between theory/model (describes a theoretical framework that
-encompasses a large number of existing visualisations) and technique
-(code is provided to implement the framework). I'm not sure if it'd be
-better to simplify the classification by omitting mention of the code
-except as a note near the end - while it's interesting, it's not core to
-the paper.
-
-Doing this would leave more space to address some other minor concerns
-with the figures: in figures 2, 3, 5, 6 and 7 it's quite clear that the
-labels on the plot have spacing issues that could be resolved with larger
-figures. Additionally, figure 4 spills over the gutter into the text in
-the right column.
-
-# Minor changes
-
-* Figure 4 overruns its column.
+* We have endeavoured to make all of the other minor improvements to spelling and grammar that the reviewers suggested.  Thanks!

### 2011-06-27 prodplots-infovis.tex
Final tweaks

@@
 
-We begin in Section~\ref{sec:primitives} by motivating the three constraints that give rise to the basic graphical component of product plots: the rectangle.
+We begin in Section~\ref{sec:related} with a review of related work.
+Then Section~\ref{sec:primitives} motivates the three specific graphical constraints at the heart of product plots: the rectangle.
 We constrain the rectangle further to produce three 1d atoms (bar, spine, tile) and one 2d atom (the fluct).
@@
 The product plots framework is heavily influenced by the work of Wilkinson \citep{wilkinson:1999, wilkinson:2005}, who suggested that both mosaic plots and treemaps can be described as plots with coordinate systems based on recursive partitioning.
-Our work builds on this by adding conditioning, and defining a ``grammar of area plots'', a domain specific language for describing, in detail, this small family of plots.
+Our work builds on this by adding conditioning and defining a ``grammar of area plots'', a domain specific language that describes this small family of plots in more detail.
 Other similar efforts to build domain specific languages for visualisation are APT \citep{mackinlay:1986} and VisQL \citep{mackinlay:2007}: our framework is driven by similar forces, but carves out a much smaller niche.
@@
 Some statistical vocabulary is useful: one way of describing the input data is as a {\bf probability mass function}, or {\sc pmf}.
-A {\sc pmf} is a function with $n$ inputs, each indexing one dimension with an integer, and gives the probability of each combination of inputs.
+A {\sc pmf} is a function with $n$ inputs, each indexing one dimension with an integer, which outputs the probability of each combination of inputs.
 A {\sc pmf} has two restrictions: every value must be greater than or equal to zero, and all values must sum to one.
@@
 The middle two tables displays two {\bf conditional distributions}: the distribution of sex given happiness, and happiness given sex.
-These correspond to restricting the row or column sums to one, and support questions such as ``what proportion of pretty happy people are female?'' --  the table tells us that it's 0.55, or ``what proportion of males are not too happy?'' -- the top left number in the table tells us that that's 0.12.
+These correspond to restricting the row or column sums to one, and support questions such as ``what proportion of pretty happy people are female?'' (the table tells us 0.55), or ``what proportion of males are not too happy?'' (the top left number in the table tells us that it's 0.12).
 Finally, the last table displays the two {\bf marginal distributions} of sex and happiness.
@@
   \begin{tabular*}{0.5\linewidth}{rrr}
-\multicolumn{3}{l}{$f(happy, sex)$, see fig. \ref{fig:fact-simple} (right) }\\
+\multicolumn{3}{l}{$f(happy, sex)$}\\
   \toprule
@@
 
-  \begin{tabular*}{0.49\linewidth}{rrr}
-\multicolumn{3}{l}{$f(sex|happy)$, see fig. \ref{fig:fact-simple} (middle) }\\
+  \begin{tabular*}{0.47\linewidth}{rp{1.5em}p{1.5em}}
+\multicolumn{3}{l}{$f(sex|happy)$}\\
   \toprule
@@
 %
-\hspace{1em}
-  \begin{tabular*}{0.47\linewidth}{rrr}
+\hspace{0.5em}
+  \begin{tabular*}{0.47\linewidth}{rp{1.5em}p{1.5em}}
 \multicolumn{3}{l}{$f(happy| sex)$}\\
@@
   \begin{tabular}{rrrr}
-\multicolumn{3}{l}{$f(happy)$, see fig. \ref{fig:fact-simple} (left) and $f(sex)$}\\
+\multicolumn{3}{l}{$f(happy)$ and $f(sex)$}\\
   \toprule
@@
   \caption{(Left) The joint distribution of marital status and age, $f(marital, age)$ as $f(marital|age) \cdot f(age)$, (in decades) partitioned by a vspine and hspine.
-(Right) The joint distribution of age and marital status, $f(age, marital)$, partitioned by a vspine and hbar.}
+(Right) The joint distribution of age and marital status, $f(age, marital)$ as $f(age T|marital) \cdot f(marital)$, partitioned by a vspine and hbar.}
   \label{fig:marital}
@@
 
-A more theoretical approach in dealing with continuous variables  lies in increasing the number of bins infinitesimally -- this leads from a probability mass function to a density function, making the 1d primitives of hbar and vbar to density plots, and changing the spine-plot representation to a conditional density plot (citation needed - probably Martin's and Heike's paper).
-While the inclusion of one continuous variable is completely feasible that way, two variables need the use of an additional aesthetic, such a s colour,  to visualize a 2d joint density.
-Three or more continuous variables break the scope of the product plots framework.
+A more theoretical approach in dealing with continuous variables lies in increasing the number of bins infinitesimally, which leads from a probability mass function to a probability density function, turning the 1d primitives bar primitives to density plots, and the 1d spine primitives to a conditional density plot \citep{hofmann:2005}.
+That approach lends itself to the inclusion of one continuous variable, but two variables need an additional aesthetic, such as colour, to visualize a 2d joint density.
+Three or more continuous variables break the current scope of the product plots framework.
 
